### 2016-03-05 Paper_v1.tex
initial commit

\documentclass[10pt,twocolumn,twoside,final]{IEEEtran}

% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}
\usepackage{lastpage,fancyhdr,graphicx}
\usepackage{amssymb,amsmath}

\bibliographystyle{plain}

% Remove brackets from numbering in List of References
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

% *** SUBFIGURE PACKAGES ***
\ifCLASSOPTIONcompsoc
  \usepackage[caption=false,font=footnotesize,labelfont=sf,textfont=sf]{subfig}
\else
  \usepackage[caption=false,font=footnotesize]{subfig}
\fi

\begin{document}

\title{Effective Dynamic Models of Metabolic Networks}

% author names and affiliations
% use a multiple column layout for up to three different
% affiliations
\author{\IEEEauthorblockN{Michael Vilkhovoy\IEEEauthorrefmark{1},
Mason Minot\IEEEauthorrefmark{2},
Jeffrey D. Varner\IEEEauthorrefmark{1}}\\
\IEEEauthorblockA{\IEEEauthorrefmark{1}School of Chemical Engineering, Purdue University, West Lafayette, IN 47907 USA}
\IEEEauthorblockA{\IEEEauthorrefmark{2}School of Chemical and Biomolecular Engineering, Cornell University, Ithaca, NY 14850 USA}
\thanks{Manuscript received December 1, 2012; revised September 17, 2014.
Corresponding author: J. Varner (email:jdvarner@purdue.edu).}}

\maketitle

\begin{abstract}
Mathematical models of biochemical networks are useful tools to understand and ultimately predict how cells utilize nutrients to produce valuable products.
Modified version of the last paragraph of the introduction.

\end{abstract}

\begin{IEEEkeywords}
Dynamic metabolic models, flux balance analysis, cybernetic models
\end{IEEEkeywords}

\section{Introduction}
Three paragraphs, all must fit (along with abstract) on the first page.
First, introduce the need to model and unstructured, structured, cybernetic, and recent palsson dynamic models of biochemical networks.
Second paragraph introduce constraints based modeling, and its modifications along with EMs. Stress the issues with calculating EMs for large networks. Ref the parallel work.
Third paragraph summarize what you have done. This paragraph must start with ``In this study, we developed a ... ''. It describes what you did, what the results were and what big picture conclusions you have drawn from the study. Last like should start with ``Taken together, ... ''. Always past tense. We did this, we saw that. SIMPLE LANGUAGE, SHORT DECLARATIVE SENTENCES.

\begin{figure}[!t]\centering
\includegraphics[width=0.40\textwidth]{./figs/Fig-1-GeneralModel-Results.pdf}
\caption{A: Proof of concept metabolic network with six metabolites and seven reactions.
Intracellular cellmass precursors $A,B$ and $C$ are balanced (no accumulation) while the extracellular metabolites ($A_{e},B_{e}$ and $C_{e}$) are not balanced (can accumulate). The blue-oval denotes the cell boundary, where $q_{j}$ denotes the jth flux across the boundaries ([mmol/gdw-hr]) and $v_{k}$ denotes the kth intracellular flux. B: Simulation of the extracellular metabolites using the FBA hybrid cybernetic approach (solid line) versus elementary modes (points) for the proof of concept metabolic network. FBA modes have comparable model performance compared to elementary modes.
}\label{fig:model-fitting}
\end{figure}

\section{Results}

General case = we showed that EMs and FBA models were equivalent for the small hypothetical model

Small E. coli case = we should that EMs and a fewer number of FBA models gave similar model performance

Large E. coli case = too many EMs, small number of FBA modes allowed us to simulate this network opening up the possibility of genome scale cybernetic models.

Sensitivity results = simple systematic method to eliminate FBA modes to give the smallest number needed, without complex weighting schemes.

In all cases, the first line of the paragraph states the result described in the paragraph. All results are in past tense.

\section{Discussion}
Three paragraphs.
First paragraph, modified version of the last paragraph of the introduction, more details.

Second paragraph, contrast what we have done in the context of the field, in particular compare the hybrid cybernetic models, constraints based models and dynamic models of
Palsson. Why is this study innovative?

Last paragraph, what is wrong with this study and where do we go from here? For example,
could we extend the work done here to animal cells producing biologics, could we extend this to genome scale.

\section{Materials and Methods}

\noindent\subsubsection*{Description of the method}
Describe the method mathematics, use index form if possible, I the vector nomenclature to be hard to follow. Define all terms and symbols.

\subsubsection*{Estimation of model parameters Modify this section}
Model parameters were estimated by minimizing the difference between simulations and experimental thrombin measurements (squared residual):
\begin{equation}\label{eqn:objective-function}
	\min_{\mathbf{k}} \sum_{\tau=1}^{\mathcal{T}}\sum_{j=1}^{\mathcal{S}}\left(\frac{\hat{x}_{j}\left(\tau\right) - x_{j}\left(\tau,\mathbf{k}\right)}{\omega_{j}\left(\tau\right)}\right)^{2}
\end{equation}where $\hat{x}_{j}\left(\tau\right)$ denotes the measured value of species $j$ at time $\tau$, $x_{j}\left(\tau,\mathbf{k}\right)$ denotes the simulated
value for species $j$ at time $\tau$, and $\omega_{j}\left(\tau\right)$ denotes the experimental measurement variance for species $j$ at time $\tau$.
The outer summation is with respect to time, while the inner summation is with respect to state. We minimized the model residual using Particle swarm optimization (PSO) \cite{PSO}. PSO uses a \textit{swarming} metaheuristic to explore parameter spaces.
For each iteration, particles in the swarm compute their local error by evaluating the model equations using their specific parameter vector realization.
From each of these local points, a globally best error is identified. Both the local and global error
are then used to update the parameter estimates of each particle using the rules:
\begin{eqnarray}
	\mathbf{\Delta}_{i} &=&\theta_{1}\mathbf{\Delta}_{i} + \theta_{2}\mathbf{r}_{1}\left(\mathcal{L}_{i} - \mathbf{k}_{i}\right) + \theta_{3}\mathbf{r}_{2}\left(\mathcal{G} - \mathbf{k}_{i}\right) \\
	\mathbf{k}_{i} &=& \mathbf{k}_{i} + \mathbf{\Delta}_{i}
\end{eqnarray}where $\left(\theta_{1},\theta_{2},\theta_{3}\right)$ are adjustable parameters, $\mathcal{L}_{i}$ denotes the local best solution found by particle $i$, and
$\mathcal{G}$ denotes the best solution found over the entire population of particles. The quantities $r_{1}$ and $r_{2}$ denote uniform random vectors with the same dimension as the number of unknown model
parameters ($\mathcal{K}\times{1}$). In thus study, we used $\left(\theta_{1},\theta_{2},\theta_{3}\right) = \left(1.0, 0.05564, 0.02886\right)$. The quality of parameter
estimates was measured using goodness of fit (model residual). The particle swarm optimization routine was implemented in the Python programming language.

% References -
\bibliography{Paper_v1}

\end{document}

### 2016-03-08 Paper_v1.tex
Results General Model Update1

@@
 Mathematical models of biochemical networks are useful tools to understand and ultimately predict how cells utilize nutrients to produce valuable products.
-Modified version of the last paragraph of the introduction.
+Modified version of the last paragraph of the introduction. Mike made this change.
 
@@
 \section{Results}
+In the general case model, we demonstrated that the performance of the hybrid cybernetic model with FBA modes was equivalent to that of the elementary modes.
+In the proof of concept metabolic network consisting of 6 metabolites and 7 reactions (Fig.
+\ref{fig:model-fitting} A), METATOOL generated 6 elementary modes.
+FBA generated 3 possible modes (pathways) for the same metabolic network.
+Temporal profiles of the extracellular metabolites and artificial biomass concentration are presented for both the HCM FBA and HCM EM models (Fig.
+\ref {fig:model-fitting} B).
+HCM FBA has 9 kinetic parameters whereas HCM EM has 15 kinetic parameters.
+The kinetic parameters of HCM FBA were varied to fit the concentration profiles of the HCM EM model.
 

### 2016-03-08 Paper_v1.tex
1st Intro Paragraph Update

@@
 SIMPLE LANGUAGE, SHORT DECLARATIVE SENTENCES.
+
+The advent of new and creative approaches to the mathematical modeling of cell metabolism allows for the accurate description and prediction of cell behavior for a wide variety of organism and culture conditions.
+Monod kinetics have traditionally been used as unstructured models to describe cell growth, substrate consumption, and product formation [REF].
+Structured models of metabolic pathways, such as Flux Balance Analysis (FBA), attempt to describe cellular behavior using the stoichiometry of biochemical networks, a pseudo-steady state assumption, an objective function, and system constraints [REF].
+Cybernetic modeling approaches cell metabolism as a resource optimization problem; control variables are used to direct the allocation of resources through the cell in a manner most favorable to cell survival, growth, and product formation (Ramkrishna \& Song, 2012).
+Dynamic Flux Balance Analysis couples FBA with dynamic or static optimization methods to describe the change in metabolic fluxes over time (Mahadevan et al.
+2002).
 

### 2016-03-09 Paper_v1.tex
Draft_1

--- 2016-03-05 Paper_v1.tex
@@
 \usepackage{amssymb,amsmath}
+\usepackage{mathtools}
+\usepackage{wrapfig}
 
-\bibliographystyle{plain}
+\bibliographystyle{unsrt}
 
@@
 Mathematical models of biochemical networks are useful tools to understand and ultimately predict how cells utilize nutrients to produce valuable products. 
-Modified version of the last paragraph of the introduction.
+Hybrid cybernetic models in combination with elementary modes (HCM EM) provides a route to model cellular metabolism. 
+However, this framework is limited to smaller networks due to the computational cost of calculating elementary modes and a high number of solutions for large networks.
+In this study, we developed a hybrid cybernetic model in combination with flux balance analysis (HCM FBA) modes instead of elementary modes. 
+Flux balance analysis provides a sufficient number of modes (metabolic options) for the network to estimate experimental observations.
+We show HCM FBA has comparable model performance to HCM EM for a hypothetical proof of concept metabolic network and for a reduced anaerobic \textit{E. coli} network.   
+HCM FBA was applied to a larger metabolic network where 29 FBA modes were computed, in contrast to over 66,000 elementary modes.
+HCM FBA had robust model performance to fit experimental observations with only 5 of the 29 FBA modes.
 
@@
 \section{Introduction}
-Three paragraphs, all must fit (along with abstract) on the first page.
-First, introduce the need to model and unstructured, structured, cybernetic, and recent palsson dynamic models of biochemical networks.
-Second paragraph introduce constraints based modeling, and its modifications along with EMs. Stress the issues with calculating EMs for large networks. Ref the parallel work.
-Third paragraph summarize what you have done.
-This paragraph must start with ``In this study, we developed a ...
-''.
-It describes what you did, what the results were and what big picture conclusions you have drawn from the study.
-Last like should start with ``Taken together, ...
-''.
-Always past tense.
-We did this, we saw that.
-SIMPLE LANGUAGE, SHORT DECLARATIVE SENTENCES.
+Metabolic networks for simple organisms such as \textit{E. coli} contain throusands of reactions that are highly interconnected. 
+These organisms have the potential to be engineered to produce valuable products.
+Mathematical models of biochemical networks are useful tools to understand and ultimately predict how cells utilize nutrients to produce valuable products. 
+Traditionally, Monod kinetics have been used as unstructured models to describe cell growth, substrate consumption, and product formation, but fail to account for intracellular metabolism\cite{shuler_book}.
+Structured modeling approaches have made great progress to describe metabolic pathways including elementary mode analysis (EM)\cite{2006_vonKamp_Metatool}, metabolic flux analysis (MFA), flux balance analysis (FBA)\cite{2010_orth_NatBiotech}, dynamic FBA\cite{1994_varma_palsson_ApplEnvMicro,2002_Mahadevan_BiophysJ}, and hybrid cybernetic modeling (HCM)\cite{2008_kim_varner_ramkrishna_BiotechProg}.
+
+Constraint based models of metabolic pathways, such as EM, FBA and MFA, attempt to describe intracellular metabolism using the stoichiometry of biochemical networks and a pseudo-steady state assumption\cite{2010_orth_NatBiotech}.
+EM analysis is a powerful tool to decompse a metabolic network into independent minimal pathways that are stoihiometrically and thermodynamically feasible. 
+However, elementary mode calculations are computational expensive and infeasible for genome scale networks\cite{2004_lee_varner_ko_ieee}.
+FBA is frequently used to study genome-scale networks\cite{2010_orth_NatBiotech} and often results in good approximations for yields, however it fails to estimate intracellular metabolism due to its underdetermined solution space.
+Compared with FBA, MFA has good estimations of intracellular metabolism, but is dependent on experimental data.
+Just like FBA, HCM approaches cell metabolism as a resource allocation problem towards a specified objective.
+But HCM has multiple pathways to select from (elementary modes) and can use a combination of modes to direct its resources through the cell.
+HCM has been shown to have comparable internal flux estimations to MFA and is able to estimate dynamic external fluxes with respect to experimental observations \cite{2008_kim_varner_ramkrishna_BiotechProg}.\
+DFBA is able to account for dynamic external fluxes but fails to incorporate the regulatory processes involved with HCM without the use of boolean rules\cite{2001_covert_schilling_palsson}.
+Despite these advantages of HCM, it is restricted to small networks due to the exponential increase of the number of elementary modes for large networks.   
+Recent work has been done to reduce the number of elementary modes by lumping modes into groups using complex weighting schemes and experimental data\cite{2010_song_ramkrishna}, however, HCM is still limited by EM decomposition.
+
+In this study, we developed a hybrid cybernetic model in combination with flux balance analysis modes instead of elementary modes. 
+We show HCM FBA has comparable model performance to HCM EM for a hypothetical metabolic network and a reduced anaerobic \textit{E. coli} network. 
+HCM was then applied to a large metabolic network where the number of elementary modes made it impractical to model with HCM EM.  
+However, HCM FBA was still feasible for the network and performed well with respect to experimental observations. 
+Taken together, HCM FBA has comparable model performace to HCM EM and may be potentially applied to genome-scale networks which was a shortcoming of HCM EM.
 
@@
 \section{Results}
+In the general case model, we demonstrated that the performance of the HCM FBA was equivalent to that of HCM EM.
+In the proof of concept metabolic network consisting of 6 metabolites and 7 reactions (Fig. \ref{fig:model-fitting}A), we calculated 6 elementary modes and 3 FBA modes.  
+Using the HCM EM modeling framework, concentration profiles were generated for the extracellular metabolites and artifical biomass (Fig. \ref {fig:model-fitting}B). 
+The kinetic parameters of HCM FBA were varied to fit the concentration profiles of the HCM EM model. 
+For this proof of concept model, HCM FBA was able to use fewer modes and kinetic parameters to mimic the simulation profiles of HCM EM.
 
-General case = we showed that EMs and FBA models were equivalent for the small hypothetical model
+In the anaerobic \textit{E. coli} case, we demonstrated that HCM FBA has comparable model performance to a published HCM EM model\cite{2008_kim_varner_ramkrishna_BiotechProg}.
+The metabolic network consisted of 12 reactions, where we calculated 9 elementary modes and 7 FBA modes.
+The simulated HCM EM concentration profiles of biomass, substrate and fermentation products (Fig.
+\ref {fig:ecoli}A (dashed line)) were generated using the kinetic parameters and elementary modes reported\cite{2008_kim_varner_ramkrishna_BiotechProg}.
+The kinetic parameters of HCM FBA were varied to fit the same experimental observations.
+HCM FBA has 17 kinetic parameters while HCM EM has 21 kinetic parameters and both models resulted in similar fits to the experimental data (Fig. \ref {fig:ecoli}A).
+HCM FBA had a slightly better fit on lactate than HCM EM.
 
-Small E. coli case = we should that EMs and a fewer number of FBA models gave similar model performance
+\begin{figure}[!t]\centering
+\includegraphics[width=0.48\textwidth]{./figs/Fig-2-Ecoli-SimulationResults.pdf}
+\caption{A: Batch anaerobic \textit{E.
+coli} fermentation data versus simulation of the hybrid cybernetic model with FBA modes (solid line) and elementary modes (dashed line).
+(Experimental data from Kim et.
+al (2008)\cite{2008_kim_varner_ramkrishna_BiotechProg} with error bars representing 90\% confidence intervals.) B: Batch aerobic \textit{E.
+coli} fermentation data of glucose, acetate and cell density measurements versus simulation of the hybrid cybernetic model with FBA modes (solid line).
+Model performance is shown when minor modes (dashed) and a major mode (dotted) is removed from HCM FBA.
+(Experimental data from Varma \& Palsson (1994)\cite{1994_varma_palsson_ApplEnvMicro} with assumed 10\% error bars).}
+\label{fig:ecoli}
+\end{figure}
 
-Large E. coli case = too many EMs, small number of FBA modes allowed us to simulate this network opening up the possibility of genome scale cybernetic models.
+In the aerobic \textit{E. coli} case, we demonstrated that HCM FBA is applicable for larger metabolic networks where elementary mode calculations increase exponentially.
+A metabolic network\cite{2007_schuetz_etal_MolSysBio} of 98 reactions and 60 metabolites with the biomass reaction from Palsson's \textit{E. coli} core model\cite{2006_Palsson_model} was used.
+We calculated over 66,000 elementary modes for this network and determined it was futile to model. 
+For the same network, FBA computed 29 modes. 
+HCM FBA was simulated to fit cell density, glucose and acetate measurements for an aerobic \textit{E. coli} batch culture\cite{1994_varma_palsson_ApplEnvMicro} (Fig.\ref{fig:ecoli}B (solid line)).
 
-Sensitivity results = simple systematic method to eliminate FBA modes to give the smallest number needed, without complex weighting schemes.
+A variance based sensitivity analysis on the aerobic \textit{E. coli} case identified essential modes that were influential on biomass yield. 
+The analysis calculated the total order sensitivity indicies of all the kinetic parameters and enzyme initial conditions (Fig. \ref{fig:sensitivity}).
+We used the sensitivity results as a simple systematic method to eliminate FBA modes to give the smallest number needed while maintaining good model performance.   
+Five FBA modes were determined to be significant by an arbitrary threshold of 0.00218. 
+The remaining 24 FBA modes were identified to be minor modes and insignificant. 
+The removal of the major mode, representing aerobic growth, (Fig, \ref{fig:ecoli}B (dotted line)) was detrimental to model performance, whereas the removal of the minor modes (dashed line) had little effect.
 
-In all cases, the first line of the paragraph states the result described in the paragraph. All results are in past tense.
+\begin{figure*}[!t]\centering
+\includegraphics[width=0.95\textwidth]{./figs/Fig-3-Sensitivity-Results.pdf}
+\caption{Total order variance based sensitivity analysis on biomass yield from glucose and acetate.
+Sensitivity indicies computed for rate constants, saturation constants and enzyme parameters.
+The most critical parameters are associated with mode 1 representing aerobic growth.
+(Error bars represent 95\% confidence intervals).
+}
+\label{fig:sensitivity}
+\end{figure*}
 
 \section{Discussion}
-Three paragraphs.
-First paragraph, modified version of the last paragraph of the introduction, more details.
+In this study, we developed a hybrid cybernetic model in combination with flux balance analysis modes instead of elementary modes. 
+First, we showed HCM FBA can have comparable model performance to HCM EM for a proof of concept metabolic network. 
+This judgement is based on the quality of fit of the model simulation to the observations. 
+Next, HCM FBA model performance was compared to an established HCM EM model\cite{2008_kim_varner_ramkrishna_BiotechProg} for an anaerobic culture of \textit{E. coli}. 
+HCM FBA had comparable fits to HCM EM on the fermentation data and a slightly better fit on lactate measurements. 
+Finally, HCM was applied to a larger metabolic network where 29 FBA modes were computed, in contrast to over 66,000 elementary modes.  
+HCM FBA was shown to have robust model performance with only requiring 5 of the 29 FBA modes (selected by the sensitivity analysis) to fit experimental observations.
 
-Second paragraph, contrast what we have done in the context of the field, in particular compare the hybrid cybernetic models, constraints based models and dynamic models of
-Palsson. Why is this study innovative?
+The hybrid cybernetic model is formulated to maximize substrate uptake based on cybernetic arguments. 
+Towards this objective, HCM uses a combination of modes to describe both the external and internal fluxes. 
+The internal fluxes still follow the pseudo steady state approximation just as in FBA, but are a combination of solutions.
+HCM EM has been shown to have comparable internal flux estimates to MFA results\cite{2008_kim_varner_ramkrishna_BiotechProg}, this has not been shown for HCM FBA.
+The HCM framework has been shown to predict dynamic external fluxes using both elementary and FBA modes.
+While DFBA already has the capacity to estimate dynamic external fluxes, it requires \textit{a priori} knowledge and boolean rules to capture the diauxic phenomena \cite{1994_varma_palsson_ApplEnvMicro,2002_Mahadevan_BiophysJ,2001_covert_schilling_palsson}.
+HCM overcomes this by incorparting the regulatory processes of substrate utilization based on its cybernetic arguments. 
+HCM EM has good model performance for reduced networks, but is impractical for larger networks.
+HCM EM has been applied to a network with 67 reactions, requiring the elementary modes to be lumped into groups by complex weighting schemes and relying on experimental data\cite{2010_song_ramkrishna}.
+Elementary mode calculations are still computational expensive and and do not scale well for larger networks\cite{2004_lee_varner_ko_ieee}. 
+This is a shortcoming of the HCM EM framework.
+In contrast, FBA does not have the computational burden associated with calculating elementary modes.  
+FBA is frequently used to study genome-scale networks\cite{2010_orth_NatBiotech} and can be used to generate modes for the HCM framework. 
+This opens up the possibility of genome scale cybernetic models.
 
-Last paragraph, what is wrong with this study and where do we go from here? For example,
-could we extend the work done here to animal cells producing biologics, could we extend this to genome scale.
+HCM FBA provides a practical and feasible approach to model large networks where elementary modes fail.
+The disadvantage of HCM FBA is it requires \textit{a priori} knowledge of the inputs into the metabolic network.
+FBA modes do not have the thorough information as elementary modes, but provide sufficient pathways to model experimental observations.
+%HCM FBA may also be applied to mammalian networks, as long as FBA accounts for compartmentalization within mammalian systems.
 
@@
 \noindent\subsubsection*{Description of the method}
-Describe the method mathematics, use index form if possible, I the vector nomenclature to be hard to follow. Define all terms and symbols.
+The model was constructed using the hybrid cybernetic approach of Kim et al.\cite{2008_kim_varner_ramkrishna_BiotechProg}. 
+The abundance of extracellular species $i$ ($x_{i}$) is governed by:
 
-\subsubsection*{Estimation of model parameters Modify this section}
-Model parameters were estimated by minimizing the difference between simulations and experimental thrombin measurements (squared residual):
+\begin{equation}
+	\frac{dx_{i}}{dt}  =  \sum_{j = 1}^{\mathcal{R}}\sum_{l = 1}^{\mathcal{L}}\sigma_{ij}z_{jl}r_{l}\left(\mathbf{e},\mathbf{k},\mathbf{s}\right)c \qquad{i=1,\hdots,\mathcal{M}} 
+\end{equation}
+
+where $\mathcal{R}$ and $\mathcal{M}$ denotes the number of reactions and extracellular species in the model, and $\mathcal{L}$ denotes the number of modes.
+The quantity $\sigma_{ij}$ denotes the stoichiometric coefficient for species $i$ in reaction $j$ and
+$z_{jl}$ denotes the normalized flux for reaction $j$ in mode $l$.
+If $\sigma_{ij}>0$, species $i$ is produced by reaction $j$, if $\sigma_{ij}<0$, species $i$ is consumed by reaction $j$, while $\sigma_{ij} = 0$ indicates species $i$ is not connected with reaction $j$.
+Extracellular species balances were subject to the initial conditions $\mathbf{x}\left(t_{o}\right) = \mathbf{x}_{o}$ determined from experimental data.
+The $r_{l}\left(\mathbf{e},\mathbf{k},\mathbf{s}\right)$ term denotes the regulated uptake rate of mode $k$, and was written as the product of a kinetic term ($\bar{r}_{l}$) and a cybernetic control variable for enzyme activity ($v_{l}$), $r_{l}\left(\mathbf{e},\mathbf{k},\mathbf{s}\right) = \bar{r}_{l}v_{l}$, where $\mathbf{e}$ denotes the enzyme level for each mode which become additional state variables, $\mathbf{k}$ ($\mathcal{K}\times{1}$) denotes the unknown kinetic parameter vector, and $\mathbf{s}$ denotes the substrate utilized.
+The reaction rate followed Michaelis-Menten kinetics.
+The abundance of the enzyme species is governed by:
+
+\begin{equation}
+	\frac{de_{l}}{dt}  = \alpha + r_{El}u_{l} - \left(\beta+r_{G}\right)e_{l}
+\end{equation}
+
+where $\alpha$ denotes the constant background enzyme synthesis, $r_{El}$ denotes the enzyme synthesis reaction and $u_{l}$ denotes the cybernetic control variable for enzyme synthesis.
+In this study, the enzyme synthesis reaction follows Michealis-Menten kinetics.
+The $\beta$ term denotes the degradation rate of enzyme, and $r_{G}$ denotes the growth rate through all modes.
+All initial conditions of the enzyme species were set to 0.9 for the anaerobic case and 0.8 for the aerobic case.
+
+\begin{equation}
+	r_{G}  = \sum_{l = 1}^{\mathcal{L}}z_{\mu l}r_{l}\left(\mathbf{e},\mathbf{k},\mathbf{s}\right)
+\end{equation}
+
+where $z_{\mu l}$ denotes the growth flux $\mu$ through mode $l$. 
+The cybernetic control variables, $u_{l}$ and $v_{l}$, control the synthesis and activity of each enzyme for each mode using the matching and proportional laws\cite{2007_young_ramkrishna_BiotechProg}.
+They are formulated to maximize the overall substrate uptake rate through each mode.
+
+\begin{equation}
+	u_{l}  = \frac{z_{s,l}\bar{r}_{l}}{\sum_{l = 1}^{\mathcal{L}}z_{sl}\bar{r}_{l}} 
+\end{equation}
+\begin{equation}
+	v_{l} = \frac{z_{s,l}\bar{r}_{l}}{\max_{l=1} z_{sl}\bar{r}_{l}}
+\end{equation}
+
+where $z_{sl}$ denotes the uptake flux of substrate $s$ through mode $l$. 
+In the anaerobic case we followed the assumption of Young's thesis\cite{2005_Young} and Kim et al.\cite{2008_kim_varner_ramkrishna_BiotechProg} that formate decomposition occurs only outside the network (applicable for strain GJT001).
+Therefore its enzyme activity was assumed to be at its maximum level of unity. 
+The reaction rate for formate was also modified following Kim et al.\cite{2008_kim_varner_ramkrishna_BiotechProg}
+
+\subsubsection*{Elementary Flux Mode Analysis}
+The elementary flux modes were calculated using METATOOL 5.1\cite{2006_vonKamp_Metatool} for all networks. 
+The input file requires the conditions for reversibility/irreversibility reactions and intracellular/extracellular metabolites to be specified.
+
+\subsubsection*{Flux Balance Analysis}
+Flux balance analysis determines an optimal flux vector $\mathbf{w}$ subject to the stoichiometric matrix and flux constraints that maximizes/minimize a specified objective. 
+FBA modes were defined as the solution flux vector through the network.
+FBA is formulated by the follwoing:
+
+\begin{equation}
+ \begin{multlined}
+	\qquad \qquad \qquad \max_{\boldsymbol{w}}{} \! \left( w_{obj} = \boldsymbol{c}^T \boldsymbol{w} \right) \\
+	\mathrm{Subject \; to:}
+	 \; \; \mathbf{S}\mathbf{w}=0 \mathrm{\; and \;} \mathbf{S_{x}}\mathbf{w} = \mathbf{b} \\ 
+\alpha_i \leq w_i \leq \beta_i  \qquad
+ \end{multlined}
+\end{equation}
+
+%\begin{alignat}{3}
+%	\max_{\boldsymbol{w}}{} \! \left( w_{obj} = \boldsymbol{c}^T \boldsymbol{w} \right) \\
+ %	 \mathrm{Subject \; to:}
+%	 \; \;  \mathbf{S}\mathbf{w}=0 \mathrm{\; and \;} \mathbf{S_{x}}\mathbf{w} = \mathbf{b} \\
+%	\alpha_i \leq w_i \leq \beta_i 
+%\end{alignat}
+where $\boldsymbol{c}$ is a vector of weights indicating how each flux contributes to the objective; $\alpha_i$ and $\beta_i$ are the lower and upper limits for the individual flux values, respectively.
+The stoichoimetric matrix $\mathbf{S}$ denotes the intracellular network where intracellular metabolism has a psuedo steady state approximation. 
+$\mathbf{S_{x}}$ denotes the extracellular metabolic network and $\mathbf{b}$ denotes the exchange flux across the cell membrane. 
+For each mode, the objective was set to maximize either the growth flux or each fermentation byproduct flux from a specified substrate.
+Multiple modes were calculated for each objective flux by allowing the oxygen and nitrate uptake rates to be either nonexistent or its maximum.
+In the aerobic case, oxygen and nitrate uptake rates were constrained to allow a maximum flux of 10 mM/gDW$\cdot$hr and 0.05 mM/gDW$\cdot$hr, respectively. 
+Each flux vector was normalized by the growth flux, otherwise by the specified objective flux.
+
+\subsubsection*{Sensitivity Analysis}
+The variance based sensitivity analysis\cite{2010_saltelli} was done using SALib, an open source Python library.
+The sensitivity analysis identified key parameters that were influential on our specified model output of biomass yield.
+All model parameters used in the Sobol sensitivity analysis varied between 50\% lower and higher than the optimal values obtained from parameter identification, except enzyme initital conditions which varied between 0 and 1.
+The sample space was set to 1000 and 182,000 parameter sets were generated from the pseudo-random Sobol sequence.
+
+\subsubsection*{Estimation of model parameters}
+Model parameters were estimated by minimizing the difference between simulations and experimental measurements (squared residual):
 \begin{equation}\label{eqn:objective-function}
@@
 value for species $j$ at time $\tau$, and $\omega_{j}\left(\tau\right)$ denotes the experimental measurement variance for species $j$ at time $\tau$.
-The outer summation is with respect to time, while the inner summation is with respect to state.
-We minimized the model residual using Particle swarm optimization (PSO) \cite{PSO}.
-PSO uses a \textit{swarming} metaheuristic to explore parameter spaces.
-For each iteration, particles in the swarm compute their local error by evaluating the model equations using their specific parameter vector realization.
-From each of these local points, a globally best error is identified. Both the local and global error
-are then used to update the parameter estimates of each particle using the rules:
-\begin{eqnarray}
-	\mathbf{\Delta}_{i} &=&\theta_{1}\mathbf{\Delta}_{i} + \theta_{2}\mathbf{r}_{1}\left(\mathcal{L}_{i} - \mathbf{k}_{i}\right) + \theta_{3}\mathbf{r}_{2}\left(\mathcal{G} - \mathbf{k}_{i}\right) \\
-	\mathbf{k}_{i} &=& \mathbf{k}_{i} + \mathbf{\Delta}_{i}
-\end{eqnarray}where $\left(\theta_{1},\theta_{2},\theta_{3}\right)$ are adjustable parameters, $\mathcal{L}_{i}$ denotes the local best solution found by particle $i$, and
-$\mathcal{G}$ denotes the best solution found over the entire population of particles.
-The quantities $r_{1}$ and $r_{2}$ denote uniform random vectors with the same dimension as the number of unknown model
-parameters ($\mathcal{K}\times{1}$). In thus study, we used $\left(\theta_{1},\theta_{2},\theta_{3}\right) = \left(1.0, 0.05564, 0.02886\right)$. The quality of parameter
-estimates was measured using goodness of fit (model residual). The particle swarm optimization routine was implemented in the Python programming language.
+The outer summation is with respect to time, while the inner summation is with respect to state. We minimized the model residual using simulated annealing.
 

### 2016-03-09 Paper_v1.tex
Draft1
Initial JV edits

@@
 \section{Introduction}
-Metabolic networks for simple organisms such as \textit{E. coli} contain throusands of reactions that are highly interconnected. 
+Metabolic networks for simple organisms such as \textit{E. coli} contain thousands of reactions that are highly interconnected.
 These organisms have the potential to be engineered to produce valuable products.
@@
 Traditionally, Monod kinetics have been used as unstructured models to describe cell growth, substrate consumption, and product formation, but fail to account for intracellular metabolism\cite{shuler_book}.
-Structured modeling approaches have made great progress to describe metabolic pathways including elementary mode analysis (EM)\cite{2006_vonKamp_Metatool}, metabolic flux analysis (MFA), flux balance analysis (FBA)\cite{2010_orth_NatBiotech}, dynamic FBA\cite{1994_varma_palsson_ApplEnvMicro,2002_Mahadevan_BiophysJ}, and hybrid cybernetic modeling (HCM)\cite{2008_kim_varner_ramkrishna_BiotechProg}.
+Structured modeling approaches have made great progress to describe metabolic pathways including elementary mode analysis (EM)\cite{2006_vonKamp_Metatool}, metabolic flux analysis (MFA), flux balance analysis (FBA)\cite{2010_orth_NatBiotech}, dynamic FBA \cite{1994_varma_palsson_ApplEnvMicro,2002_Mahadevan_BiophysJ}, and hybrid cybernetic modeling (HCM) \cite{2008_kim_varner_ramkrishna_BiotechProg}.
 
-Constraint based models of metabolic pathways, such as EM, FBA and MFA, attempt to describe intracellular metabolism using the stoichiometry of biochemical networks and a pseudo-steady state assumption\cite{2010_orth_NatBiotech}.
+Convex and constraints based analysis methods, such as EM, FBA and MFA,
+describe intracellular metabolism using the stoichiometry of the biochemical networks and the pseudo-steady state assumption \cite{2010_orth_NatBiotech}.
 EM analysis is a powerful tool to decompse a metabolic network into independent minimal pathways that are stoihiometrically and thermodynamically feasible.
@@
 \section{Results}
-In the general case model, we demonstrated that the performance of the HCM FBA was equivalent to that of HCM EM.
+The performance of HCM FBA was equivalent to HCM EM for a proof of concept metabolic network (Fig. \ref{fig:model-fitting}).
 In the proof of concept metabolic network consisting of 6 metabolites and 7 reactions (Fig. \ref{fig:model-fitting}A), we calculated 6 elementary modes and 3 FBA modes.
@@
 
-In the anaerobic \textit{E. coli} case, we demonstrated that HCM FBA has comparable model performance to a published HCM EM model\cite{2008_kim_varner_ramkrishna_BiotechProg}.
+The performance of HCM FBA was equivalent to HCM EM for a small model of anaerobic \textit{E. coli} metabolism (Fig. \ref{fig:ecoli}, left).
+In the  case, we demonstrated that  has comparable model performance to a published HCM EM model \cite{2008_kim_varner_ramkrishna_BiotechProg}.
 The metabolic network consisted of 12 reactions, where we calculated 9 elementary modes and 7 FBA modes.
@@
 
-In the aerobic \textit{E. coli} case, we demonstrated that HCM FBA is applicable for larger metabolic networks where elementary mode calculations increase exponentially.
+HCM FBA outperformed HCM EM for a large model of aerobic \textit{E. coli} metabolism (Fig. \ref{fig:ecoli}, right).
+In the  case, we demonstrated that HCM FBA is applicable for larger metabolic networks where elementary mode calculations increase exponentially.
 A metabolic network\cite{2007_schuetz_etal_MolSysBio} of 98 reactions and 60 metabolites with the biomass reaction from Palsson's \textit{E. coli} core model\cite{2006_Palsson_model} was used.
@@
 
-A variance based sensitivity analysis on the aerobic \textit{E. coli} case identified essential modes that were influential on biomass yield. 
-The analysis calculated the total order sensitivity indicies of all the kinetic parameters and enzyme initial conditions (Fig. \ref{fig:sensitivity}).
+Variance based sensitivity of aerobic \textit{E. coli} metabolism identified modes that were essential to biomass yield (Fig. \ref{fig:sensitivity}).
+The analysis calculated the total order sensitivity indicies of all the kinetic parameters and enzyme initial conditions.
 We used the sensitivity results as a simple systematic method to eliminate FBA modes to give the smallest number needed while maintaining good model performance.
@@
 \begin{figure*}[!t]\centering
-\includegraphics[width=0.95\textwidth]{./figs/Fig-3-Sensitivity-Results.pdf}
+\includegraphics[width=0.85\textwidth]{./figs/Fig-3-Sensitivity-Results.pdf}
 \caption{Total order variance based sensitivity analysis on biomass yield from glucose and acetate.
@@
 
-\noindent\subsubsection*{Description of the method}
+\noindent\subsubsection*{HCM FBA method.}
 The model was constructed using the hybrid cybernetic approach of Kim et al.\cite{2008_kim_varner_ramkrishna_BiotechProg}.
 The abundance of extracellular species $i$ ($x_{i}$) is governed by:
-
 \begin{equation}
@@
 \end{equation}
-
 where $\mathcal{R}$ and $\mathcal{M}$ denotes the number of reactions and extracellular species in the model, and $\mathcal{L}$ denotes the number of modes.
@@
 \begin{equation}
-	\frac{de_{l}}{dt}  = \alpha + r_{El}u_{l} - \left(\beta+r_{G}\right)e_{l}
+	\frac{de_{l}}{dt}  = \alpha_{l} + r_{E,l}u_{l} - \left(\beta_{l}+r_{G}\right)e_{l}
 \end{equation}
-
-where $\alpha$ denotes the constant background enzyme synthesis, $r_{El}$ denotes the enzyme synthesis reaction and $u_{l}$ denotes the cybernetic control variable for enzyme synthesis.
-In this study, the enzyme synthesis reaction follows Michealis-Menten kinetics.
-The $\beta$ term denotes the degradation rate of enzyme, and $r_{G}$ denotes the growth rate through all modes.
-All initial conditions of the enzyme species were set to 0.9 for the anaerobic case and 0.8 for the aerobic case.
+where $\alpha_{l}$ denotes the specific rate of constitutive enzyme synthesis for enzyme l, $r_{E,l}$ denotes the specific rate enzyme synthesis for enzyme l, and $u_{l}$ denotes the cybernetic variable controlling the synthesis of enzyme l.
+The term $\beta_{l}$ denotes the rate constant governing enzyme degradation, and $r_{G}$ denotes the growth rate through all modes.
+The specific rate of enzyme synthesis was modeled using saturation kinetics.
+All enzyme initial conditions were set to 0.9 for the anaerobic case, and 0.8 for the aerobic case.
 
@@
 \end{equation}
-
 where $z_{\mu l}$ denotes the growth flux $\mu$ through mode $l$.
@@
 \end{equation}
-
 where $z_{sl}$ denotes the uptake flux of substrate $s$ through mode $l$.
@@
 \end{equation}
-
-%\begin{alignat}{3}
-%	\max_{\boldsymbol{w}}{} \! \left( w_{obj} = \boldsymbol{c}^T \boldsymbol{w} \right) \\
- %	 \mathrm{Subject \; to:}
-%	 \; \;  \mathbf{S}\mathbf{w}=0 \mathrm{\; and \;} \mathbf{S_{x}}\mathbf{w} = \mathbf{b} \\
-%	\alpha_i \leq w_i \leq \beta_i 
-%\end{alignat}
 where $\boldsymbol{c}$ is a vector of weights indicating how each flux contributes to the objective; $\alpha_i$ and $\beta_i$ are the lower and upper limits for the individual flux values, respectively.
@@
 \end{document}
+\grid

### 2016-03-09 Paper_v1.tex
Draft2

@@
 \usepackage{mathtools}
-\usepackage{wrapfig}
-
-\bibliographystyle{unsrt}
 
@@
 HCM FBA was applied to a larger metabolic network where 29 FBA modes were computed, in contrast to over 66,000 elementary modes.
-HCM FBA had robust model performance to fit experimental observations with only 5 of the 29 FBA modes.
+A sensitivity analysis was performed to reduce the number of FBA modes from 29 to 5.
+This reduction maintained a robust model performance to fit experimental observations.
 
@@
 Traditionally, Monod kinetics have been used as unstructured models to describe cell growth, substrate consumption, and product formation, but fail to account for intracellular metabolism\cite{shuler_book}.
-Structured modeling approaches have made great progress to describe metabolic pathways including elementary mode analysis (EM)\cite{2006_vonKamp_Metatool}, metabolic flux analysis (MFA), flux balance analysis (FBA)\cite{2010_orth_NatBiotech}, dynamic FBA \cite{1994_varma_palsson_ApplEnvMicro,2002_Mahadevan_BiophysJ}, and hybrid cybernetic modeling (HCM) \cite{2008_kim_varner_ramkrishna_BiotechProg}.
+Structured modeling approaches have made great progress to describe metabolic pathways including elementary mode analysis (EM)\cite{2006_vonKamp_Metatool}, metabolic flux analysis (MFA), flux balance analysis (FBA)\cite{2010_orth_NatBiotech}, dynamic FBA (DFBA) \cite{1994_varma_palsson_ApplEnvMicro,2002_Mahadevan_BiophysJ}, and hybrid cybernetic modeling (HCM) \cite{2008_kim_varner_ramkrishna_BiotechProg}.
 
@@
 DFBA is able to account for dynamic external fluxes but fails to incorporate the regulatory processes involved with HCM without the use of boolean rules\cite{2001_covert_schilling_palsson}.
-Despite these advantages of HCM, it is restricted to small networks due to the exponential increase of the number of elementary modes for large networks.
+Despite the advantages of HCM, it is restricted to small networks due to the exponential increase of the number of elementary modes for large networks.
 Recent work has been done to reduce the number of elementary modes by lumping modes into groups using complex weighting schemes and experimental data\cite{2010_song_ramkrishna}, however, HCM is still limited by EM decomposition.
@@
 We show HCM FBA has comparable model performance to HCM EM for a hypothetical metabolic network and a reduced anaerobic \textit{E. coli} network.
-HCM was then applied to a large metabolic network where the number of elementary modes made it impractical to model with HCM EM.
-However, HCM FBA was still feasible for the network and performed well with respect to experimental observations.
+HCM EM and HCM FBA were then applied to a large metabolic network. 
+The number of elementary modes made it impractical to model with HCM EM, but HCM FBA was computationally feasible and performed well with respect to experimental observations.
 Taken together, HCM FBA has comparable model performace to HCM EM and may be potentially applied to genome-scale networks which was a shortcoming of HCM EM.
@@
 The performance of HCM FBA was equivalent to HCM EM for a small model of anaerobic \textit{E. coli} metabolism (Fig. \ref{fig:ecoli}, left).
-In the  case, we demonstrated that  has comparable model performance to a published HCM EM model \cite{2008_kim_varner_ramkrishna_BiotechProg}.
+In the case, we demonstrated that HCM FBA has comparable model performance to a published HCM EM model \cite{2008_kim_varner_ramkrishna_BiotechProg}.
 The metabolic network consisted of 12 reactions, where we calculated 9 elementary modes and 7 FBA modes.
@@
 We calculated over 66,000 elementary modes for this network and determined it was futile to model.
-For the same network, FBA computed 29 modes.
+For the same network, we computed 29 FBA modes.
 HCM FBA was simulated to fit cell density, glucose and acetate measurements for an aerobic \textit{E. coli} batch culture\cite{1994_varma_palsson_ApplEnvMicro} (Fig.\ref{fig:ecoli}B (solid line)).
@@
 Finally, HCM was applied to a larger metabolic network where 29 FBA modes were computed, in contrast to over 66,000 elementary modes.
-HCM FBA was shown to have robust model performance with only requiring 5 of the 29 FBA modes (selected by the sensitivity analysis) to fit experimental observations.
+HCM FBA was shown to have robust model performance only requiring 5 of the 29 FBA modes (selected by the sensitivity analysis) to fit experimental observations.
 
@@
 Towards this objective, HCM uses a combination of modes to describe both the external and internal fluxes.
-The internal fluxes still follow the pseudo steady state approximation just as in FBA, but are a combination of solutions.
+The internal fluxes still follow the pseudo steady state approximation applied in FBA, but HCM uses a combination of pathways.
 HCM EM has been shown to have comparable internal flux estimates to MFA results\cite{2008_kim_varner_ramkrishna_BiotechProg}, this has not been shown for HCM FBA.
@@
 HCM EM has been applied to a network with 67 reactions, requiring the elementary modes to be lumped into groups by complex weighting schemes and relying on experimental data\cite{2010_song_ramkrishna}.
-Elementary mode calculations are still computational expensive and and do not scale well for larger networks\cite{2004_lee_varner_ko_ieee}.
-This is a shortcoming of the HCM EM framework.
+A shortcoming of the HCM EM framework is that EM calculations are computationally expensive and the number of modes exponentially increases for larger networks \cite{2004_lee_varner_ko_ieee}.
 In contrast, FBA does not have the computational burden associated with calculating elementary modes.
@@
 \begin{equation}
-	\frac{de_{l}}{dt}  = \alpha_{l} + r_{E,l}u_{l} - \left(\beta_{l}+r_{G}\right)e_{l}
+	\frac{de_{l}}{dt}  = \alpha_{l} + r_{E,l}\left(\mathbf{k},\mathbf{s}\right)u_{l} - \left(\beta_{l}+r_{G}\right)e_{l}
 \end{equation}
-where $\alpha_{l}$ denotes the specific rate of constitutive enzyme synthesis for enzyme l, $r_{E,l}$ denotes the specific rate enzyme synthesis for enzyme l, and $u_{l}$ denotes the cybernetic variable controlling the synthesis of enzyme l.
+where $\alpha_{l}$ denotes the specific rate of constitutive enzyme synthesis for enzyme l, $r_{E,l}\left(\mathbf{k},\mathbf{s}\right)$ denotes the specific rate enzyme synthesis for enzyme l, and $u_{l}$ denotes the cybernetic variable controlling the synthesis of enzyme l.
 The term $\beta_{l}$ denotes the rate constant governing enzyme degradation, and $r_{G}$ denotes the growth rate through all modes.
@@
 $\mathbf{S_{x}}$ denotes the extracellular metabolic network and $\mathbf{b}$ denotes the exchange flux across the cell membrane.
+FBA was solved using linear programming by a GNU Linear Programming Kit, GLPK version 4.52.
 For each mode, the objective was set to maximize either the growth flux or each fermentation byproduct flux from a specified substrate.
@@
 In the aerobic case, oxygen and nitrate uptake rates were constrained to allow a maximum flux of 10 mM/gDW$\cdot$hr and 0.05 mM/gDW$\cdot$hr, respectively.
-Each flux vector was normalized by the growth flux, otherwise by the specified objective flux.
+Each flux vector was normalized by the growth flux, or by the specified objective flux.
 
@@
 Model parameters were estimated by minimizing the difference between simulations and experimental measurements (squared residual):
+
 \begin{equation}\label{eqn:objective-function}
 	\min_{\mathbf{k}} \sum_{\tau=1}^{\mathcal{T}}\sum_{j=1}^{\mathcal{S}}\left(\frac{\hat{x}_{j}\left(\tau\right) - x_{j}\left(\tau,\mathbf{k}\right)}{\omega_{j}\left(\tau\right)}\right)^{2}
-\end{equation}where $\hat{x}_{j}\left(\tau\right)$ denotes the measured value of species $j$ at time $\tau$, $x_{j}\left(\tau,\mathbf{k}\right)$ denotes the simulated
+\end{equation}
+
+where $\hat{x}_{j}\left(\tau\right)$ denotes the measured value of species $j$ at time $\tau$, $x_{j}\left(\tau,\mathbf{k}\right)$ denotes the simulated
 value for species $j$ at time $\tau$, and $\omega_{j}\left(\tau\right)$ denotes the experimental measurement variance for species $j$ at time $\tau$.
@@
 % References -
+\bibliographystyle{naturemag_noURL}
 \bibliography{Paper_v1}

### 2016-03-10 Paper_v1.tex
More changes

@@
 \section{Materials and Methods}
-
-\noindent\subsubsection*{HCM FBA method.}
-The model was constructed using the hybrid cybernetic approach of Kim et al.\cite{2008_kim_varner_ramkrishna_BiotechProg}.
-The abundance of extracellular species $i$ ($x_{i}$) is governed by:
-\begin{equation}
-	\frac{dx_{i}}{dt}  =  \sum_{j = 1}^{\mathcal{R}}\sum_{l = 1}^{\mathcal{L}}\sigma_{ij}z_{jl}r_{l}\left(\mathbf{e},\mathbf{k},\mathbf{s}\right)c \qquad{i=1,\hdots,\mathcal{M}}
-\end{equation}
-where $\mathcal{R}$ and $\mathcal{M}$ denotes the number of reactions and extracellular species in the model, and $\mathcal{L}$ denotes the number of modes.
-The quantity $\sigma_{ij}$ denotes the stoichiometric coefficient for species $i$ in reaction $j$ and
-$z_{jl}$ denotes the normalized flux for reaction $j$ in mode $l$.
-If $\sigma_{ij}>0$, species $i$ is produced by reaction $j$, if $\sigma_{ij}<0$, species $i$ is consumed by reaction $j$, while $\sigma_{ij} = 0$ indicates species $i$ is not connected with reaction $j$.
+The HCM FBA approach is a modification of the HCM EM strategy of Kim et al.
+\cite{2008_kim_varner_ramkrishna_BiotechProg}.
+However, unlike HCM EM, we replaced elementary modes with flux balance analysis solutions.
+The abundance of extracellular species $i$ ($x_{i}$), the pseudo enzyme $e_{l}$ and cellmass are governed by:
+\begin{eqnarray}\nonumber
+	\frac{dx_{i}}{dt}  & = &  \sum_{j = 1}^{\mathcal{R}}\sum_{l = 1}^{\mathcal{L}}\sigma_{ij}z_{jl}r_{l}\left(\mathbf{e},\mathbf{k},\mathbf{x}\right)c \qquad{i=1,\hdots,\mathcal{M}}\\\nonumber
+  \frac{de_{l}}{dt}  & = & \alpha_{l} + r_{E,l}\left(\mathbf{k},\mathbf{x}\right)u_{l} - \left(\beta_{l}+r_{G}\right)e_{l} \qquad l=1,\hdots,\mathcal{L} \\\nonumber
+  \frac{dc}{dt} & = & r_{G}c
+\end{eqnarray}
+where $\mathcal{R}$ and $\mathcal{M}$ denote the number of reactions and extracellular species in the model, and $\mathcal{L}$ denotes the number of FBA modes.
+The quantity $\sigma_{ij}$ denotes the stoichiometric coefficient for species $i$ in reaction $j$ and $z_{jl}$ denotes the normalized flux for reaction $j$ in mode $l$.
+If $\sigma_{ij}>0$, species $i$ is produced by reaction $j$,
+if $\sigma_{ij}<0$, species $i$ is consumed by reaction $j$, while $\sigma_{ij} = 0$ indicates species $i$ is not connected with reaction $j$.
 Extracellular species balances were subject to the initial conditions $\mathbf{x}\left(t_{o}\right) = \mathbf{x}_{o}$ determined from experimental data.
-The $r_{l}\left(\mathbf{e},\mathbf{k},\mathbf{s}\right)$ term denotes the regulated uptake rate of mode $k$, and was written as the product of a kinetic term ($\bar{r}_{l}$) and a cybernetic control variable for enzyme activity ($v_{l}$), $r_{l}\left(\mathbf{e},\mathbf{k},\mathbf{s}\right) = \bar{r}_{l}v_{l}$, where $\mathbf{e}$ denotes the enzyme level for each mode which become additional state variables, $\mathbf{k}$ ($\mathcal{K}\times{1}$) denotes the unknown kinetic parameter vector, and $\mathbf{s}$ denotes the substrate utilized.
-The reaction rate followed Michaelis-Menten kinetics.
-The abundance of the enzyme species is governed by:
-
-\begin{equation}
-	\frac{de_{l}}{dt}  = \alpha_{l} + r_{E,l}\left(\mathbf{k},\mathbf{s}\right)u_{l} - \left(\beta_{l}+r_{G}\right)e_{l}
-\end{equation}
-where $\alpha_{l}$ denotes the specific rate of constitutive enzyme synthesis for enzyme l, $r_{E,l}\left(\mathbf{k},\mathbf{s}\right)$ denotes the specific rate enzyme synthesis for enzyme l, and $u_{l}$ denotes the cybernetic variable controlling the synthesis of enzyme l.
+The term $r_{l}\left(\mathbf{e},\mathbf{k},\mathbf{x}\right)$ denotes the specific rate of flux through mode $l$, and was written as the product of a kinetic term ($\bar{r}_{l}$) and a cybernetic control variable governing enzyme activity.
+Flux through each mode was catalyzed by a pseudo enzyme $e_{l}$, where enzyme $e_{l}$ was synthesized at the regulated specific rate $r_{E,l}\left(\mathbf{k},\mathbf{s}\right)$ and constitutively at the rate $\alpha_{l}$.
+The term $r_{E,l}$ denotes the specific rate enzyme synthesis for enzyme l, and $u_{l}$ denotes the cybernetic variable controlling the synthesis of enzyme l.
 The term $\beta_{l}$ denotes the rate constant governing enzyme degradation, and $r_{G}$ denotes the growth rate through all modes.
-The specific rate of enzyme synthesis was modeled using saturation kinetics.
+The specific rate of flux through an FBA mode, and the specific rate of enzyme synthesis were modeled using saturation kinetics.
 All enzyme initial conditions were set to 0.9 for the anaerobic case, and 0.8 for the aerobic case.
-
+Lastly, cellmass was produced at the specific growth rate:
 \begin{equation}
@@
 where $z_{\mu l}$ denotes the growth flux $\mu$ through mode $l$.
-The cybernetic control variables, $u_{l}$ and $v_{l}$, control the synthesis and activity of each enzyme for each mode using the matching and proportional laws\cite{2007_young_ramkrishna_BiotechProg}.
-They are formulated to maximize the overall substrate uptake rate through each mode.
-
-\begin{equation}
-	u_{l}  = \frac{z_{s,l}\bar{r}_{l}}{\sum_{l = 1}^{\mathcal{L}}z_{sl}\bar{r}_{l}}
-\end{equation}
-\begin{equation}
-	v_{l} = \frac{z_{s,l}\bar{r}_{l}}{\max_{l=1} z_{sl}\bar{r}_{l}}
-\end{equation}
-where $z_{sl}$ denotes the uptake flux of substrate $s$ through mode $l$.
-In the anaerobic case we followed the assumption of Young's thesis\cite{2005_Young} and Kim et al.\cite{2008_kim_varner_ramkrishna_BiotechProg} that formate decomposition occurs only outside the network (applicable for strain GJT001).
+The cybernetic control variables $u_{l}$ and $v_{l}$, which control the synthesis and activity of each enzyme respectively, were given by:
+\begin{align*}
+	u_{l}  = \frac{z_{s,l}\bar{r}_{l}}{\sum\limits_{l = 1}^{\mathcal{L}}z_{sl}\bar{r}_{l}} && v_{l} = \frac{z_{s,l}\bar{r}_{l}}{\max\limits_{l=1} z_{sl}\bar{r}_{l}}
+\end{align*}
+where $z_{sl}$ denotes the uptake flux of substrate $x$ through mode $l$.
+In the anaerobic case we followed the assumption of Kim et al.\cite{2008_kim_varner_ramkrishna_BiotechProg},
+that formate decomposition occurs only outside the network (applicable for strain GJT001).
 Therefore its enzyme activity was assumed to be at its maximum level of unity.
@@
 
-\subsubsection*{Elementary Flux Mode Analysis}
-The elementary flux modes were calculated using METATOOL 5.1\cite{2006_vonKamp_Metatool} for all networks.
-The input file requires the conditions for reversibility/irreversibility reactions and intracellular/extracellular metabolites to be specified.
-
-\subsubsection*{Flux Balance Analysis}
-Flux balance analysis determines an optimal flux vector $\mathbf{w}$ subject to the stoichiometric matrix and flux constraints that maximizes/minimize a specified objective.
-FBA modes were defined as the solution flux vector through the network.
-FBA is formulated by the follwoing:
-
+\noindent\subsubsection*{Elementary Mode and Flux Balance Analysis}
+Elementary modes were calculated using METATOOL 5.1 \cite{2006_vonKamp_Metatool}.
+FBA modes were defined as the solution flux vector through the network connecting substrate uptake to cellmass and extracellular product formation.
+The FBA problem was formulated as:
 \begin{equation}
  \begin{multlined}
-	\qquad \qquad \qquad \max_{\boldsymbol{w}}{} \! \left( w_{obj} = \boldsymbol{c}^T \boldsymbol{w} \right) \\
+	\qquad \qquad \qquad \max_{\boldsymbol{w}}{} \! \left( w_{obj} = \mathbf{\theta}^T \boldsymbol{w} \right) \\
 	\mathrm{Subject \; to:}
-	 \; \; \mathbf{S}\mathbf{w}=0 \mathrm{\; and \;} \mathbf{S_{x}}\mathbf{w} = \mathbf{b} \\
+	 \; \; \mathbf{S}\mathbf{w}=\mathbf{0} \\
 \alpha_i \leq w_i \leq \beta_i  \qquad
@@
 \end{equation}
-where $\boldsymbol{c}$ is a vector of weights indicating how each flux contributes to the objective; $\alpha_i$ and $\beta_i$ are the lower and upper limits for the individual flux values, respectively.
-The stoichoimetric matrix $\mathbf{S}$ denotes the intracellular network where intracellular metabolism has a psuedo steady state approximation.
-$\mathbf{S_{x}}$ denotes the extracellular metabolic network and $\mathbf{b}$ denotes the exchange flux across the cell membrane.
-FBA was solved using linear programming by a GNU Linear Programming Kit, GLPK version 4.52.
-For each mode, the objective was set to maximize either the growth flux or each fermentation byproduct flux from a specified substrate.
-Multiple modes were calculated for each objective flux by allowing the oxygen and nitrate uptake rates to be either nonexistent or its maximum.
-In the aerobic case, oxygen and nitrate uptake rates were constrained to allow a maximum flux of 10 mM/gDW$\cdot$hr and 0.05 mM/gDW$\cdot$hr, respectively.
+where $\mathbf{S}$ denotes the stoichiometric matrix, $\mathbf{w}$ denotes the unknown flux vector, $\boldsymbol{\theta}$ denotes the objective selection vector
+and $\alpha_i$ and $\beta_i$ denote the lower and upper bounds on flux $w_{i}$, respectively.
+The flux balance analysis problem was solved using the GNU Linear Programming Kit, GLPK version 4.52 [REF].
+For each FBA mode, the objective was to maximize either the growth rate, or the specific rate of byproduct formation from a specified starting substrate.
+Multiple FBA modes were calculated for each objective flux by allowing the oxygen and nitrate uptake rates to be either zero or maximal.
+For aerobic metabolism, the specific oxygen and nitrate uptake rates were constrained to allow a maximum flux of 10 mM/gDW$\cdot$hr and 0.05 mM/gDW$\cdot$hr, respectively.
 Each flux vector was normalized by the growth flux, or by the specified objective flux.
 
-\subsubsection*{Sensitivity Analysis}
-The variance based sensitivity analysis\cite{2010_saltelli} was done using SALib, an open source Python library.
-The sensitivity analysis identified key parameters that were influential on our specified model output of biomass yield.
-All model parameters used in the Sobol sensitivity analysis varied between 50\% lower and higher than the optimal values obtained from parameter identification, except enzyme initital conditions which varied between 0 and 1.
-The sample space was set to 1000 and 182,000 parameter sets were generated from the pseudo-random Sobol sequence.
+\subsubsection*{Global Sensitivity Analysis}
+Variance based sensitivity analysis was used to estimate which FBA modes were critical to model performance \cite{2010_saltelli}.
+The performance function used in this study was the biomass yield on substrate.
+Candidate parameter sets (N = 182,000) were generated using Sobol sampling by perturbing the best fit parameter set $\pm50\%$ [REF].
+Model performance, calculated for each of these parameter sets, was then used to estimate the total-order sensitivity coefficient for each model parameter.
 
@@
 \end{equation}
-
 where $\hat{x}_{j}\left(\tau\right)$ denotes the measured value of species $j$ at time $\tau$, $x_{j}\left(\tau,\mathbf{k}\right)$ denotes the simulated
 value for species $j$ at time $\tau$, and $\omega_{j}\left(\tau\right)$ denotes the experimental measurement variance for species $j$ at time $\tau$.
-The outer summation is with respect to time, while the inner summation is with respect to state. We minimized the model residual using simulated annealing.
+The outer summation is with respect to time, while the inner summation is with respect to state. The model residual was minimized using the simulated annealing routine
+in the JuMP package [REF].
 

### 2016-03-10 Paper_v1.tex
update_MV

@@
 describe intracellular metabolism using the stoichiometry of the biochemical networks and the pseudo-steady state assumption \cite{2010_orth_NatBiotech}.
-EM analysis is a powerful tool to decompse a metabolic network into independent minimal pathways that are stoihiometrically and thermodynamically feasible.
+EM analysis is a powerful tool to decompse a metabolic network into independent minimal pathways that are stoichiometrically and thermodynamically feasible.
 However, elementary mode calculations are computational expensive and infeasible for genome scale networks\cite{2004_lee_varner_ko_ieee}.
@@
 In the case, we demonstrated that HCM FBA has comparable model performance to a published HCM EM model \cite{2008_kim_varner_ramkrishna_BiotechProg}.
-The metabolic network consisted of 12 reactions, where we calculated 9 elementary modes and 7 FBA modes.
+The metabolic network consisted of 12 reactions and 19 metabolites, where we calculated 9 elementary modes and 7 FBA modes.
 The simulated HCM EM concentration profiles of biomass, substrate and fermentation products (Fig.
@@
 HCM FBA has 17 kinetic parameters while HCM EM has 21 kinetic parameters and both models resulted in similar fits to the experimental data (Fig. \ref {fig:ecoli}A).
-HCM FBA had a slightly better fit on lactate than HCM EM.
+HCM FBA had a better fit on lactate than HCM EM by 35.6\% on the residual standard error. 
+However, HCM EM had a lower residual standard error on the overall performance by 4.5\%.
 
@@
 The term $r_{l}\left(\mathbf{e},\mathbf{k},\mathbf{x}\right)$ denotes the specific rate of flux through mode $l$, and was written as the product of a kinetic term ($\bar{r}_{l}$) and a cybernetic control variable governing enzyme activity.
-Flux through each mode was catalyzed by a pseudo enzyme $e_{l}$, where enzyme $e_{l}$ was synthesized at the regulated specific rate $r_{E,l}\left(\mathbf{k},\mathbf{s}\right)$ and constitutively at the rate $\alpha_{l}$.
-The term $r_{E,l}$ denotes the specific rate enzyme synthesis for enzyme l, and $u_{l}$ denotes the cybernetic variable controlling the synthesis of enzyme l.
+Flux through each mode was catalyzed by a pseudo enzyme $e_{l}$, where enzyme $e_{l}$ was synthesized at the regulated specific rate $r_{E,l}\left(\mathbf{k},\mathbf{x}\right)$ and constitutively at the rate $\alpha_{l}$.
+The term $r_{E,l}$ denotes the specific rate enzyme synthesis for enzyme $l$, and $u_{l}$ denotes the cybernetic variable controlling the synthesis of enzyme $l$.
 The term $\beta_{l}$ denotes the rate constant governing enzyme degradation, and $r_{G}$ denotes the growth rate through all modes.
@@
 \begin{equation}
-	r_{G}  = \sum_{l = 1}^{\mathcal{L}}z_{\mu l}r_{l}\left(\mathbf{e},\mathbf{k},\mathbf{s}\right)
+	r_{G}  = \sum_{l = 1}^{\mathcal{L}}z_{\mu l}r_{l}\left(\mathbf{e},\mathbf{k},\mathbf{x}\right)
 \end{equation}
@@
 \begin{align*}
-	u_{l}  = \frac{z_{s,l}\bar{r}_{l}}{\sum\limits_{l = 1}^{\mathcal{L}}z_{sl}\bar{r}_{l}} && v_{l} = \frac{z_{s,l}\bar{r}_{l}}{\max\limits_{l=1} z_{sl}\bar{r}_{l}}
+	u_{l}  = \frac{z_{sl}\bar{r}_{l}}{\sum\limits_{l = 1}^{\mathcal{L}}z_{sl}\bar{r}_{l}} && v_{l} = \frac{z_{sl}\bar{r}_{l}}{\max\limits_{\mathbf{l}} z_{sl}\bar{r}_{l}}
 \end{align*}
@@
 Therefore its enzyme activity was assumed to be at its maximum level of unity.
-The reaction rate for formate was also modified following Kim et al.\cite{2008_kim_varner_ramkrishna_BiotechProg}
+The reaction rate for formate was also modified following Kim et al.\cite{2008_kim_varner_ramkrishna_BiotechProg}.
+All numerical simulations were conducted in Julia version 0.4.2, a high-level, high performance dynamic programming language \cite{Julia}.
+Species balances were solved using a Sundials \cite{Sundials} wrapper for julia.
 
@@
 and $\alpha_i$ and $\beta_i$ denote the lower and upper bounds on flux $w_{i}$, respectively.
-The flux balance analysis problem was solved using the GNU Linear Programming Kit, GLPK version 4.52 [REF].
+The flux balance analysis problem was solved using the GNU Linear Programming Kit, GLPK version 4.52 \cite{GLPK}.
 For each FBA mode, the objective was to maximize either the growth rate, or the specific rate of byproduct formation from a specified starting substrate.
@@
 \subsubsection*{Global Sensitivity Analysis}
-Variance based sensitivity analysis was used to estimate which FBA modes were critical to model performance \cite{2010_saltelli}.
+Variance based sensitivity analysis was used to estimate which FBA modes were critical to model performance.
 The performance function used in this study was the biomass yield on substrate.
-Candidate parameter sets (N = 182,000) were generated using Sobol sampling by perturbing the best fit parameter set $\pm50\%$ [REF].
+Candidate parameter sets (N = 182,000) were generated using Sobol sampling by perturbing the best fit parameter set $\pm50\%$ \cite{SALib}.
 Model performance, calculated for each of these parameter sets, was then used to estimate the total-order sensitivity coefficient for each model parameter.
@@
 value for species $j$ at time $\tau$, and $\omega_{j}\left(\tau\right)$ denotes the experimental measurement variance for species $j$ at time $\tau$.
-The outer summation is with respect to time, while the inner summation is with respect to state. The model residual was minimized using the simulated annealing routine
-in the JuMP package [REF].
+The outer summation is with respect to time, while the inner summation is with respect to state. The model residual was minimized using simulated annealing.
 

### 2016-03-13 Paper_v1.tex
More changes afternoon 3/12

@@
 But HCM has multiple pathways to select from (elementary modes) and can use a combination of modes to direct its resources through the cell.
-HCM has been shown to have comparable internal flux estimations to MFA and is able to estimate dynamic external fluxes with respect to experimental observations \cite{2008_kim_varner_ramkrishna_BiotechProg}.\
+HCM has been shown to have comparable internal flux estimations to MFA and is able to estimate dynamic external fluxes with respect to experimental observations \cite{2008_kim_varner_ramkrishna_BiotechProg}.
 DFBA is able to account for dynamic external fluxes but fails to incorporate the regulatory processes involved with HCM without the use of boolean rules\cite{2001_covert_schilling_palsson}.
@@
 \section{Results}
-The performance of HCM FBA was equivalent to HCM EM for a proof of concept metabolic network (Fig. \ref{fig:model-fitting}).
-In the proof of concept metabolic network consisting of 6 metabolites and 7 reactions (Fig. \ref{fig:model-fitting}A), we calculated 6 elementary modes and 3 FBA modes.
-Using the HCM EM modeling framework, concentration profiles were generated for the extracellular metabolites and artifical biomass (Fig. \ref {fig:model-fitting}B).
-The kinetic parameters of HCM FBA were varied to fit the concentration profiles of the HCM EM model.
-For this proof of concept model, HCM FBA was able to use fewer modes and kinetic parameters to mimic the simulation profiles of HCM EM.
+The performance of the HCM FBA approach was equivalent to HCM EM for a proof of concept metabolic network (Fig. \ref{fig:model-fitting}).
+The proof of concept network, consisting of six metabolites and seven reactions (Fig. \ref{fig:model-fitting}A), generated three FBA and six elementary modes.
+Using the elementary modes and synthetic parameters, we generated test data from which we estimated the HCM FBA model parameters.
+The best fit HCM FBA model replicated the synthetic data (Fig. \ref {fig:model-fitting}B).
+The HCM EM and HCM FBA kinetic parameters were not quantitatively identical, but had similar orders of magnitude;
+the FBA approach had three fewer modes, thus you would not expect identical parameter values.
+Taken together, the HCM FBA approach replicated synthetic data generated using HCM EM, despite having three fewer modes.
+Next, we tested the ability of HCM FBA to replicate experimental data.
+
+%In the case, we demonstrated that HCM FBA has comparable model performance to a published HCM EM model \cite{2008_kim_varner_ramkrishna_BiotechProg}.
+%and elementary modes reported
+%The kinetic parameters of HCM FBA were varied to fit the same experimental observations.
+%HCM FBA has 17 kinetic parameters while HCM EM has 21 kinetic parameters and both models resulted in similar fits to the experimental data .
+%HCM FBA had a better fit on lactate than HCM EM by 35.6\% on the
+%However, HCM EM had a lower residual standard error on the overall performance by 4.5\%.
+%However, ZZZ has shown that the EM algorithm has exponential runtime complexity [REFHERE].
+%Thus, HCM EM may not be possible for large networks, but HCM FBA will not suffer from this shortcoming.
+%In the case, we demonstrated that HCM FBA is applicable for larger metabolic networks where elementary mode calculations increase exponentially.
 
 The performance of HCM FBA was equivalent to HCM EM for a small model of anaerobic \textit{E. coli} metabolism (Fig. \ref{fig:ecoli}, left).
-In the case, we demonstrated that HCM FBA has comparable model performance to a published HCM EM model \cite{2008_kim_varner_ramkrishna_BiotechProg}.
-The metabolic network consisted of 12 reactions and 19 metabolites, where we calculated 9 elementary modes and 7 FBA modes.
-The simulated HCM EM concentration profiles of biomass, substrate and fermentation products (Fig.
-\ref {fig:ecoli}A (dashed line)) were generated using the kinetic parameters and elementary modes reported\cite{2008_kim_varner_ramkrishna_BiotechProg}.
-The kinetic parameters of HCM FBA were varied to fit the same experimental observations.
-HCM FBA has 17 kinetic parameters while HCM EM has 21 kinetic parameters and both models resulted in similar fits to the experimental data (Fig. \ref {fig:ecoli}A).
-HCM FBA had a better fit on lactate than HCM EM by 35.6\% on the residual standard error. 
-However, HCM EM had a lower residual standard error on the overall performance by 4.5\%.
+The anaerobic \textit{E. coli} network, consisting of 12 reactions and 19 metabolites, generated nine elementary and seven FBA modes.
+HCM EM cellmass, glucose and fermentation product trajectories were generated
+using these elementary modes and kinetic parameters reported by Kim et al. \cite{2008_kim_varner_ramkrishna_BiotechProg},
+where the experimental data was taken from ZZZ et al. [REFHERE] (Fig. \ref{fig:ecoli}A, dashed).
+We estimated the HCM FBA model parameters from the experimental measurements using simulated annealing.
+Overall, HCM FBA performed only 5\% worse than HCM EM (Fig. \ref {fig:ecoli}A, solid), despite having two fewer modes and four fewer parameters (17 versus 21 parameters).
+Taken together, these results suggested HCM FBA is an alternative to HCM EM even for cases when elementary modes can be easily calculated.
 
@@
 HCM FBA outperformed HCM EM for a large model of aerobic \textit{E. coli} metabolism (Fig. \ref{fig:ecoli}, right).
-In the case, we demonstrated that HCM FBA is applicable for larger metabolic networks where elementary mode calculations increase exponentially.
-A metabolic network\cite{2007_schuetz_etal_MolSysBio} of 98 reactions and 60 metabolites with the biomass reaction from Palsson's \textit{E. coli} core model\cite{2006_Palsson_model} was used.
-We calculated over 66,000 elementary modes for this network and determined it was futile to model.
-For the same network, we computed 29 FBA modes.
-HCM FBA was simulated to fit cell density, glucose and acetate measurements for an aerobic \textit{E. coli} batch culture\cite{1994_varma_palsson_ApplEnvMicro} (Fig.\ref{fig:ecoli}B (solid line)).
+An \emph{E. coli} metabolic network, consisting of 60 metabolites and 110 reactions, was constructed from literature \cite{2007_schuetz_etal_MolSysBio,2006_Palsson_model}.
+Elementary mode decomposition of this network was not feasible; we generated $>$ 150,000 elementary modes before prematurely terminating the calculation.
+Conversely, flux balance analysis generated only 29 modes for the same network.
+HCM FBA model parameters were estimated from cellmass, glucose and acetate measurements generated by Palsson and coworkers \cite{1994_varma_palsson_ApplEnvMicro} (Fig. \ref{fig:ecoli}B, solid).
+HCM FBA captured glucose consumption, cellmass formation and the switch between acetate production and consumption following glucose exhaustion.
+Taken together, the aerobic \textit{E. coli} results demonstrated that HCM FBA was able to describe cell growth, including the switch between glucose and acetate consumption,
+for a network that was infeasible to decompose using elementary modes.
+However, it was unclear whether all 29 FBA modes were required to capture the experimental data.
 
-Variance based sensitivity of aerobic \textit{E. coli} metabolism identified modes that were essential to biomass yield (Fig. \ref{fig:sensitivity}).
+Variance based sensitivity of aerobic \textit{E. coli} metabolism identified which FBA modes that were essential to biomass yield (Fig. \ref{fig:sensitivity}).
 The analysis calculated the total order sensitivity indicies of all the kinetic parameters and enzyme initial conditions.
@@
 The remaining 24 FBA modes were identified to be minor modes and insignificant.
-The removal of the major mode, representing aerobic growth, (Fig, \ref{fig:ecoli}B (dotted line)) was detrimental to model performance, whereas the removal of the minor modes (dashed line) had little effect.
+The removal of the major mode, representing aerobic growth, (Fig, \ref{fig:ecoli}B, dotted)
+was detrimental to model performance, whereas the removal of the minor modes (dashed line) had little effect.
 
 \begin{figure*}[!t]\centering
-\includegraphics[width=0.85\textwidth]{./figs/Fig-3-Sensitivity-Results.pdf}
+\includegraphics[width=0.80\textwidth]{./figs/Fig-3-Sensitivity-Results.pdf}
 \caption{Total order variance based sensitivity analysis on biomass yield from glucose and acetate.

### 2016-03-13 Paper_v1.tex
More changes afternoon 3/12

@@
 
-The performance of HCM FBA was equivalent to HCM EM for a small model of anaerobic \textit{E. coli} metabolism (Fig. \ref{fig:ecoli}, left).
+The performance of HCM FBA was equivalent to HCM EM for anaerobic \textit{E. coli} metabolism (Fig. \ref{fig:ecoli}, left).
 The anaerobic \textit{E. coli} network, consisting of 12 reactions and 19 metabolites, generated nine elementary and seven FBA modes.
 HCM EM cellmass, glucose and fermentation product trajectories were generated
-using these elementary modes and kinetic parameters reported by Kim et al. \cite{2008_kim_varner_ramkrishna_BiotechProg},
-where the experimental data was taken from ZZZ et al. [REFHERE] (Fig. \ref{fig:ecoli}A, dashed).
-We estimated the HCM FBA model parameters from the experimental measurements using simulated annealing.
+using these elementary modes and kinetic parameters reported by Kim et al. \cite{2008_kim_varner_ramkrishna_BiotechProg} (Fig. \ref{fig:ecoli}A, dashed).
+We estimated the HCM FBA model parameters from the experimental data set taken from Kim et al.  \cite{2008_kim_varner_ramkrishna_BiotechProg} using simulated annealing.
 Overall, HCM FBA performed only 5\% worse than HCM EM (Fig. \ref {fig:ecoli}A, solid), despite having two fewer modes and four fewer parameters (17 versus 21 parameters).

### 2016-03-13 Paper_v1.tex
Results section done

@@
 The performance of HCM FBA was equivalent to HCM EM for anaerobic \textit{E. coli} metabolism (Fig. \ref{fig:ecoli}, left).
-The anaerobic \textit{E. coli} network, consisting of 12 reactions and 19 metabolites, generated nine elementary and seven FBA modes.
-HCM EM cellmass, glucose and fermentation product trajectories were generated
-using these elementary modes and kinetic parameters reported by Kim et al. \cite{2008_kim_varner_ramkrishna_BiotechProg} (Fig. \ref{fig:ecoli}A, dashed).
-We estimated the HCM FBA model parameters from the experimental data set taken from Kim et al.  \cite{2008_kim_varner_ramkrishna_BiotechProg} using simulated annealing.
-Overall, HCM FBA performed only 5\% worse than HCM EM (Fig. \ref {fig:ecoli}A, solid), despite having two fewer modes and four fewer parameters (17 versus 21 parameters).
-Taken together, these results suggested HCM FBA is an alternative to HCM EM even for cases when elementary modes can be easily calculated.
+We constructed an anaerobic \textit{E. coli} network, consisting of 12 reactions and 19 metabolites, which generated nine elementary and seven FBA modes.
+HCM EM cellmass, glucose and fermentation product trajectories, generated
+using kinetic parameters reported by Kim et al. \cite{2008_kim_varner_ramkrishna_BiotechProg},
+were compared with experimental measurements (Fig. \ref{fig:ecoli}A, dashed).
+We estimated HCM FBA model parameters from the Kim et al. experimental data set using simulated annealing.
+Overall, HCM FBA performed within 5\% of HCM EM (on a residual standard error basis) for the anaerobic \textit{E. coli} data (Fig. \ref {fig:ecoli}A, solid),
+despite having two fewer modes and four fewer parameters (17 versus 21 parameters).
+Taken together, these results suggested HCM FBA  was capable of describing experimental data, and is a viable reduced-order alternative to HCM EM.
 
@@
 
-HCM FBA outperformed HCM EM for a large model of aerobic \textit{E. coli} metabolism (Fig. \ref{fig:ecoli}, right).
-An \emph{E. coli} metabolic network, consisting of 60 metabolites and 110 reactions, was constructed from literature \cite{2007_schuetz_etal_MolSysBio,2006_Palsson_model}.
-Elementary mode decomposition of this network was not feasible; we generated $>$ 150,000 elementary modes before prematurely terminating the calculation.
+HCM FBA captured the glucose to acetate metabolic shift for a large model of aerobic \textit{E. coli} metabolism (Fig. \ref{fig:ecoli}, right).
+A core \emph{E. coli} metabolic network, consisting of 60 metabolites and 105 reactions, was constructed from literature \cite{2007_schuetz_etal_MolSysBio,2006_Palsson_model}.
+Elementary mode decomposition (and thus HCM EM) was not feasible for this network; 153,000 elementary modes were generated before the calculation prematurely terminated.
 Conversely, flux balance analysis generated only 29 modes for the same network.
-HCM FBA model parameters were estimated from cellmass, glucose and acetate measurements generated by Palsson and coworkers \cite{1994_varma_palsson_ApplEnvMicro} (Fig. \ref{fig:ecoli}B, solid).
-HCM FBA captured glucose consumption, cellmass formation and the switch between acetate production and consumption following glucose exhaustion.
-Taken together, the aerobic \textit{E. coli} results demonstrated that HCM FBA was able to describe cell growth, including the switch between glucose and acetate consumption,
-for a network that was infeasible to decompose using elementary modes.
-However, it was unclear whether all 29 FBA modes were required to capture the experimental data.
+HCM FBA model parameters were estimated from cellmass, glucose and acetate measurements \cite{1994_varma_palsson_ApplEnvMicro} using simulated annealing (Fig. \ref{fig:ecoli}B, solid).
+HCM FBA captured glucose consumption, cellmass formation and the dynamic switch between acetate production and consumption following glucose exhaustion.
+Taken together, the aerobic \textit{E. coli} results demonstrated that HCM FBA described metabolic dynamics for large
+networks that were infeasible for HCM EM. However, it was unclear whether all 29 FBA modes were required to capture the experimental data.
+Next, we considered a systematic approach to identify the critical subset of FBA modes required for model performance.
 
-Variance based sensitivity of aerobic \textit{E. coli} metabolism identified which FBA modes that were essential to biomass yield (Fig. \ref{fig:sensitivity}).
-The analysis calculated the total order sensitivity indicies of all the kinetic parameters and enzyme initial conditions.
-We used the sensitivity results as a simple systematic method to eliminate FBA modes to give the smallest number needed while maintaining good model performance.
-Five FBA modes were determined to be significant by an arbitrary threshold of 0.00218.
-The remaining 24 FBA modes were identified to be minor modes and insignificant.
-The removal of the major mode, representing aerobic growth, (Fig, \ref{fig:ecoli}B, dotted)
-was detrimental to model performance, whereas the removal of the minor modes (dashed line) had little effect.
+Global variance based sensitivity analysis of aerobic \textit{E. coli} metabolism identified the FBA modes essential to biomass yield (Fig. \ref{fig:sensitivity}).
+Total order sensitivity coefficients were calculated for all kinetic parameters and enzyme initial conditions.
+Five of the 29 FBA modes were determined to be significant by a threshold test.
+Removal of the most significant mode (encoding aerobic growth on glucose) destroyed model performance (Fig. \ref{fig:ecoli}B, dotted).
+Conversely, removal of the remaining 24 modes had negligible effect on model performance (Fig. \ref{fig:ecoli}B, dashed).
+Taken together, these results demonstrate that sensitivity analysis can identify critical FBA modes, thereby allowing further reduction to the minimal model structure
+required to explain the data.
 

### 2016-03-15 Paper_v1.tex
More results updates

@@
 \section{Results}
-The performance of the HCM FBA approach was equivalent to HCM EM for a proof of concept metabolic network (Fig. \ref{fig:model-fitting}).
+HCM FBA was equivalent to HCM EM for a proof of concept metabolic network (Fig. \ref{fig:model-fitting}).
 The proof of concept network, consisting of six metabolites and seven reactions (Fig. \ref{fig:model-fitting}A), generated three FBA and six elementary modes.
@@
 We constructed an anaerobic \textit{E. coli} network, consisting of 12 reactions and 19 metabolites, which generated nine elementary and seven FBA modes.
-HCM EM cellmass, glucose and fermentation product trajectories, generated
-using kinetic parameters reported by Kim et al. \cite{2008_kim_varner_ramkrishna_BiotechProg},
-were compared with experimental measurements (Fig. \ref{fig:ecoli}A, dashed).
-We estimated HCM FBA model parameters from the Kim et al. experimental data set using simulated annealing.
+HCM EM reproduced cellmass, glucose and byproduct trajectories
+using the kinetic parameters reported by Kim et al. \cite{2008_kim_varner_ramkrishna_BiotechProg} (Fig. \ref{fig:ecoli}A, points versus dashed).
+On the other hand, HCM FBA model parameters were estimated in this study from the Kim et al. data set using simulated annealing.
 Overall, HCM FBA performed within 5\% of HCM EM (on a residual standard error basis) for the anaerobic \textit{E. coli} data (Fig. \ref {fig:ecoli}A, solid),
 despite having two fewer modes and four fewer parameters (17 versus 21 parameters).
-Taken together, these results suggested HCM FBA  was capable of describing experimental data, and is a viable reduced-order alternative to HCM EM.
+Taken together, while both HCM EM and HCM FBA described experimental data, HCM FBA had fewer parameters and modes.
 
@@
 
-HCM FBA captured the glucose to acetate metabolic shift for a large model of aerobic \textit{E. coli} metabolism (Fig. \ref{fig:ecoli}, right).
+HCM FBA captured the shift from glucose to acetate consumption for a model of aerobic \textit{E. coli} metabolism that was infeasible for HCM EM (Fig. \ref{fig:ecoli}, right).
 A core \emph{E. coli} metabolic network, consisting of 60 metabolites and 105 reactions, was constructed from literature \cite{2007_schuetz_etal_MolSysBio,2006_Palsson_model}.
-Elementary mode decomposition (and thus HCM EM) was not feasible for this network; 153,000 elementary modes were generated before the calculation prematurely terminated.
+Elementary mode decomposition of this network (and thus HCM EM) was not feasible; 153,000 elementary modes were generated before the calculation became infeasible.
 Conversely, flux balance analysis generated only 29 modes for the same network.
 HCM FBA model parameters were estimated from cellmass, glucose and acetate measurements \cite{1994_varma_palsson_ApplEnvMicro} using simulated annealing (Fig. \ref{fig:ecoli}B, solid).
-HCM FBA captured glucose consumption, cellmass formation and the dynamic switch between acetate production and consumption following glucose exhaustion.
-Taken together, the aerobic \textit{E. coli} results demonstrated that HCM FBA described metabolic dynamics for large
-networks that were infeasible for HCM EM. However, it was unclear whether all 29 FBA modes were required to capture the experimental data.
+HCM FBA captured glucose consumption, cellmass formation and the switch to acetate consumption following glucose exhaustion.
+Taken together, HCM FBA described the dynamics of a network that was infeasible for HCM EM, thereby demonstrating the power of the approach for large networks.
 Next, we considered a systematic approach to identify the critical subset of FBA modes required for model performance.
 
-Global variance based sensitivity analysis of aerobic \textit{E. coli} metabolism identified the FBA modes essential to biomass yield (Fig. \ref{fig:sensitivity}).
-Total order sensitivity coefficients were calculated for all kinetic parameters and enzyme initial conditions.
-Five of the 29 FBA modes were determined to be significant by a threshold test.
-Removal of the most significant mode (encoding aerobic growth on glucose) destroyed model performance (Fig. \ref{fig:ecoli}B, dotted).
-Conversely, removal of the remaining 24 modes had negligible effect on model performance (Fig. \ref{fig:ecoli}B, dashed).
-Taken together, these results demonstrate that sensitivity analysis can identify critical FBA modes, thereby allowing further reduction to the minimal model structure
-required to explain the data.
+Global sensitivity analysis identified the FBA modes essential to model performance (Fig. \ref{fig:sensitivity}).
+Total order sensitivity coefficients were calculated for all kinetic parameters and enzyme initial conditions in the aerobic \textit{E. coli} model.
+Five of the 29 FBA modes were significant;
+removal of the most significant of these modes (encoding aerobic growth on glucose) destroyed model performance (Fig. \ref{fig:ecoli}B, dotted).
+Conversely, removal of the remaining 24 modes had a negligible effect upon model performance (Fig. \ref{fig:ecoli}B, dashed).
+Taken together, sensitivity analysis identified the minimal model structure required to explain the experimental data.
 
@@
 \caption{Total order variance based sensitivity analysis on biomass yield from glucose and acetate.
-Sensitivity indicies computed for rate constants, saturation constants and enzyme parameters.
-The most critical parameters are associated with mode 1 representing aerobic growth.
-(Error bars represent 95\% confidence intervals).
+Sensitivity coefficients were computed for kinetic parameters and enzyme initial conditions.
+The most critical parameters are associated with mode 1 representing aerobic growth. (Error bars represent 95\% confidence intervals).
 }
@@
 \section{Discussion}
-In this study, we developed a hybrid cybernetic model in combination with flux balance analysis modes instead of elementary modes.
-First, we showed HCM FBA can have comparable model performance to HCM EM for a proof of concept metabolic network.
-This judgement is based on the quality of fit of the model simulation to the observations.
-Next, HCM FBA model performance was compared to an established HCM EM model\cite{2008_kim_varner_ramkrishna_BiotechProg} for an anaerobic culture of \textit{E. coli}.
-HCM FBA had comparable fits to HCM EM on the fermentation data and a slightly better fit on lactate measurements.
-Finally, HCM was applied to a larger metabolic network where 29 FBA modes were computed, in contrast to over 66,000 elementary modes.
-HCM FBA was shown to have robust model performance only requiring 5 of the 29 FBA modes (selected by the sensitivity analysis) to fit experimental observations.
+In this study, we developed HCM FBA, an effective modeling technique to simulate metabolic dynamics.
+HCM FBA is a modification of the hybrid cybernetic approach of Ramkrishna and coworkers \cite{2008_kim_varner_ramkrishna_BiotechProg}.
+HCM FBA uses flux balance analysis solutions (instead of elementary modes) in conjunction with cybernetic control variables to dynamically simulate metabolism.
+We studied the performance of HCM FBA on a proof of concept metabolic network, along with two \emph{E. coli} networks.
+First, we showed the performance of HCM FBA and HCM EM were comparable for both a proof of concept metabolic network, and a small model of anaerobic \textit{E. coli} metabolism.
+For the anaerobic case, both approaches described experimental data with HCM FBA within 5\% of HCM EM for most of the experimental measurements, and slightly better for lactate secretion.
+Next, HCM FBA was applied to an aerobic \textit{E. coli} metabolic network that was not feasible for HCM EM.
+Elementary mode decomposition of the core aerobic network generated over 150,000 elementary modes.
+Conversely, the HCM FBA approach described cellmass growth, and the shift from glucose to acetate consumption, with only 29 FBA modes.
+Global sensitivity analysis further showed that only five of the 29 FBA modes were critical to model performance.
+Removal of these modes crippled the model, but removal of the remaining 24 modes had a negligible impact.
+Taken together, HCM FBA is a viable alternative to HCM EM, especially for large networks where the generation of elementary modes is unfeasible.
 
-The hybrid cybernetic model is formulated to maximize substrate uptake based on cybernetic arguments.
-Towards this objective, HCM uses a combination of modes to describe both the external and internal fluxes.
+Hybrid cybernetic models are formulated to maximize substrate uptake based on cybernetic arguments.
 The internal fluxes still follow the pseudo steady state approximation applied in FBA, but HCM uses a combination of pathways.
@@
 
-HCM FBA provides a practical and feasible approach to model large networks where elementary modes fail.
-The disadvantage of HCM FBA is it requires \textit{a priori} knowledge of the inputs into the metabolic network.
-FBA modes do not have the thorough information as elementary modes, but provide sufficient pathways to model experimental observations.
+%HCM FBA provides a practical and feasible approach to model large networks where elementary modes fail.
+%The disadvantage of HCM FBA is it requires \textit{a priori} knowledge of the inputs into the metabolic network.
+%FBA modes do not have the thorough information as elementary modes, but provide sufficient pathways to model experimental observations.
 %HCM FBA may also be applied to mammalian networks, as long as FBA accounts for compartmentalization within mammalian systems.
@@
 The reaction rate for formate was also modified following Kim et al.\cite{2008_kim_varner_ramkrishna_BiotechProg}.
-All numerical simulations were conducted in Julia version 0.4.2, a high-level, high performance dynamic programming language \cite{Julia}.
-Species balances were solved using a Sundials \cite{Sundials} wrapper for julia.
+All numerical simulations were conducted in the high-performance computing language Julia (v0.4.2) \cite{Julia}.
+Species balances were solved using the SUNDIALS package for Julia \cite{Sundials}.
 
-\noindent\subsubsection*{Elementary Mode and Flux Balance Analysis}
+\noindent\subsubsection*{Elementary mode and flux balance analysis}
 Elementary modes were calculated using METATOOL 5.1 \cite{2006_vonKamp_Metatool}.
@@
 
-\subsubsection*{Global Sensitivity Analysis}
+\subsubsection*{Global sensitivity analysis}
 Variance based sensitivity analysis was used to estimate which FBA modes were critical to model performance.

### 2016-03-15 Paper_v1.tex
More results

@@
 despite having two fewer modes and four fewer parameters (17 versus 21 parameters).
-Taken together, while both HCM EM and HCM FBA described experimental data, HCM FBA had fewer parameters and modes.
+Thus, while both HCM EM and HCM FBA described the experimental data, HCM FBA did so with fewer modes and parameters.
 
@@
 HCM FBA captured glucose consumption, cellmass formation and the switch to acetate consumption following glucose exhaustion.
-Taken together, HCM FBA described the dynamics of a network that was infeasible for HCM EM, thereby demonstrating the power of the approach for large networks.
-Next, we considered a systematic approach to identify the critical subset of FBA modes required for model performance.
+Thus, HCM FBA described the dynamics of a network that was infeasible for HCM EM, thereby demonstrating the power of the approach for large networks.
+Next, we demonstrated a systematic strategy to identify the critical subset of FBA modes required for model performance.
 
@@
 removal of the most significant of these modes (encoding aerobic growth on glucose) destroyed model performance (Fig. \ref{fig:ecoli}B, dotted).
-Conversely, removal of the remaining 24 modes had a negligible effect upon model performance (Fig. \ref{fig:ecoli}B, dashed).
-Taken together, sensitivity analysis identified the minimal model structure required to explain the experimental data.
+Conversely, removing the remaining 24 modes had a negligible effect upon model performance (Fig. \ref{fig:ecoli}B, dashed).
+Thus, sensitivity analysis identified the minimal model structure required to explain the experimental data.
 

### 2016-03-16 Paper_v1.tex
Working on introduction

@@
 Mathematical models of biochemical networks are useful tools to understand and ultimately predict how cells utilize nutrients to produce valuable products.
-Hybrid cybernetic models in combination with elementary modes (HCM EM) provides a route to model cellular metabolism.
+Hybrid cybernetic models in combination with elementary modes (HCM-EM) provides a route to model cellular metabolism.
 However, this framework is limited to smaller networks due to the computational cost of calculating elementary modes and a high number of solutions for large networks.
-In this study, we developed a hybrid cybernetic model in combination with flux balance analysis (HCM FBA) modes instead of elementary modes.
+In this study, we developed the hybrid cybernetic modeling with flux balance analysis or HCM-FBA technique, which uses flux balance solutions instead of elementary modes,
+to dynamically model metabolism.
 Flux balance analysis provides a sufficient number of modes (metabolic options) for the network to estimate experimental observations.
-We show HCM FBA has comparable model performance to HCM EM for a hypothetical proof of concept metabolic network and for a reduced anaerobic \textit{E. coli} network.
-HCM FBA was applied to a larger metabolic network where 29 FBA modes were computed, in contrast to over 66,000 elementary modes.
+We show HCM-FBA has comparable model performance to HCM-EM for a hypothetical proof of concept metabolic network and for a reduced anaerobic \textit{E. coli} network.
+HCM-FBA was applied to a larger metabolic network where 29 FBA modes were computed, in contrast to over 66,000 elementary modes.
 A sensitivity analysis was performed to reduce the number of FBA modes from 29 to 5.
@@
 \section{Introduction}
-Metabolic networks for simple organisms such as \textit{E. coli} contain thousands of reactions that are highly interconnected.
-These organisms have the potential to be engineered to produce valuable products.
-Mathematical models of biochemical networks are useful tools to understand and ultimately predict how cells utilize nutrients to produce valuable products.
-Traditionally, Monod kinetics have been used as unstructured models to describe cell growth, substrate consumption, and product formation, but fail to account for intracellular metabolism\cite{shuler_book}.
-Structured modeling approaches have made great progress to describe metabolic pathways including elementary mode analysis (EM)\cite{2006_vonKamp_Metatool}, metabolic flux analysis (MFA), flux balance analysis (FBA)\cite{2010_orth_NatBiotech}, dynamic FBA (DFBA) \cite{1994_varma_palsson_ApplEnvMicro,2002_Mahadevan_BiophysJ}, and hybrid cybernetic modeling (HCM) \cite{2008_kim_varner_ramkrishna_BiotechProg}.
+%Metabolic networks for simple organisms such as \textit{E. coli} contain thousands of reactions that are highly interconnected.
+%These organisms have the potential to be engineered to produce valuable products.
+%Mathematical models of biochemical networks are useful tools to understand and ultimately predict how cells utilize nutrients to produce valuable products.
+%Traditionally, Monod kinetics have been used as unstructured models to describe cell growth, substrate consumption, and product formation, but fail to account for intracellular metabolism\cite{shuler_book}.
+%Structured modeling approaches have made great progress to describe metabolic pathways including elementary mode analysis (EM)\cite{2006_vonKamp_Metatool}, metabolic flux analysis (MFA), flux balance analysis (FBA)\cite{2010_orth_NatBiotech}, dynamic FBA (DFBA) \cite{1994_varma_palsson_ApplEnvMicro,2002_Mahadevan_BiophysJ}, and hybrid cybernetic modeling (HCM) \cite{2008_kim_varner_ramkrishna_BiotechProg}.
+%thermodynamically feasible
 
-Convex and constraints based analysis methods, such as EM, FBA and MFA,
-describe intracellular metabolism using the stoichiometry of the biochemical networks and the pseudo-steady state assumption \cite{2010_orth_NatBiotech}.
-EM analysis is a powerful tool to decompse a metabolic network into independent minimal pathways that are stoichiometrically and thermodynamically feasible.
+Metabolic networks, even for simple organisms such as \textit{E. coli}, contain thousands of intracellular metabolites and reactions that are highly interconnected.
+Constraints based models and convex analysis techniques have become the standard approaches to interrogate these complex networks.
+Constraints based methods such as flux balance analysis (FBA) \cite{2010_orth_NatBiotech} and convex network decomposition approaches such elementary modes (EMs) [REF] or extreme pathways (EPs) [REF] describe intracellular metabolism using the biochemical stoichiometry, and other constraints such as thermodynamically feasibility [REF]
+under pseudo-steady state conditions.
 However, elementary mode calculations are computational expensive and infeasible for genome scale networks\cite{2004_lee_varner_ko_ieee}.
-FBA is frequently used to study genome-scale networks\cite{2010_orth_NatBiotech} and often results in good approximations for yields, however it fails to estimate intracellular metabolism due to its underdetermined solution space.
+FBA is frequently used to study genome-scale networks and often results in good approximations for yields, however it fails to estimate intracellular metabolism due to its underdetermined solution space.
 Compared with FBA, MFA has good estimations of intracellular metabolism, but is dependent on experimental data.
@@
 
-In this study, we developed a hybrid cybernetic model in combination with flux balance analysis modes instead of elementary modes.
-We show HCM FBA has comparable model performance to HCM EM for a hypothetical metabolic network and a reduced anaerobic \textit{E. coli} network.
-HCM EM and HCM FBA were then applied to a large metabolic network.
-The number of elementary modes made it impractical to model with HCM EM, but HCM FBA was computationally feasible and performed well with respect to experimental observations.
-Taken together, HCM FBA has comparable model performace to HCM EM and may be potentially applied to genome-scale networks which was a shortcoming of HCM EM.
+In this study, we developed the hybrid cybernetic modeling with flux balance analysis or HCM-FBA technique.
+HCM-FBA is a modification of the hybrid cybernetic approach of Ramkrishna and coworkers which uses flux balance analysis solutions
+(instead of elementary modes) in conjunction with cybernetic control variables to dynamically simulate metabolism.
+We compared the performance of HCM-FBA to HCM-EM for a proof of concept metabolic network, along with two \emph{E. coli} networks.
+HCM-FBA performed similar to HCM-EM for the hypothetical network and a reduced anaerobic \textit{E. coli} network, despite having fewer parameters in each case.
+Next, HCM-FBA was applied to an aerobic \textit{E. coli} metabolic network that was not feasible for HCM-EM.
+The HCM-FBA approach described cellmass growth, and the shift from glucose to acetate consumption, with only a few modes.
+Global sensitivity analysis allowed us to further reduce the aerobic \textit{E. coli} HCM-FBA model to the minimal model required to describe the data.
+Thus, we demonstrated that HCM-FBA is a promising approach for the development of reduced order dynamic metabolic models,
+and a viable alternative to the HCM-EM approach especially for large networks where the generation of elementary modes is infeasible.
+
+%HCM-FBA also captured the dynamics of a larger anaerobic \textit{E. coli} network, that was infeasible for HCM-EM.
+%Taken together, HCM-FBA has comparable model performace to HCM-EM and may be potentially applied to genome-scale networks which was a shortcoming of HCM-EM.
+%HCM-EM and HCM-FBA were then applied to a large metabolic network.
+%The number of elementary modes made it impractical to model with HCM-EM, but HCM-FBA was computationally feasible and performed well with respect to experimental observations.
 
@@
 \includegraphics[width=0.40\textwidth]{./figs/Fig-1-GeneralModel-Results.pdf}
-\caption{A: Proof of concept metabolic network with six metabolites and seven reactions.
+\caption{HCM-FBA proof of concept metabolic study.
+A: Prototypical network with six metabolites and seven reactions.
 Intracellular cellmass precursors $A,B$ and $C$ are balanced (no accumulation) while the extracellular metabolites ($A_{e},B_{e}$ and $C_{e}$) are not balanced (can accumulate).
-The blue-oval denotes the cell boundary, where $q_{j}$ denotes the jth flux across the boundaries ([mmol/gdw-hr]) and $v_{k}$ denotes the kth intracellular flux.
-B: Simulation of the extracellular metabolites using the FBA hybrid cybernetic approach (solid line) versus elementary modes (points) for the proof of concept metabolic network.
-FBA modes have comparable model performance compared to elementary modes.
+The oval denotes the cell boundary, $q_{j}$ is the jth flux across the boundary and $v_{k}$ denotes the kth intracellular flux.
+B: Simulation of extracellular metabolite trajectories using HCM-FBA (solid line) versus HCM-EM (points) for the proof of concept metabolic network.
 }\label{fig:model-fitting}
@@
 \section{Results}
-HCM FBA was equivalent to HCM EM for a proof of concept metabolic network (Fig. \ref{fig:model-fitting}).
+HCM-FBA was equivalent to HCM-EM for a proof of concept metabolic network (Fig. \ref{fig:model-fitting}).
 The proof of concept network, consisting of six metabolites and seven reactions (Fig. \ref{fig:model-fitting}A), generated three FBA and six elementary modes.
-Using the elementary modes and synthetic parameters, we generated test data from which we estimated the HCM FBA model parameters.
-The best fit HCM FBA model replicated the synthetic data (Fig. \ref {fig:model-fitting}B).
-The HCM EM and HCM FBA kinetic parameters were not quantitatively identical, but had similar orders of magnitude;
+Using the elementary modes and synthetic parameters, we generated test data from which we estimated the HCM-FBA model parameters.
+The best fit HCM-FBA model replicated the synthetic data (Fig. \ref {fig:model-fitting}B).
+The HCM-EM and HCM-FBA kinetic parameters were not quantitatively identical, but had similar orders of magnitude;
 the FBA approach had three fewer modes, thus you would not expect identical parameter values.
-Taken together, the HCM FBA approach replicated synthetic data generated using HCM EM, despite having three fewer modes.
-Next, we tested the ability of HCM FBA to replicate experimental data.
+Taken together, the HCM-FBA approach replicated synthetic data generated using HCM-EM, despite having three fewer modes.
+Next, we tested the ability of HCM-FBA to replicate experimental data.
 
-%In the case, we demonstrated that HCM FBA has comparable model performance to a published HCM EM model \cite{2008_kim_varner_ramkrishna_BiotechProg}.
+%In the case, we demonstrated that HCM-FBA has comparable model performance to a published HCM-EM model \cite{2008_kim_varner_ramkrishna_BiotechProg}.
 %and elementary modes reported
-%The kinetic parameters of HCM FBA were varied to fit the same experimental observations.
-%HCM FBA has 17 kinetic parameters while HCM EM has 21 kinetic parameters and both models resulted in similar fits to the experimental data .
-%HCM FBA had a better fit on lactate than HCM EM by 35.6\% on the
-%However, HCM EM had a lower residual standard error on the overall performance by 4.5\%.
+%The kinetic parameters of HCM-FBA were varied to fit the same experimental observations.
+%HCM-FBA has 17 kinetic parameters while HCM-EM has 21 kinetic parameters and both models resulted in similar fits to the experimental data .
+%HCM-FBA had a better fit on lactate than HCM-EM by 35.6\% on the
+%However, HCM-EM had a lower residual standard error on the overall performance by 4.5\%.
 %However, ZZZ has shown that the EM algorithm has exponential runtime complexity [REFHERE].
-%Thus, HCM EM may not be possible for large networks, but HCM FBA will not suffer from this shortcoming.
-%In the case, we demonstrated that HCM FBA is applicable for larger metabolic networks where elementary mode calculations increase exponentially.
+%Thus, HCM-EM may not be possible for large networks, but HCM-FBA will not suffer from this shortcoming.
+%In the case, we demonstrated that HCM-FBA is applicable for larger metabolic networks where elementary mode calculations increase exponentially.
 
-The performance of HCM FBA was equivalent to HCM EM for anaerobic \textit{E. coli} metabolism (Fig. \ref{fig:ecoli}, left).
+The performance of HCM-FBA was equivalent to HCM-EM for anaerobic \textit{E. coli} metabolism (Fig. \ref{fig:ecoli}, left).
 We constructed an anaerobic \textit{E. coli} network, consisting of 12 reactions and 19 metabolites, which generated nine elementary and seven FBA modes.
-HCM EM reproduced cellmass, glucose and byproduct trajectories
+HCM-EM reproduced cellmass, glucose and byproduct trajectories
 using the kinetic parameters reported by Kim et al. \cite{2008_kim_varner_ramkrishna_BiotechProg} (Fig. \ref{fig:ecoli}A, points versus dashed).
-On the other hand, HCM FBA model parameters were estimated in this study from the Kim et al. data set using simulated annealing.
-Overall, HCM FBA performed within 5\% of HCM EM (on a residual standard error basis) for the anaerobic \textit{E. coli} data (Fig. \ref {fig:ecoli}A, solid),
+On the other hand, HCM-FBA model parameters were estimated in this study from the Kim et al. data set using simulated annealing.
+Overall, HCM-FBA performed within 5\% of HCM-EM (on a residual standard error basis) for the anaerobic \textit{E. coli} data (Fig. \ref {fig:ecoli}A, solid),
 despite having two fewer modes and four fewer parameters (17 versus 21 parameters).
-Thus, while both HCM EM and HCM FBA described the experimental data, HCM FBA did so with fewer modes and parameters.
+Thus, while both HCM-EM and HCM-FBA described the experimental data, HCM-FBA did so with fewer modes and parameters.
 
@@
 \includegraphics[width=0.48\textwidth]{./figs/Fig-2-Ecoli-SimulationResults.pdf}
-\caption{A: Batch anaerobic \textit{E.
-coli} fermentation data versus simulation of the hybrid cybernetic model with FBA modes (solid line) and elementary modes (dashed line).
-(Experimental data from Kim et.
-al (2008)\cite{2008_kim_varner_ramkrishna_BiotechProg} with error bars representing 90\% confidence intervals.) B: Batch aerobic \textit{E.
-coli} fermentation data of glucose, acetate and cell density measurements versus simulation of the hybrid cybernetic model with FBA modes (solid line).
-Model performance is shown when minor modes (dashed) and a major mode (dotted) is removed from HCM FBA.
-(Experimental data from Varma \& Palsson (1994)\cite{1994_varma_palsson_ApplEnvMicro} with assumed 10\% error bars).}
+\caption{HCM-FBA versus HCM-EM performance for small and large metabolic networks.
+A: Batch anaerobic \textit{E. coli} fermentation data versus HCM-FBA (solid) and HCM-EM (dashed).
+The experimental data was reproduced from from Kim et. al \cite{2008_kim_varner_ramkrishna_BiotechProg}. Error bars represent the 90\% confidence interval.
+B: Batch aerobic \textit{E. coli} fermentation data versus HCM-FBA (solid).
+Model performance is also shown when minor modes (dashed) and major modes (dotted) were removed from the HCM-FBA model.
+The experimental data was reproduced from Varma \& Palsson \cite{1994_varma_palsson_ApplEnvMicro}. Error bars denote a 10\% coefficient of variation.}
 \label{fig:ecoli}
@@
 
-HCM FBA captured the shift from glucose to acetate consumption for a model of aerobic \textit{E. coli} metabolism that was infeasible for HCM EM (Fig. \ref{fig:ecoli}, right).
+HCM-FBA captured the shift from glucose to acetate consumption for a model of aerobic \textit{E. coli} metabolism that was infeasible for HCM-EM (Fig. \ref{fig:ecoli}, right).
 A core \emph{E. coli} metabolic network, consisting of 60 metabolites and 105 reactions, was constructed from literature \cite{2007_schuetz_etal_MolSysBio,2006_Palsson_model}.
-Elementary mode decomposition of this network (and thus HCM EM) was not feasible; 153,000 elementary modes were generated before the calculation became infeasible.
+Elementary mode decomposition of this network (and thus HCM-EM) was not feasible; 153,000 elementary modes were generated before the calculation became infeasible.
 Conversely, flux balance analysis generated only 29 modes for the same network.
-HCM FBA model parameters were estimated from cellmass, glucose and acetate measurements \cite{1994_varma_palsson_ApplEnvMicro} using simulated annealing (Fig. \ref{fig:ecoli}B, solid).
-HCM FBA captured glucose consumption, cellmass formation and the switch to acetate consumption following glucose exhaustion.
-Thus, HCM FBA described the dynamics of a network that was infeasible for HCM EM, thereby demonstrating the power of the approach for large networks.
+HCM-FBA model parameters were estimated from cellmass, glucose and acetate measurements \cite{1994_varma_palsson_ApplEnvMicro} using simulated annealing (Fig. \ref{fig:ecoli}B, solid).
+HCM-FBA captured glucose consumption, cellmass formation and the switch to acetate consumption following glucose exhaustion.
+Thus, HCM-FBA described the dynamics of a network that was infeasible for HCM-EM, thereby demonstrating the power of the approach for large networks.
 Next, we demonstrated a systematic strategy to identify the critical subset of FBA modes required for model performance.
@@
 \includegraphics[width=0.80\textwidth]{./figs/Fig-3-Sensitivity-Results.pdf}
-\caption{Total order variance based sensitivity analysis on biomass yield from glucose and acetate.
-Sensitivity coefficients were computed for kinetic parameters and enzyme initial conditions.
-The most critical parameters are associated with mode 1 representing aerobic growth. (Error bars represent 95\% confidence intervals).
+\caption{Global sensitivity analysis of the aerobic \textit{E. coli} model.
+Total order variance based sensitivity coefficients were calculated for the biomass yield on glucose and acetate.
+Sensitivity coefficients were computed for kinetic parameters and enzyme initial conditions (N = 183,000).
+Error bars represent the 95\% confidence intervals of the sensitivity coefficients.
 }
@@
 \section{Discussion}
-In this study, we developed HCM FBA, an effective modeling technique to simulate metabolic dynamics.
-HCM FBA is a modification of the hybrid cybernetic approach of Ramkrishna and coworkers \cite{2008_kim_varner_ramkrishna_BiotechProg}.
-HCM FBA uses flux balance analysis solutions (instead of elementary modes) in conjunction with cybernetic control variables to dynamically simulate metabolism.
-We studied the performance of HCM FBA on a proof of concept metabolic network, along with two \emph{E. coli} networks.
-First, we showed the performance of HCM FBA and HCM EM were comparable for both a proof of concept metabolic network, and a small model of anaerobic \textit{E. coli} metabolism.
-For the anaerobic case, both approaches described experimental data with HCM FBA within 5\% of HCM EM for most of the experimental measurements, and slightly better for lactate secretion.
-Next, HCM FBA was applied to an aerobic \textit{E. coli} metabolic network that was not feasible for HCM EM.
+In this study, we developed HCM-FBA, an effective modeling technique to simulate metabolic dynamics.
+HCM-FBA is a modification of the hybrid cybernetic approach of Ramkrishna and coworkers \cite{2008_kim_varner_ramkrishna_BiotechProg}.
+HCM-FBA uses flux balance analysis solutions (instead of elementary modes) in conjunction with cybernetic control variables to dynamically simulate metabolism.
+We studied the performance of HCM-FBA on a proof of concept metabolic network, along with two \emph{E. coli} networks.
+First, we showed the performance of HCM-FBA and HCM-EM were comparable for a proof of concept metabolic network, and a small model of anaerobic \textit{E. coli} metabolism.
+For the anaerobic case, both approaches described the experimental data.
+However HCM-FBA (which was within 5\% of HCM-EM and slightly better than HCM-EM for lactate secretion) had fewer modes and parameters.
+Next, HCM-FBA was applied to an aerobic \textit{E. coli} metabolic network that was not feasible for HCM-EM.
 Elementary mode decomposition of the core aerobic network generated over 150,000 elementary modes.
-Conversely, the HCM FBA approach described cellmass growth, and the shift from glucose to acetate consumption, with only 29 FBA modes.
+Conversely, the HCM-FBA approach described cellmass growth, and the shift from glucose to acetate consumption, with only 29 FBA modes.
 Global sensitivity analysis further showed that only five of the 29 FBA modes were critical to model performance.
 Removal of these modes crippled the model, but removal of the remaining 24 modes had a negligible impact.
-Taken together, HCM FBA is a viable alternative to HCM EM, especially for large networks where the generation of elementary modes is unfeasible.
+Thus, HCM-FBA is a viable alternative to HCM-EM, especially for large networks where the generation of elementary modes is infeasible.
 
@@
 The internal fluxes still follow the pseudo steady state approximation applied in FBA, but HCM uses a combination of pathways.
-HCM EM has been shown to have comparable internal flux estimates to MFA results\cite{2008_kim_varner_ramkrishna_BiotechProg}, this has not been shown for HCM FBA.
+HCM-EM has been shown to have comparable internal flux estimates to MFA results\cite{2008_kim_varner_ramkrishna_BiotechProg}, this has not been shown for HCM-FBA.
 The HCM framework has been shown to predict dynamic external fluxes using both elementary and FBA modes.
@@
 HCM overcomes this by incorparting the regulatory processes of substrate utilization based on its cybernetic arguments.
-HCM EM has good model performance for reduced networks, but is impractical for larger networks.
-HCM EM has been applied to a network with 67 reactions, requiring the elementary modes to be lumped into groups by complex weighting schemes and relying on experimental data\cite{2010_song_ramkrishna}.
-A shortcoming of the HCM EM framework is that EM calculations are computationally expensive and the number of modes exponentially increases for larger networks \cite{2004_lee_varner_ko_ieee}.
+HCM-EM has good model performance for reduced networks, but is impractical for larger networks.
+HCM-EM has been applied to a network with 67 reactions, requiring the elementary modes to be lumped into groups by complex weighting schemes and relying on experimental data\cite{2010_song_ramkrishna}.
+A shortcoming of the HCM-EM framework is that EM calculations are computationally expensive and the number of modes exponentially increases for larger networks \cite{2004_lee_varner_ko_ieee}.
 In contrast, FBA does not have the computational burden associated with calculating elementary modes.
@@
 
-%HCM FBA provides a practical and feasible approach to model large networks where elementary modes fail.
-%The disadvantage of HCM FBA is it requires \textit{a priori} knowledge of the inputs into the metabolic network.
+%HCM-FBA provides a practical and feasible approach to model large networks where elementary modes fail.
+%The disadvantage of HCM-FBA is it requires \textit{a priori} knowledge of the inputs into the metabolic network.
 %FBA modes do not have the thorough information as elementary modes, but provide sufficient pathways to model experimental observations.
-%HCM FBA may also be applied to mammalian networks, as long as FBA accounts for compartmentalization within mammalian systems.
+%HCM-FBA may also be applied to mammalian networks, as long as FBA accounts for compartmentalization within mammalian systems.
 
 \section{Materials and Methods}
-The HCM FBA approach is a modification of the HCM EM strategy of Kim et al.
+The HCM-FBA approach is a modification of the HCM-EM strategy of Kim et al.
 \cite{2008_kim_varner_ramkrishna_BiotechProg}.
-However, unlike HCM EM, we replaced elementary modes with flux balance analysis solutions.
+However, unlike HCM-EM, we replaced elementary modes with flux balance analysis solutions.
 The abundance of extracellular species $i$ ($x_{i}$), the pseudo enzyme $e_{l}$ and cellmass are governed by:

### 2016-03-17 Paper_v1.tex
Working in introduction

@@
 
-Metabolic networks, even for simple organisms such as \textit{E. coli}, contain thousands of intracellular metabolites and reactions that are highly interconnected.
-Constraints based models and convex analysis techniques have become the standard approaches to interrogate these complex networks.
+Biotechnology harnesses the power of metabolism to produce products that benefit society.
+Thus, a core challenge in biotechnology is the understanding of intracellular metabolic networks.
+Mathematical models of metabolic networks are useful tools to understand and ultimately predict how cells utilize nutrients to produce valuable products.
+Genome scale metabolic networks contain thousands of coupled intracellular metabolites and biochemical reactions.
+Constraints based techniques have become the standard approach to interrogating these complex networks.
 Constraints based methods such as flux balance analysis (FBA) \cite{2010_orth_NatBiotech} and convex network decomposition approaches such elementary modes (EMs) [REF] or extreme pathways (EPs) [REF] describe intracellular metabolism using the biochemical stoichiometry, and other constraints such as thermodynamically feasibility [REF]

### 2016-03-18 Paper_v1.tex
Still working on introduction

--- 2016-03-16 Paper_v1.tex
@@
 %thermodynamically feasible
+%Genome scale metabolic networks contain thousands of coupled intracellular metabolites and biochemical reactions.
 
-Metabolic networks, even for simple organisms such as \textit{E. coli}, contain thousands of intracellular metabolites and reactions that are highly interconnected.
-Constraints based models and convex analysis techniques have become the standard approaches to interrogate these complex networks.
+Biotechnology harnesses the power of metabolism to produce products that benefit society.
+Thus, a core challenge in biotechnology is understanding intracellular metabolic networks.
+Constraints based models are useful tools to understand and ultimately predict how cells utilize nutrients to produce valuable products.
 Constraints based methods such as flux balance analysis (FBA) \cite{2010_orth_NatBiotech} and convex network decomposition approaches such elementary modes (EMs) [REF] or extreme pathways (EPs) [REF] describe intracellular metabolism using the biochemical stoichiometry, and other constraints such as thermodynamically feasibility [REF]

### 2016-03-18 Paper_v1.tex
Working on introduction

@@
 Constraints based models are useful tools to understand and ultimately predict how cells utilize nutrients to produce valuable products.
-Constraints based methods such as flux balance analysis (FBA) \cite{2010_orth_NatBiotech} and convex network decomposition approaches such elementary modes (EMs) [REF] or extreme pathways (EPs) [REF] describe intracellular metabolism using the biochemical stoichiometry, and other constraints such as thermodynamically feasibility [REF]
-under pseudo-steady state conditions.
-However, elementary mode calculations are computational expensive and infeasible for genome scale networks\cite{2004_lee_varner_ko_ieee}.
-FBA is frequently used to study genome-scale networks and often results in good approximations for yields, however it fails to estimate intracellular metabolism due to its underdetermined solution space.
-Compared with FBA, MFA has good estimations of intracellular metabolism, but is dependent on experimental data.
-Just like FBA, HCM approaches cell metabolism as a resource allocation problem towards a specified objective.
+Constraints based methods such as flux balance analysis (FBA) \cite{2010_orth_NatBiotech} and convex network decomposition approaches such elementary modes (EMs) [REF]
+or extreme pathways (EPs) \cite{Schilling:2000aa}
+describe intracellular metabolism using the biochemical stoichiometry, and other constraints such as thermodynamically feasibility [REF] under pseudo-steady state conditions.
+FBA is frequently used to study genome-scale networks and often results in good approximations for yields.
+However, elementary mode calculations are computational expensive and infeasible for genome scale networks \cite{2004_lee_varner_ko_ieee}.
+
+Cybernetic models successfully predicted choice behavior in metabolism, e.g., diauxie behavior \cite{1986_kompala_ramkrishna_tsao_BiotechBioeng}, complex phenomena such as steady-state multiplicity \cite{2012_kim_ramkrishna_BiotechProg} as well as cell response to metabolic engineering strategies \cite{1999_varner_ramkrishna_MetaEng}.
+However, a perceived limitation of cybernetic models, has been their reductionist representation
+of biological mechanism, posing difficulties in formulation and requiring potentially unobservable parametric information.
+HCM approaches cell metabolism as a resource allocation problem towards a specified objective.
 But HCM has multiple pathways to select from (elementary modes) and can use a combination of modes to direct its resources through the cell.
 HCM has been shown to have comparable internal flux estimations to MFA and is able to estimate dynamic external fluxes with respect to experimental observations \cite{2008_kim_varner_ramkrishna_BiotechProg}.
-DFBA is able to account for dynamic external fluxes but fails to incorporate the regulatory processes involved with HCM without the use of boolean rules\cite{2001_covert_schilling_palsson}.
+DFBA is able to account for dynamic external fluxes but fails to incorporate the regulatory processes involved with HCM without the use of boolean rules \cite{2001_covert_schilling_palsson}.
 Despite the advantages of HCM, it is restricted to small networks due to the exponential increase of the number of elementary modes for large networks.
-Recent work has been done to reduce the number of elementary modes by lumping modes into groups using complex weighting schemes and experimental data\cite{2010_song_ramkrishna}, however, HCM is still limited by EM decomposition.
+Recent work has been done to reduce the number of elementary modes by lumping modes into groups using complex weighting schemes and experimental data \cite{2010_song_ramkrishna}, however, HCM is still limited by EM decomposition.
 
@@
 HCM-FBA performed similar to HCM-EM for the hypothetical network and a reduced anaerobic \textit{E. coli} network, despite having fewer parameters in each case.
-Next, HCM-FBA was applied to an aerobic \textit{E. coli} metabolic network that was not feasible for HCM-EM.
-The HCM-FBA approach described cellmass growth, and the shift from glucose to acetate consumption, with only a few modes.
+Next, HCM-FBA was applied to an aerobic \textit{E. coli} metabolic network that was infeasible for HCM-EM.
+HCM-FBA described cellmass growth, and the shift from glucose to acetate consumption, with only a few modes.
 Global sensitivity analysis allowed us to further reduce the aerobic \textit{E. coli} HCM-FBA model to the minimal model required to describe the data.
-Thus, we demonstrated that HCM-FBA is a promising approach for the development of reduced order dynamic metabolic models,
-and a viable alternative to the HCM-EM approach especially for large networks where the generation of elementary modes is infeasible.
+Thus, HCM-FBA is a promising approach for the development of reduced order dynamic metabolic models,
+and a viable alternative to HCM-EM especially for large networks where the generation of elementary modes is infeasible.
 
@@
 \begin{figure}[!t]\centering
-\includegraphics[width=0.40\textwidth]{./figs/Fig-1-GeneralModel-Results.pdf}
+\includegraphics[width=0.35\textwidth]{./figs/Fig-1-GeneralModel-Results.pdf}
 \caption{HCM-FBA proof of concept metabolic study.
@@
 \begin{figure}[!t]\centering
-\includegraphics[width=0.48\textwidth]{./figs/Fig-2-Ecoli-SimulationResults.pdf}
+\includegraphics[width=0.45\textwidth]{./figs/Fig-2-Ecoli-SimulationResults.pdf}
 \caption{HCM-FBA versus HCM-EM performance for small and large metabolic networks.

### 2016-03-19 Paper_v1.tex
Still working on the introduction

@@
 %Genome scale metabolic networks contain thousands of coupled intracellular metabolites and biochemical reactions.
+%Thus, a core challenge in biotechnology is understanding intracellular metabolic networks.
+%Cybernetic models successfully predicted choice behavior in metabolism, e.g., diauxie behavior \cite{1986_kompala_ramkrishna_tsao_BiotechBioeng}, complex phenomena such as steady-state multiplicity \cite{2012_kim_ramkrishna_BiotechProg} as well as cell the response of metabolism to metabolic engineering \cite{1999_varner_ramkrishna_MetaEng}.
 
 Biotechnology harnesses the power of metabolism to produce products that benefit society.
-Thus, a core challenge in biotechnology is understanding intracellular metabolic networks.
-Constraints based models are useful tools to understand and ultimately predict how cells utilize nutrients to produce valuable products.
-Constraints based methods such as flux balance analysis (FBA) \cite{2010_orth_NatBiotech} and convex network decomposition approaches such elementary modes (EMs) [REF]
+Constraints based models are useful tools to understand and ultimately predict how cells utilize nutrients to produce products.
+Constraints based methods such as flux balance analysis (FBA) \cite{2010_orth_NatBiotech} and network decomposition approaches such elementary modes (EMs) \cite{Schuster:2000aa}
 or extreme pathways (EPs) \cite{Schilling:2000aa}
-describe intracellular metabolism using the biochemical stoichiometry, and other constraints such as thermodynamically feasibility [REF] under pseudo-steady state conditions.
-FBA is frequently used to study genome-scale networks and often results in good approximations for yields.
-However, elementary mode calculations are computational expensive and infeasible for genome scale networks \cite{2004_lee_varner_ko_ieee}.
+model intracellular metabolism using the biochemical stoichiometry, and other constraints such as thermodynamically feasibility under pseudo-steady state conditions.
+Flux balance analysis has predicted the global performance of genome scale networks \cite{Covert:2004aa}.
+On the other hand, elementary modes (or extreme pathways) catalog all possible metabolic behaviors such that any flux distribution predicted by FBA is a convex combination
+of the elementary modes (or extreme pathways) \cite{Wiback:2003aa}.
+However, the calculation of elementary modes is computational expensive and currently infeasible for genome scale networks \cite{2004_lee_varner_ko_ieee}.
 
-Cybernetic models successfully predicted choice behavior in metabolism, e.g., diauxie behavior \cite{1986_kompala_ramkrishna_tsao_BiotechBioeng}, complex phenomena such as steady-state multiplicity \cite{2012_kim_ramkrishna_BiotechProg} as well as cell response to metabolic engineering strategies \cite{1999_varner_ramkrishna_MetaEng}.
-However, a perceived limitation of cybernetic models, has been their reductionist representation
-of biological mechanism, posing difficulties in formulation and requiring potentially unobservable parametric information.
-HCM approaches cell metabolism as a resource allocation problem towards a specified objective.
-But HCM has multiple pathways to select from (elementary modes) and can use a combination of modes to direct its resources through the cell.
-HCM has been shown to have comparable internal flux estimations to MFA and is able to estimate dynamic external fluxes with respect to experimental observations \cite{2008_kim_varner_ramkrishna_BiotechProg}.
-DFBA is able to account for dynamic external fluxes but fails to incorporate the regulatory processes involved with HCM without the use of boolean rules \cite{2001_covert_schilling_palsson}.
-Despite the advantages of HCM, it is restricted to small networks due to the exponential increase of the number of elementary modes for large networks.
-Recent work has been done to reduce the number of elementary modes by lumping modes into groups using complex weighting schemes and experimental data \cite{2010_song_ramkrishna}, however, HCM is still limited by EM decomposition.
+Cybernetic models are an alternative to the constraints based approach which hypothesize that metabolic control is the output of an optimal decision.
+Cybernetic models have predicted mutant behavior \cite{1999_varner_ramkrishna_MetaEng,Song:2012aa}, steady-state multiplicity \cite{2012_kim_ramkrishna_BiotechProg}, strain specific metabolism \cite{Song:2011aa} and have been used in bioprocess control applications \cite{Gadkar:2003aa}.
+Hybrid cybernetic models (HCM) have addressed earlier shortcomings of the approach by integrating cybernetic optimality concepts with elementary modes.
+HCMs dynamically choose combinations of elementary modes to achieve hypothetical physiological objectives.
+HCMs generate intracellular flux distributions consistent with other approaches such as metabolic flux analysis, and also describe extracellular measurements \cite{2008_kim_varner_ramkrishna_BiotechProg}.
+However, HCMs are restricted to networks which can be decomposed into elementary modes.
+
+%small networks due to the exponential increase of the number of elementary modes for large networks.
+%has been shown to have comparable internal flux estimations to MFA and is able to estimate dynamic external fluxes with respect to experimental observations
+%Despite the advantages of HCM, it is
+%Recent work has been done to reduce the number of elementary modes by lumping modes into groups using complex weighting schemes and experimental data \cite{2010_song_ramkrishna},
+%however, HCM is still limited by EM decomposition.
 
@@
 In this study, we developed HCM-FBA, an effective modeling technique to simulate metabolic dynamics.
-HCM-FBA is a modification of the hybrid cybernetic approach of Ramkrishna and coworkers \cite{2008_kim_varner_ramkrishna_BiotechProg}.
+HCM-FBA is a modification of the hybrid cybernetic approach of Ramkrishna and coworkers.
 HCM-FBA uses flux balance analysis solutions (instead of elementary modes) in conjunction with cybernetic control variables to dynamically simulate metabolism.
@@
 
-Hybrid cybernetic models are formulated to maximize substrate uptake based on cybernetic arguments.
-The internal fluxes still follow the pseudo steady state approximation applied in FBA, but HCM uses a combination of pathways.
-HCM-EM has been shown to have comparable internal flux estimates to MFA results\cite{2008_kim_varner_ramkrishna_BiotechProg}, this has not been shown for HCM-FBA.
-The HCM framework has been shown to predict dynamic external fluxes using both elementary and FBA modes.
-While DFBA already has the capacity to estimate dynamic external fluxes, it requires \textit{a priori} knowledge and boolean rules to capture the diauxic phenomena \cite{1994_varma_palsson_ApplEnvMicro,2002_Mahadevan_BiophysJ,2001_covert_schilling_palsson}.
-HCM overcomes this by incorparting the regulatory processes of substrate utilization based on its cybernetic arguments.
-HCM-EM has good model performance for reduced networks, but is impractical for larger networks.
-HCM-EM has been applied to a network with 67 reactions, requiring the elementary modes to be lumped into groups by complex weighting schemes and relying on experimental data\cite{2010_song_ramkrishna}.
-A shortcoming of the HCM-EM framework is that EM calculations are computationally expensive and the number of modes exponentially increases for larger networks \cite{2004_lee_varner_ko_ieee}.
-In contrast, FBA does not have the computational burden associated with calculating elementary modes.
-FBA is frequently used to study genome-scale networks\cite{2010_orth_NatBiotech} and can be used to generate modes for the HCM framework.
-This opens up the possibility of genome scale cybernetic models.
+%Hybrid cybernetic models are formulated to maximize substrate uptake based on cybernetic arguments.
+%The internal fluxes still follow the pseudo steady state approximation applied in FBA, but HCM uses a combination of pathways.
+%HCM-EM has been shown to have comparable internal flux estimates to MFA results\cite{2008_kim_varner_ramkrishna_BiotechProg}, this has not been shown for HCM-FBA.
+%The HCM framework has been shown to predict dynamic external fluxes using both elementary and FBA modes.
+%While DFBA already has the capacity to estimate dynamic external fluxes, it requires \textit{a priori} knowledge and boolean rules to capture the diauxic phenomena \cite{1994_varma_palsson_ApplEnvMicro,2002_Mahadevan_BiophysJ,2001_covert_schilling_palsson}.
+%HCM overcomes this by incorparting the regulatory processes of substrate utilization based on its cybernetic arguments.
+%HCM-EM has good model performance for reduced networks, but is impractical for larger networks.
+%HCM-EM has been applied to a network with 67 reactions, requiring the elementary modes to be lumped into groups by complex weighting schemes and relying on experimental data\cite{2010_song_ramkrishna}.
+%A shortcoming of the HCM-EM framework is that EM calculations are computationally expensive and the number of modes exponentially increases for larger networks \cite{2004_lee_varner_ko_ieee}.
+%In contrast, FBA does not have the computational burden associated with calculating elementary modes.
+%FBA is frequently used to study genome-scale networks\cite{2010_orth_NatBiotech} and can be used to generate modes for the HCM framework.
+%This opens up the possibility of genome scale cybernetic models.
 
@@
 where $z_{sl}$ denotes the uptake flux of substrate $x$ through mode $l$.
-In the anaerobic case we followed the assumption of Kim et al.\cite{2008_kim_varner_ramkrishna_BiotechProg},
-that formate decomposition occurs only outside the network (applicable for strain GJT001).
-Therefore its enzyme activity was assumed to be at its maximum level of unity.
-The reaction rate for formate was also modified following Kim et al.\cite{2008_kim_varner_ramkrishna_BiotechProg}.
-All numerical simulations were conducted in the high-performance computing language Julia (v0.4.2) \cite{Julia}.
-Species balances were solved using the SUNDIALS package for Julia \cite{Sundials}.
+All model equations were implemented in Julia (v0.4.2) \cite{Julia}, and solved using SUNDIALS \cite{Sundials}.
+
+%In the anaerobic case we followed the assumption of Kim et al.\cite{2008_kim_varner_ramkrishna_BiotechProg},
+%that formate decomposition occurs only outside the network (applicable for strain GJT001).
+%Therefore its enzyme activity was assumed to be at its maximum level of unity.
+%The reaction rate for formate was also modified following Kim et al.\cite{2008_kim_varner_ramkrishna_BiotechProg}.
 
@@
 The flux balance analysis problem was solved using the GNU Linear Programming Kit, GLPK version 4.52 \cite{GLPK}.
-For each FBA mode, the objective was to maximize either the growth rate, or the specific rate of byproduct formation from a specified starting substrate.
-Multiple FBA modes were calculated for each objective flux by allowing the oxygen and nitrate uptake rates to be either zero or maximal.
+For each FBA mode, the objective was to maximize either the specific growth rate, or the specific rate of byproduct formation.
+Multiple FBA modes were calculated for each objective flux by allowing the oxygen and nitrate uptake rates to vary.
 For aerobic metabolism, the specific oxygen and nitrate uptake rates were constrained to allow a maximum flux of 10 mM/gDW$\cdot$hr and 0.05 mM/gDW$\cdot$hr, respectively.
-Each flux vector was normalized by the growth flux, or by the specified objective flux.
+Each flux vector was normalized by the specified objective flux.
 

### 2016-03-23 Paper_v1.tex
updated discussion

@@
 HCM-FBA is a modification of the hybrid cybernetic approach of Ramkrishna and coworkers which uses flux balance analysis solutions
-(instead of elementary modes) in conjunction with cybernetic control variables to dynamically simulate metabolism.
+(instead of elementary modes) in conjunction with control variables to dynamically simulate metabolism.
 We compared the performance of HCM-FBA to HCM-EM for a proof of concept metabolic network, along with two \emph{E. coli} networks.
@@
 HCM-FBA is a modification of the hybrid cybernetic approach of Ramkrishna and coworkers.
-HCM-FBA uses flux balance analysis solutions (instead of elementary modes) in conjunction with cybernetic control variables to dynamically simulate metabolism.
+HCM-FBA uses flux balance analysis solutions (instead of elementary modes) in conjunction with control variables to dynamically simulate metabolism.
 We studied the performance of HCM-FBA on a proof of concept metabolic network, along with two \emph{E. coli} networks.
@@
 Thus, HCM-FBA is a viable alternative to HCM-EM, especially for large networks where the generation of elementary modes is infeasible.
+
+Hybrid cybernetic models are formulated to split the uptake flux among different pathways (defined as modes) to satisfy some chosen metabolic objective (i.e. maximum growth or substrate uptake).
+%The overall flux through each mode is determined by a kinetic uptake rate that acts as a weight for the utilization of that mode. 
+The HCM framework has been shown to predict dynamic external fluxes of cell growth, substrate utilization and byproduct formation. 
+HCM-EM has good model performance for reduced networks, but is often impractical for larger networks.
+However, Song and Ramkrishna address this shortcoming by lumping elementary modes into "families" each containing several modes with unifying characteristics based on metabolic function \cite{2010_song,Song:2011aa}.
+The elementary modes are lumped into groups by weighting schemes typically based on yields and may be further tuned with experimental data.
+The framework has been applied to a \textit{E. coli} network with 67 reactions and a \textit{Saccharomyces cerevisiae} network with 53 reactions, both had satisfactory fits to experimental data.
+The lumped EMs are considered convex basis vectors in yield space since it spans every physiological yield vector of extracellular metabolites. 
+The HCM-FBA framework is a different approach of reducing/lumping EMs, since the FBA mode is already a convex combination of EMs \cite{Wiback:2003aa}. 
+The advantage of HCM-FBA is it does not require the computational burden of calculating elementary modes.
+
+HCM-FBA provides a practical and feasible approach to model large networks where elementary modes fail.
+%The disadvantage of HCM-FBA is it requires \textit{a priori} knowledge of the inputs into FBA.
+FBA modes do not have the thorough information as elementary modes, but provide sufficient pathways to model experimental observations.
+HCM-EM has been shown to have comparable internal flux estimates to metabolic flux analysis results\cite{2008_kim_varner_ramkrishna_BiotechProg}, this has not been shown for HCM-FBA.
+Just as HCM-EM has been applied to \textit{S. cerevisiae}, we can apply the HCM-FBA framework to eukaryotes as long as FBA accounts for compartmentalization.
+FBA is frequently used to study genome-scale networks\cite{2010_orth_NatBiotech} and can be used to generate modes for the HCM framework.
+This opens up the possibility of genome scale cybernetic models.
 
@@
 \begin{eqnarray}\nonumber
-	\frac{dx_{i}}{dt}  & = &  \sum_{j = 1}^{\mathcal{R}}\sum_{l = 1}^{\mathcal{L}}\sigma_{ij}z_{jl}r_{l}\left(\mathbf{e},\mathbf{k},\mathbf{x}\right)c \qquad{i=1,\hdots,\mathcal{M}}\\\nonumber
-  \frac{de_{l}}{dt}  & = & \alpha_{l} + r_{E,l}\left(\mathbf{k},\mathbf{x}\right)u_{l} - \left(\beta_{l}+r_{G}\right)e_{l} \qquad l=1,\hdots,\mathcal{L} \\\nonumber
+	\frac{dx_{i}}{dt}  & = &  \sum_{j = 1}^{\mathcal{R}}\sum_{l = 1}^{\mathcal{L}}\sigma_{ij}z_{jl}q_{l}\left(\mathbf{e},\mathbf{k},\mathbf{x}\right)c \qquad{i=1,\hdots,\mathcal{M}}\\\nonumber
+  \frac{de_{l}}{dt}  & = & \alpha_{l} + r_{El}\left(\mathbf{k},\mathbf{x}\right)u_{l} - \left(\beta_{l}+r_{G}\right)e_{l} \qquad l=1,\hdots,\mathcal{L} \\\nonumber
   \frac{dc}{dt} & = & r_{G}c
@@
 Extracellular species balances were subject to the initial conditions $\mathbf{x}\left(t_{o}\right) = \mathbf{x}_{o}$ determined from experimental data.
-The term $r_{l}\left(\mathbf{e},\mathbf{k},\mathbf{x}\right)$ denotes the specific rate of flux through mode $l$, and was written as the product of a kinetic term ($\bar{r}_{l}$) and a cybernetic control variable governing enzyme activity.
+The term $q_{l}\left(\mathbf{e},\mathbf{k},\mathbf{x}\right)$ denotes the specific rate of flux through mode $l$, and was written as the product of a kinetic term ($\bar{q}_{l}$) and a  control variable governing enzyme activity.
 Flux through each mode was catalyzed by a pseudo enzyme $e_{l}$, where enzyme $e_{l}$ was synthesized at the regulated specific rate $r_{E,l}\left(\mathbf{k},\mathbf{x}\right)$ and constitutively at the rate $\alpha_{l}$.
-The term $r_{E,l}$ denotes the specific rate enzyme synthesis for enzyme $l$, and $u_{l}$ denotes the cybernetic variable controlling the synthesis of enzyme $l$.
+The term $r_{E,l}$ denotes the specific rate enzyme synthesis for enzyme $l$, and $u_{l}$ denotes the variable controlling the synthesis of enzyme $l$.
 The term $\beta_{l}$ denotes the rate constant governing enzyme degradation, and $r_{G}$ denotes the growth rate through all modes.
 The specific rate of flux through an FBA mode, and the specific rate of enzyme synthesis were modeled using saturation kinetics.
-All enzyme initial conditions were set to 0.9 for the anaerobic case, and 0.8 for the aerobic case.
+%All enzyme initial conditions were set to 0.9 for the anaerobic case, and 0.8 for the aerobic case.
 Lastly, cellmass was produced at the specific growth rate:
 \begin{equation}
-	r_{G}  = \sum_{l = 1}^{\mathcal{L}}z_{\mu l}r_{l}\left(\mathbf{e},\mathbf{k},\mathbf{x}\right)
+	r_{G}  = \sum_{l = 1}^{\mathcal{L}}z_{\mu l}q_{l}\left(\mathbf{e},\mathbf{k},\mathbf{x}\right)
 \end{equation}
 where $z_{\mu l}$ denotes the growth flux $\mu$ through mode $l$.
-The cybernetic control variables $u_{l}$ and $v_{l}$, which control the synthesis and activity of each enzyme respectively, were given by:
+The control variables $u_{l}$ and $v_{l}$, which control the synthesis and activity of each enzyme respectively, were given by:
 \begin{align*}
-	u_{l}  = \frac{z_{sl}\bar{r}_{l}}{\sum\limits_{l = 1}^{\mathcal{L}}z_{sl}\bar{r}_{l}} && v_{l} = \frac{z_{sl}\bar{r}_{l}}{\max\limits_{\mathbf{l}} z_{sl}\bar{r}_{l}}
+	u_{l}  = \frac{z_{sl}\bar{q}_{l}}{\sum\limits_{l = 1}^{\mathcal{L}}z_{sl}\bar{q}_{l}} && v_{l} = \frac{z_{sl}\bar{q}_{l}}{\max\limits_{\mathbf{l}} z_{sl}\bar{q}_{l}}
 \end{align*}
 where $z_{sl}$ denotes the uptake flux of substrate $x$ through mode $l$.
-All model equations were implemented in Julia (v0.4.2) \cite{Julia}, and solved using SUNDIALS \cite{Sundials}.
+All model equations were implemented in Julia (v.0.4.2) \cite{Julia}, and solved using SUNDIALS \cite{Sundials}. The model code is available online:
 

### 2016-03-23 Paper_v1.tex
new fig 1

@@
 \begin{figure}[!t]\centering
-\includegraphics[width=0.35\textwidth]{./figs/Fig-1-GeneralModel-Results.pdf}
-\caption{HCM-FBA proof of concept metabolic study.
-A: Prototypical network with six metabolites and seven reactions.
+\includegraphics[width=0.45\textwidth]{./figs/Fig-1-Method-General.pdf}
+\caption{HCM proof of concept metabolic study.
+A: HCM is formulated to split the uptake flux among different pathways (defined as modes) to satisfy some chosen metabolic objective.
+HCM-EM considers all possible modes within a network, whereas HCM-FBA combines modes by FBA.
+B: Prototypical network with six metabolites and seven reactions.
 Intracellular cellmass precursors $A,B$ and $C$ are balanced (no accumulation) while the extracellular metabolites ($A_{e},B_{e}$ and $C_{e}$) are not balanced (can accumulate).
 The oval denotes the cell boundary, $q_{j}$ is the jth flux across the boundary and $v_{k}$ denotes the kth intracellular flux.
-B: Simulation of extracellular metabolite trajectories using HCM-FBA (solid line) versus HCM-EM (points) for the proof of concept metabolic network.
+C: Simulation of extracellular metabolite trajectories using HCM-FBA (solid line) versus HCM-EM (points) for the proof of concept metabolic network.
 }\label{fig:model-fitting}

### 2016-03-24 Paper_v1.tex
Final edits

@@
 \IEEEauthorblockA{\IEEEauthorrefmark{2}School of Chemical and Biomolecular Engineering, Cornell University, Ithaca, NY 14850 USA}
-\thanks{Manuscript received December 1, 2012; revised September 17, 2014.
-Corresponding author: J. Varner (email:jdvarner@purdue.edu).}}
+\thanks{Corresponding author: J. Varner (email:jdvarner@purdue.edu).}}
 
@@
 Hybrid cybernetic models in combination with elementary modes (HCM-EM) provides a route to model cellular metabolism.
-However, this framework is limited to smaller networks due to the computational cost of calculating elementary modes and a high number of solutions for large networks.
+However, this framework is limited to small intracellular networks due to the computational burden of calculating elementary modes.
 In this study, we developed the hybrid cybernetic modeling with flux balance analysis or HCM-FBA technique, which uses flux balance solutions instead of elementary modes,
 to dynamically model metabolism.
-Flux balance analysis provides a sufficient number of modes (metabolic options) for the network to estimate experimental observations.
-We show HCM-FBA has comparable model performance to HCM-EM for a hypothetical proof of concept metabolic network and for a reduced anaerobic \textit{E. coli} network.
-HCM-FBA was applied to a larger metabolic network where 29 FBA modes were computed, in contrast to over 66,000 elementary modes.
-A sensitivity analysis was performed to reduce the number of FBA modes from 29 to 5.
-This reduction maintained a robust model performance to fit experimental observations.
+We show HCM-FBA has comparable performance to HCM-EM for a hypothetical proof of concept metabolic network and for a reduced anaerobic \textit{E. coli} network.
+HCM-FBA was applied to a larger metabolic network of aerobic \textit{E. coli} metabolism which was infeasible for HCM-EM (29 FBA modes, versus more than 153,000 elementary modes).
+Global sensitivity analysis further reduced the number of FBA modes required to describe the data from 29 to five, while maintaining model fit.
+Thus, HCM-FBA is an alternative to HCM-EM for large networks where the generation of elementary modes is infeasible.
 
@@
 Biotechnology harnesses the power of metabolism to produce products that benefit society.
-Constraints based models are useful tools to understand and ultimately predict how cells utilize nutrients to produce products.
+Constraints based models are important tools to understand and ultimately predict how cells utilize nutrients to produce products.
 Constraints based methods such as flux balance analysis (FBA) \cite{2010_orth_NatBiotech} and network decomposition approaches such elementary modes (EMs) \cite{Schuster:2000aa}
@@
 \caption{HCM proof of concept metabolic study.
-A: HCM is formulated to split the uptake flux among different pathways (defined as modes) to satisfy some chosen metabolic objective.
-HCM-EM considers all possible modes within a network, whereas HCM-FBA combines modes by FBA.
+A: HCMs distribute uptake and secretion fluxes amongst different pathways. For HCM-EM, these pathways are elementary modes, for HCM-FBA they are flux balance analysis solutions.
+Thus, HCM-EM combines all possible modes within a network, whereas HCM-FBA combines only steady-state paths estimated by flux balance analysis.
 B: Prototypical network with six metabolites and seven reactions.
@@
 HCM-FBA captured the shift from glucose to acetate consumption for a model of aerobic \textit{E. coli} metabolism that was infeasible for HCM-EM (Fig. \ref{fig:ecoli}, right).
-A core \emph{E. coli} metabolic network, consisting of 60 metabolites and 105 reactions, was constructed from literature \cite{2007_schuetz_etal_MolSysBio,2006_Palsson_model}.
+An \emph{E. coli} metabolic network (60 metabolites and 105 reactions) was constructed from literature \cite{2007_schuetz_etal_MolSysBio,2006_Palsson_model}.
 Elementary mode decomposition of this network (and thus HCM-EM) was not feasible; 153,000 elementary modes were generated before the calculation became infeasible.
@@
 Removal of these modes crippled the model, but removal of the remaining 24 modes had a negligible impact.
-Thus, HCM-FBA is a viable alternative to HCM-EM, especially for large networks where the generation of elementary modes is infeasible.
+Thus, HCM-FBA is an alternative approach to HCM-EM, especially for large networks where the generation of elementary modes is infeasible.
 
-Hybrid cybernetic models are formulated to split the uptake flux among different pathways (defined as modes) to satisfy some chosen metabolic objective (i.e. maximum growth or substrate uptake).
+We have shown that HCM-FBA is an alternative approach to model large metabolic networks where elementary modes calculations are infeasible.
+However, there are additional studies that should be performed to further test the HCM-FBA approach.
+First, the intracellular flux distribution predicted by HCM-FBA should be compared to HCM-EM, and to flux measurements calculated using metabolic flux analysis or flux balance analysis in combination with carbon labeling.
+HCM-EM has comparable internal flux estimates compared to metabolic flux analysis results \cite{2008_kim_varner_ramkrishna_BiotechProg}.
+Next, we should compare the performance of HCM-FBA to lumped hybrid cybernetic models (L-HCM).
+L-HCMs, which combine elementary modes into mode families based upon metabolic function \cite{2010_song,Song:2011aa},
+has been applied to an \textit{E. coli} network with 67 reactions and a \textit{Saccharomyces cerevisiae} network (excluding compartmentalization) with 53 reactions;
+both cases had satisfactory fits to extracellular experimental data.
+However, while L-HCM reduces the dimension of modes participating in the optimal decision, it still requires the calculation of modes which for networks of even moderate size is not possible.
+On the other had, the generation of flux balance solutions (which are convex combinations of the elementary modes) is trivial, even for genome scale metabolic networks.
+Thus, HCM-FBA opens up the possibility of manageable dynamic genome scale models of bacterial and perhaps even mammalian systems.
+
+%Hybrid cybernetic models are formulated to split the uptake flux among different pathways (defined as modes) to satisfy some chosen metabolic objective (i.e. maximum growth or substrate uptake).
 %The overall flux through each mode is determined by a kinetic uptake rate that acts as a weight for the utilization of that mode.
-The HCM framework has been shown to predict dynamic external fluxes of cell growth, substrate utilization and byproduct formation. 
-HCM-EM has good model performance for reduced networks, but is often impractical for larger networks.
-However, Song and Ramkrishna address this shortcoming by lumping elementary modes into "families" each containing several modes with unifying characteristics based on metabolic function \cite{2010_song,Song:2011aa}.
-The elementary modes are lumped into groups by weighting schemes typically based on yields and may be further tuned with experimental data.
-The framework has been applied to a \textit{E. coli} network with 67 reactions and a \textit{Saccharomyces cerevisiae} network with 53 reactions, both had satisfactory fits to experimental data.
-The lumped EMs are considered convex basis vectors in yield space since it spans every physiological yield vector of extracellular metabolites. 
-The HCM-FBA framework is a different approach of reducing/lumping EMs, since the FBA mode is already a convex combination of EMs \cite{Wiback:2003aa}. 
-The advantage of HCM-FBA is it does not require the computational burden of calculating elementary modes.
-
-HCM-FBA provides a practical and feasible approach to model large networks where elementary modes fail.
+%The HCM framework has been shown to predict dynamic external fluxes of cell growth, substrate utilization and byproduct formation.
+%HCM-EM has good model performance for reduced networks, but is often impractical for larger networks.
+%However, Song and Ramkrishna address this shortcoming by lumping .
+%The elementary modes are lumped into groups by weighting schemes typically based on yields and may be further tuned with experimental data.
+%The lumped EMs are considered convex basis vectors in yield space since it spans every physiological yield vector of extracellular metabolites.
+%The HCM-FBA framework is a different approach of reducing/lumping EMs, since the FBA mode is already a convex combination of EMs \cite{Wiback:2003aa}.
+%The advantage of HCM-FBA is it does not require the computational burden of calculating elementary modes.
 %The disadvantage of HCM-FBA is it requires \textit{a priori} knowledge of the inputs into FBA.
-FBA modes do not have the thorough information as elementary modes, but provide sufficient pathways to model experimental observations.
-HCM-EM has been shown to have comparable internal flux estimates to metabolic flux analysis results\cite{2008_kim_varner_ramkrishna_BiotechProg}, this has not been shown for HCM-FBA.
-Just as HCM-EM has been applied to \textit{S. cerevisiae}, we can apply the HCM-FBA framework to eukaryotes as long as FBA accounts for compartmentalization.
-FBA is frequently used to study genome-scale networks\cite{2010_orth_NatBiotech} and can be used to generate modes for the HCM framework.
-This opens up the possibility of genome scale cybernetic models.
-
+%FBA modes do not have the thorough information as elementary modes, but provide sufficient pathways to model experimental observations.
+%, this has not been shown for HCM-FBA.
+%Just as HCM-EM has been applied to \textit{S. cerevisiae}, we can apply the HCM-FBA framework to eukaryotes as long as FBA accounts for compartmentalization.
+%FBA is frequently used to study genome-scale networks\cite{2010_orth_NatBiotech} and can be used to generate modes for the HCM framework.
 %Hybrid cybernetic models are formulated to maximize substrate uptake based on cybernetic arguments.
@@
 %This opens up the possibility of genome scale cybernetic models.
-
 %HCM-FBA provides a practical and feasible approach to model large networks where elementary modes fail.
@@
 \section{Materials and Methods}
-The HCM-FBA approach is a modification of the HCM-EM strategy of Kim et al.
-\cite{2008_kim_varner_ramkrishna_BiotechProg}.
-However, unlike HCM-EM, we replaced elementary modes with flux balance analysis solutions.
-The abundance of extracellular species $i$ ($x_{i}$), the pseudo enzyme $e_{l}$ and cellmass are governed by:
+The HCM-FBA approach is a modification of HCM-EM, where elementary modes are replaced with flux balance analysis solutions.
+Thus, extracellular variables are dynamic, while intracellular metabolites are at a pseudo steady state.
+The abundance of extracellular species $i$ ($x_{i}$), the pseudo enzyme $e_{l}$ (catalyzes flux through mode l) and cellmass are governed by:
 \begin{eqnarray}\nonumber
@@
 Extracellular species balances were subject to the initial conditions $\mathbf{x}\left(t_{o}\right) = \mathbf{x}_{o}$ determined from experimental data.
-The term $q_{l}\left(\mathbf{e},\mathbf{k},\mathbf{x}\right)$ denotes the specific rate of flux through mode $l$, and was written as the product of a kinetic term ($\bar{q}_{l}$) and a  control variable governing enzyme activity.
-Flux through each mode was catalyzed by a pseudo enzyme $e_{l}$, where enzyme $e_{l}$ was synthesized at the regulated specific rate $r_{E,l}\left(\mathbf{k},\mathbf{x}\right)$ and constitutively at the rate $\alpha_{l}$.
+The term $q_{l}\left(\mathbf{e},\mathbf{k},\mathbf{x}\right)$ denotes the specific uptake/secretion rate for mode $l$;
+$q_{l}\left(\mathbf{e},\mathbf{k},\mathbf{x}\right)$ is the product of a kinetic term ($\bar{q}_{l}$) and a control variable governing enzyme activity.
+Flux through each mode was catalyzed by a pseudo enzyme $e_{l}$, synthesized at the regulated specific rate $r_{E,l}\left(\mathbf{k},\mathbf{x}\right)$ and constitutively at the rate $\alpha_{l}$.
 The term $r_{E,l}$ denotes the specific rate enzyme synthesis for enzyme $l$, and $u_{l}$ denotes the variable controlling the synthesis of enzyme $l$.
-The term $\beta_{l}$ denotes the rate constant governing enzyme degradation, and $r_{G}$ denotes the growth rate through all modes.
-The specific rate of flux through an FBA mode, and the specific rate of enzyme synthesis were modeled using saturation kinetics.
-%All enzyme initial conditions were set to 0.9 for the anaerobic case, and 0.8 for the aerobic case.
+The term $\beta_{l}$ denotes the rate constant governing non-specific enzyme degradation, and $r_{G}$ denotes the specific growth rate through all modes.
+The specific uptake/secretion rates, and the specific rate of enzyme synthesis were modeled using saturation kinetics.
 Lastly, cellmass was produced at the specific growth rate:
@@
 \begin{align*}
-	u_{l}  = \frac{z_{sl}\bar{q}_{l}}{\sum\limits_{l = 1}^{\mathcal{L}}z_{sl}\bar{q}_{l}} && v_{l} = \frac{z_{sl}\bar{q}_{l}}{\max\limits_{\mathbf{l}} z_{sl}\bar{q}_{l}}
+	u_{l}  = \frac{z_{sl}\bar{q}_{l}}{\sum\limits_{l = 1}^{\mathcal{L}}z_{sl}\bar{q}_{l}} && v_{l} = \frac{z_{sl}\bar{q}_{l}}{\max\limits_{l=1,\hdots,\mathcal{L}}z_{sl}\bar{q}_{l}}
 \end{align*}
 where $z_{sl}$ denotes the uptake flux of substrate $x$ through mode $l$.
-All model equations were implemented in Julia (v.0.4.2) \cite{Julia}, and solved using SUNDIALS \cite{Sundials}. The model code is available online:
+All model equations were implemented in Julia (v.0.4.2) \cite{Julia}, and solved using SUNDIALS \cite{Sundials}.
+The model code for each case study is available for download at http://www.varnerlab.org.
 

### 2016-03-24 Paper_v1.tex
Final draft

@@
 First, the intracellular flux distribution predicted by HCM-FBA should be compared to HCM-EM, and to flux measurements calculated using metabolic flux analysis or flux balance analysis in combination with carbon labeling.
-HCM-EM has comparable internal flux estimates compared to metabolic flux analysis results \cite{2008_kim_varner_ramkrishna_BiotechProg}.
-Next, we should compare the performance of HCM-FBA to lumped hybrid cybernetic models (L-HCM).
+HCM-EM had predicted intracellular fluxes that were similar to metabolic flux analysis results \cite{2008_kim_varner_ramkrishna_BiotechProg}.
+Next, the performance of HCM-FBA to lumped hybrid cybernetic models (L-HCM) should be compared.
 L-HCMs, which combine elementary modes into mode families based upon metabolic function \cite{2010_song,Song:2011aa},
-has been applied to an \textit{E. coli} network with 67 reactions and a \textit{Saccharomyces cerevisiae} network (excluding compartmentalization) with 53 reactions;
+have been applied to an \textit{E. coli} network with 67 reactions and a \textit{Saccharomyces cerevisiae} network (excluding compartmentalization) with 53 reactions;
 both cases had satisfactory fits to extracellular experimental data.
-However, while L-HCM reduces the dimension of modes participating in the optimal decision, it still requires the calculation of modes which for networks of even moderate size is not possible.
+However, while L-HCM reduces the dimension of possible alternative modes that must be considered, it still requires the calculation of initial set of modes.
+For metabolic networks of even moderate size this may not possible.
 On the other had, the generation of flux balance solutions (which are convex combinations of the elementary modes) is trivial, even for genome scale metabolic networks.
-Thus, HCM-FBA opens up the possibility of manageable dynamic genome scale models of bacterial and perhaps even mammalian systems.
+Thus, HCM-FBA opens up the possibility of dynamic genome scale models of bacterial and perhaps even mammalian metabolism.
 
@@
 For aerobic metabolism, the specific oxygen and nitrate uptake rates were constrained to allow a maximum flux of 10 mM/gDW$\cdot$hr and 0.05 mM/gDW$\cdot$hr, respectively.
-Each flux vector was normalized by the specified objective flux.
+Each FBA mode was normalized by the specified objective flux.
 

### 2016-03-26 Paper_v1.tex
Final Edits

@@
 \usepackage{mathtools}
+\usepackage[square,sort,comma,numbers]{natbib}
+\renewcommand{\bibfont}{\footnotesize}
 
@@
 However, this framework is limited to small intracellular networks due to the computational burden of calculating elementary modes.
-In this study, we developed the hybrid cybernetic modeling with flux balance analysis or HCM-FBA technique, which uses flux balance solutions instead of elementary modes,
+In this study, we developed the hybrid cybernetic modeling with flux balance analysis or HCM-FBA technique which uses flux balance solutions instead of elementary modes
 to dynamically model metabolism.
 We show HCM-FBA has comparable performance to HCM-EM for a hypothetical proof of concept metabolic network and for a reduced anaerobic \textit{E. coli} network.
-HCM-FBA was applied to a larger metabolic network of aerobic \textit{E. coli} metabolism which was infeasible for HCM-EM (29 FBA modes, versus more than 153,000 elementary modes).
-Global sensitivity analysis further reduced the number of FBA modes required to describe the data from 29 to five, while maintaining model fit.
+HCM-FBA was applied to a larger metabolic network of aerobic \textit{E. coli} metabolism which was infeasible for HCM-EM (29 FBA modes versus more than 153,000 elementary modes).
+Global sensitivity analysis further reduced the number of FBA modes required to describe the data from 29 to 5, while maintaining model fit.
 Thus, HCM-FBA is an alternative to HCM-EM for large networks where the generation of elementary modes is infeasible.
@@
 Biotechnology harnesses the power of metabolism to produce products that benefit society.
-Constraints based models are important tools to understand and ultimately predict how cells utilize nutrients to produce products.
-Constraints based methods such as flux balance analysis (FBA) \cite{2010_orth_NatBiotech} and network decomposition approaches such elementary modes (EMs) \cite{Schuster:2000aa}
+Constraints based models are important tools to understand and ultimately to predict how cells utilize nutrients to produce products.
+Constraints based methods such as flux balance analysis (FBA) \cite{2010_orth_NatBiotech} and network decomposition approaches such as elementary modes (EMs) \cite{Schuster:2000aa}
 or extreme pathways (EPs) \cite{Schilling:2000aa}
-model intracellular metabolism using the biochemical stoichiometry, and other constraints such as thermodynamically feasibility under pseudo-steady state conditions.
-Flux balance analysis has predicted the global performance of genome scale networks \cite{Covert:2004aa}.
-On the other hand, elementary modes (or extreme pathways) catalog all possible metabolic behaviors such that any flux distribution predicted by FBA is a convex combination
-of the elementary modes (or extreme pathways) \cite{Wiback:2003aa}.
-However, the calculation of elementary modes is computational expensive and currently infeasible for genome scale networks \cite{2004_lee_varner_ko_ieee}.
+model intracellular metabolism using the biochemical stoichiometry and other constraints such as thermodynamical feasibility under pseudo-steady state conditions.
+FBA has predicted the global performance of genome scale networks \cite{Covert:2004aa}.
+On the other hand, EMs (or EPs) catalog all possible metabolic behaviors such that any flux distribution predicted by FBA is a convex combination
+of the EMs (or EPs) \cite{Wiback:2003aa}.
+However, the calculation of elementary modes is computationally expensive and currently infeasible for genome scale networks \cite{2004_lee_varner_ko_ieee}.
 
 Cybernetic models are an alternative to the constraints based approach which hypothesize that metabolic control is the output of an optimal decision.
-Cybernetic models have predicted mutant behavior \cite{1999_varner_ramkrishna_MetaEng,Song:2012aa}, steady-state multiplicity \cite{2012_kim_ramkrishna_BiotechProg}, strain specific metabolism \cite{Song:2011aa} and have been used in bioprocess control applications \cite{Gadkar:2003aa}.
-Hybrid cybernetic models (HCM) have addressed earlier shortcomings of the approach by integrating cybernetic optimality concepts with elementary modes.
-HCMs dynamically choose combinations of elementary modes to achieve hypothetical physiological objectives.
-HCMs generate intracellular flux distributions consistent with other approaches such as metabolic flux analysis, and also describe extracellular measurements \cite{2008_kim_varner_ramkrishna_BiotechProg}.
-However, HCMs are restricted to networks which can be decomposed into elementary modes.
+Cybernetic models have predicted mutant behavior \cite{1999_varner_ramkrishna_MetaEng,Song:2012aa}, steady-state multiplicity \cite{2012_kim_ramkrishna_BiotechProg}, strain specific metabolism \cite{Song:2011aa}, and have been used in bioprocess control applications \cite{Gadkar:2003aa}.
+Hybrid cybernetic models (HCM) have addressed earlier shortcomings of the approach by integrating cybernetic optimality concepts with EMs.
+HCMs dynamically choose combinations of modes to achieve hypothetical physiological objectives (Fig. \ref{fig:model-fitting}A).
+HCMs generate intracellular flux distributions consistent with other approaches such as metabolic flux analysis (MFA), and also describe extracellular measurements \cite{2008_kim_varner_ramkrishna_BiotechProg}.
+However, HCMs are restricted to networks which can be decomposed into EMs.
 
@@
 
-In this study, we developed the hybrid cybernetic modeling with flux balance analysis or HCM-FBA technique.
-HCM-FBA is a modification of the hybrid cybernetic approach of Ramkrishna and coworkers which uses flux balance analysis solutions
-(instead of elementary modes) in conjunction with control variables to dynamically simulate metabolism.
-We compared the performance of HCM-FBA to HCM-EM for a proof of concept metabolic network, along with two \emph{E. coli} networks.
-HCM-FBA performed similar to HCM-EM for the hypothetical network and a reduced anaerobic \textit{E. coli} network, despite having fewer parameters in each case.
+In this study, we developed the hybrid cybernetic modeling with flux balance analysis (HCM-FBA) technique.
+HCM-FBA is a modification of the hybrid cybernetic approach of Ramkrishna and coworkers \cite{2008_kim_varner_ramkrishna_BiotechProg} which uses FBA solutions
+(instead of EMs) in conjunction with control variables to dynamically simulate metabolism.
+We compared the performance of HCM-FBA to HCM-EM for a proof of concept metabolic network along with two \emph{E. coli} networks.
+HCM-FBA performed comparably to HCM-EM for the hypothetical network and a reduced anaerobic \textit{E. coli} network, despite having fewer parameters in each case.
 Next, HCM-FBA was applied to an aerobic \textit{E. coli} metabolic network that was infeasible for HCM-EM.
-HCM-FBA described cellmass growth, and the shift from glucose to acetate consumption, with only a few modes.
+HCM-FBA described cellmass growth and the shift from glucose to acetate consumption with only a few modes.
 Global sensitivity analysis allowed us to further reduce the aerobic \textit{E. coli} HCM-FBA model to the minimal model required to describe the data.
-Thus, HCM-FBA is a promising approach for the development of reduced order dynamic metabolic models,
-and a viable alternative to HCM-EM especially for large networks where the generation of elementary modes is infeasible.
+Thus, HCM-FBA is a promising approach for the development of reduced order dynamic metabolic models
+and a viable alternative to HCM-EM, especially for large networks where the generation of EMs is infeasible.
 
@@
 \caption{HCM proof of concept metabolic study.
-A: HCMs distribute uptake and secretion fluxes amongst different pathways. For HCM-EM, these pathways are elementary modes, for HCM-FBA they are flux balance analysis solutions.
-Thus, HCM-EM combines all possible modes within a network, whereas HCM-FBA combines only steady-state paths estimated by flux balance analysis.
+A: HCMs distribute uptake and secretion fluxes amongst different pathways. For HCM-EM, these pathways are elementary modes; for HCM-FBA these are flux balance analysis solutions.
+HCM-EM combines all possible modes within a network; whereas HCM-FBA combines only steady-state paths estimated by flux balance analysis.
 B: Prototypical network with six metabolites and seven reactions.
-Intracellular cellmass precursors $A,B$ and $C$ are balanced (no accumulation) while the extracellular metabolites ($A_{e},B_{e}$ and $C_{e}$) are not balanced (can accumulate).
-The oval denotes the cell boundary, $q_{j}$ is the jth flux across the boundary and $v_{k}$ denotes the kth intracellular flux.
+Intracellular cellmass precursors $A,B$, and $C$ are balanced (no accumulation) while the extracellular metabolites ($A_{e},B_{e}$, and $C_{e}$) are not balanced (can accumulate).
+The oval denotes the cell boundary, $q_{j}$ is the $j$th flux across the boundary, and $v_{k}$ denotes the $k$th intracellular flux.
 C: Simulation of extracellular metabolite trajectories using HCM-FBA (solid line) versus HCM-EM (points) for the proof of concept metabolic network.
@@
 HCM-FBA was equivalent to HCM-EM for a proof of concept metabolic network (Fig. \ref{fig:model-fitting}).
-The proof of concept network, consisting of six metabolites and seven reactions (Fig. \ref{fig:model-fitting}A), generated three FBA and six elementary modes.
-Using the elementary modes and synthetic parameters, we generated test data from which we estimated the HCM-FBA model parameters.
-The best fit HCM-FBA model replicated the synthetic data (Fig. \ref {fig:model-fitting}B).
+The proof of concept network, consisting of 6 metabolites and 7 reactions (Fig. \ref{fig:model-fitting}B), generated 3 FBA modes and 6 EMs.
+Using the EMs and synthetic parameters, we generated test data from which we estimated the HCM-FBA model parameters.
+The best fit HCM-FBA model replicated the synthetic data (Fig. \ref {fig:model-fitting}C).
 The HCM-EM and HCM-FBA kinetic parameters were not quantitatively identical, but had similar orders of magnitude;
-the FBA approach had three fewer modes, thus you would not expect identical parameter values.
-Taken together, the HCM-FBA approach replicated synthetic data generated using HCM-EM, despite having three fewer modes.
+the FBA approach had 3 fewer modes, thus identical parameter values were not expected.
+Taken together, the HCM-FBA approach replicated synthetic data generated by HCM-EM, despite having 3 fewer modes.
 Next, we tested the ability of HCM-FBA to replicate experimental data.
@@
 
-The performance of HCM-FBA was equivalent to HCM-EM for anaerobic \textit{E. coli} metabolism (Fig. \ref{fig:ecoli}, left).
-We constructed an anaerobic \textit{E. coli} network, consisting of 12 reactions and 19 metabolites, which generated nine elementary and seven FBA modes.
-HCM-EM reproduced cellmass, glucose and byproduct trajectories
-using the kinetic parameters reported by Kim et al. \cite{2008_kim_varner_ramkrishna_BiotechProg} (Fig. \ref{fig:ecoli}A, points versus dashed).
-On the other hand, HCM-FBA model parameters were estimated in this study from the Kim et al. data set using simulated annealing.
+The performance of HCM-FBA was equivalent to HCM-EM for anaerobic \textit{E. coli} metabolism (Fig. \ref{fig:ecoli}A).
+We constructed an anaerobic \textit{E. coli} network, consisting of 12 reactions and 19 metabolites, which generated 7 FBA modes and 9 EMs.
+HCM-EM reproduced cellmass, glucose, and byproduct trajectories using the kinetic parameters reported by Kim \textit{et al.} \cite{2008_kim_varner_ramkrishna_BiotechProg} (Fig.
+\ref{fig:ecoli}A, points versus dashed).
+HCM-FBA model parameters were estimated in this study from the Kim \textit{et al.} data set using simulated annealing.
 Overall, HCM-FBA performed within 5\% of HCM-EM (on a residual standard error basis) for the anaerobic \textit{E. coli} data (Fig. \ref {fig:ecoli}A, solid),
-despite having two fewer modes and four fewer parameters (17 versus 21 parameters).
+despite having 2 fewer modes and 4 fewer parameters (17 versus 21 parameters).
 Thus, while both HCM-EM and HCM-FBA described the experimental data, HCM-FBA did so with fewer modes and parameters.
@@
 A: Batch anaerobic \textit{E. coli} fermentation data versus HCM-FBA (solid) and HCM-EM (dashed).
-The experimental data was reproduced from from Kim et. al \cite{2008_kim_varner_ramkrishna_BiotechProg}. Error bars represent the 90\% confidence interval.
+The experimental data was reproduced from Kim \textit{et al.} \cite{2008_kim_varner_ramkrishna_BiotechProg}. Error bars represent the 90\% confidence interval.
 B: Batch aerobic \textit{E. coli} fermentation data versus HCM-FBA (solid).
@@
 
-HCM-FBA captured the shift from glucose to acetate consumption for a model of aerobic \textit{E. coli} metabolism that was infeasible for HCM-EM (Fig. \ref{fig:ecoli}, right).
+HCM-FBA captured the shift from glucose to acetate consumption for a model of aerobic \textit{E. coli} metabolism that was infeasible for HCM-EM (Fig. \ref{fig:ecoli}B).
 An \emph{E. coli} metabolic network (60 metabolites and 105 reactions) was constructed from literature \cite{2007_schuetz_etal_MolSysBio,2006_Palsson_model}.
@@
 Conversely, flux balance analysis generated only 29 modes for the same network.
-HCM-FBA model parameters were estimated from cellmass, glucose and acetate measurements \cite{1994_varma_palsson_ApplEnvMicro} using simulated annealing (Fig. \ref{fig:ecoli}B, solid).
-HCM-FBA captured glucose consumption, cellmass formation and the switch to acetate consumption following glucose exhaustion.
-Thus, HCM-FBA described the dynamics of a network that was infeasible for HCM-EM, thereby demonstrating the power of the approach for large networks.
+HCM-FBA model parameters were estimated from cellmass, glucose, and acetate measurements \cite{1994_varma_palsson_ApplEnvMicro} using simulated annealing (Fig. \ref{fig:ecoli}B, solid).
+HCM-FBA captured glucose consumption, cellmass formation, and the switch to acetate consumption following glucose exhaustion.
+HCM-FBA described the dynamics of a network that was infeasible for HCM-EM, thereby demonstrating the power of the approach for large networks.
 Next, we demonstrated a systematic strategy to identify the critical subset of FBA modes required for model performance.
@@
 Conversely, removing the remaining 24 modes had a negligible effect upon model performance (Fig. \ref{fig:ecoli}B, dashed).
-Thus, sensitivity analysis identified the minimal model structure required to explain the experimental data.
+The sensitivity analysis identified the minimal model structure required to explain the experimental data.
 
@@
 In this study, we developed HCM-FBA, an effective modeling technique to simulate metabolic dynamics.
-HCM-FBA is a modification of the hybrid cybernetic approach of Ramkrishna and coworkers.
+HCM-FBA is a modification of the hybrid cybernetic approach of Ramkrishna and coworkers \cite{2008_kim_varner_ramkrishna_BiotechProg}.
 HCM-FBA uses flux balance analysis solutions (instead of elementary modes) in conjunction with control variables to dynamically simulate metabolism.
 We studied the performance of HCM-FBA on a proof of concept metabolic network, along with two \emph{E. coli} networks.
-First, we showed the performance of HCM-FBA and HCM-EM were comparable for a proof of concept metabolic network, and a small model of anaerobic \textit{E. coli} metabolism.
+First, we showed that the performance of HCM-FBA and HCM-EM were comparable for a proof of concept metabolic network and a small model of anaerobic \textit{E. coli} metabolism.
 For the anaerobic case, both approaches described the experimental data.
@@
 Elementary mode decomposition of the core aerobic network generated over 150,000 elementary modes.
-Conversely, the HCM-FBA approach described cellmass growth, and the shift from glucose to acetate consumption, with only 29 FBA modes.
-Global sensitivity analysis further showed that only five of the 29 FBA modes were critical to model performance.
+Conversely, the HCM-FBA approach described cellmass growth and the shift from glucose to acetate consumption with only 29 FBA modes.
+Global sensitivity analysis further showed that only 5 of the 29 FBA modes were critical to model performance.
 Removal of these modes crippled the model, but removal of the remaining 24 modes had a negligible impact.
@@
 However, there are additional studies that should be performed to further test the HCM-FBA approach.
-First, the intracellular flux distribution predicted by HCM-FBA should be compared to HCM-EM, and to flux measurements calculated using metabolic flux analysis or flux balance analysis in combination with carbon labeling.
-HCM-EM had predicted intracellular fluxes that were similar to metabolic flux analysis results \cite{2008_kim_varner_ramkrishna_BiotechProg}.
+First, the intracellular flux distribution predicted by HCM-FBA should be compared to HCM-EM and to flux measurements calculated using MFA or FBA in combination with carbon labeling.
+HCM-EM predicted intracellular fluxes that were similar to MFA results \cite{2008_kim_varner_ramkrishna_BiotechProg}.
 Next, the performance of HCM-FBA to lumped hybrid cybernetic models (L-HCM) should be compared.
 L-HCMs, which combine elementary modes into mode families based upon metabolic function \cite{2010_song,Song:2011aa},
-have been applied to an \textit{E. coli} network with 67 reactions and a \textit{Saccharomyces cerevisiae} network (excluding compartmentalization) with 53 reactions;
+have been applied to an \textit{E. coli} network with 67 reactions and a \textit{Saccharomyces cerevisiae} network with 70 reactions;
 both cases had satisfactory fits to extracellular experimental data.
-However, while L-HCM reduces the dimension of possible alternative modes that must be considered, it still requires the calculation of initial set of modes.
-For metabolic networks of even moderate size this may not possible.
-On the other had, the generation of flux balance solutions (which are convex combinations of the elementary modes) is trivial, even for genome scale metabolic networks.
-Thus, HCM-FBA opens up the possibility of dynamic genome scale models of bacterial and perhaps even mammalian metabolism.
+However, while L-HCM reduces the dimension of possible alternative modes that must be considered, it still requires the calculation of an initial set of modes.
+For metabolic networks of even moderate size, EM decomposition may not be possible.
+On the other hand, the generation of flux balance solutions (convex combinations of the elementary modes) is trivial, even for genome scale metabolic networks.
+Thus, HCM-FBA opens up the possibility for dynamic genome scale models of bacterial and perhaps even of mammalian metabolism.
 
@@
 The HCM-FBA approach is a modification of HCM-EM, where elementary modes are replaced with flux balance analysis solutions.
-Thus, extracellular variables are dynamic, while intracellular metabolites are at a pseudo steady state.
-The abundance of extracellular species $i$ ($x_{i}$), the pseudo enzyme $e_{l}$ (catalyzes flux through mode l) and cellmass are governed by:
+Thus, extracellular variables are dynamic while intracellular metabolites are at a pseudo steady state.
+The abundance of extracellular species $i$ ($x_{i}$), the pseudo enzyme $e_{l}$ (catalyzes flux through mode l), and cellmass are governed by:
 \begin{eqnarray}\nonumber
@@
 \end{eqnarray}
-where $\mathcal{R}$ and $\mathcal{M}$ denote the number of reactions and extracellular species in the model, and $\mathcal{L}$ denotes the number of FBA modes.
+where $\mathcal{R}$ and $\mathcal{M}$ denote the number of reactions and extracellular species in the model and $\mathcal{L}$ denotes the number of FBA modes.
 The quantity $\sigma_{ij}$ denotes the stoichiometric coefficient for species $i$ in reaction $j$ and $z_{jl}$ denotes the normalized flux for reaction $j$ in mode $l$.
-If $\sigma_{ij}>0$, species $i$ is produced by reaction $j$,
-if $\sigma_{ij}<0$, species $i$ is consumed by reaction $j$, while $\sigma_{ij} = 0$ indicates species $i$ is not connected with reaction $j$.
+If $\sigma_{ij}>0$, species $i$ is produced by reaction $j$;
+if $\sigma_{ij}<0$, species $i$ is consumed by reaction $j$; if $\sigma_{ij} = 0$, species $i$ is not connected with reaction $j$.
 Extracellular species balances were subject to the initial conditions $\mathbf{x}\left(t_{o}\right) = \mathbf{x}_{o}$ determined from experimental data.
-The term $q_{l}\left(\mathbf{e},\mathbf{k},\mathbf{x}\right)$ denotes the specific uptake/secretion rate for mode $l$;
+The term $q_{l}\left(\mathbf{e},\mathbf{k},\mathbf{x}\right)$ denotes the specific uptake/secretion rate for mode $l$
+where $\mathbf{e}$ denotes the pseudo enzyme vector, $\mathbf{k}$ denotes the unknown kinetic parameter vector, and $\mathbf{x}$ denotes the extracellular species vector;
 $q_{l}\left(\mathbf{e},\mathbf{k},\mathbf{x}\right)$ is the product of a kinetic term ($\bar{q}_{l}$) and a control variable governing enzyme activity.
-Flux through each mode was catalyzed by a pseudo enzyme $e_{l}$, synthesized at the regulated specific rate $r_{E,l}\left(\mathbf{k},\mathbf{x}\right)$ and constitutively at the rate $\alpha_{l}$.
-The term $r_{E,l}$ denotes the specific rate enzyme synthesis for enzyme $l$, and $u_{l}$ denotes the variable controlling the synthesis of enzyme $l$.
+Flux through each mode was catalyzed by a pseudo enzyme $e_{l}$, synthesized at the regulated specific rate $r_{E,l}\left(\mathbf{k},\mathbf{x}\right)$, and constitutively at the rate $\alpha_{l}$.
+The term $r_{E,l}$ denotes the specific enzyme synthesis rate for enzyme $l$ and $u_{l}$ denotes the variable controlling the synthesis of enzyme $l$.
 The term $\beta_{l}$ denotes the rate constant governing non-specific enzyme degradation, and $r_{G}$ denotes the specific growth rate through all modes.
-The specific uptake/secretion rates, and the specific rate of enzyme synthesis were modeled using saturation kinetics.
+The specific uptake/secretion rates and the specific rate of enzyme synthesis were modeled using saturation kinetics.
 Lastly, cellmass was produced at the specific growth rate:
@@
 \end{align*}
-where $z_{sl}$ denotes the uptake flux of substrate $x$ through mode $l$.
-All model equations were implemented in Julia (v.0.4.2) \cite{Julia}, and solved using SUNDIALS \cite{Sundials}.
+where $z_{sl}$ denotes the uptake flux of substrate $s$ from $\mathbf{x}$ through mode $l$.
+All model equations were implemented in Julia (v.0.4.2) \cite{Julia} and solved using SUNDIALS \cite{Sundials}. 
 The model code for each case study is available for download at http://www.varnerlab.org.
@@
  \begin{multlined}
-	\qquad \qquad \qquad \max_{\boldsymbol{w}}{} \! \left( w_{obj} = \mathbf{\theta}^T \boldsymbol{w} \right) \\
+	\qquad \qquad \qquad \max_{\boldsymbol{w}}{} \! \left( w_{obj} = \mathbf{\boldsymbol{\theta}}^T \boldsymbol{w} \right) \\
 	\mathrm{Subject \; to:}
@@
 The flux balance analysis problem was solved using the GNU Linear Programming Kit, GLPK version 4.52 \cite{GLPK}.
-For each FBA mode, the objective was to maximize either the specific growth rate, or the specific rate of byproduct formation.
+For each FBA mode, the objective $w_{obj}$ was to maximize either the specific growth rate or the specific rate of byproduct formation.
 Multiple FBA modes were calculated for each objective flux by allowing the oxygen and nitrate uptake rates to vary.

### 2016-03-29 Paper_v1.tex
Final edits

@@
 Mathematical models of biochemical networks are useful tools to understand and ultimately predict how cells utilize nutrients to produce valuable products.
-Hybrid cybernetic models in combination with elementary modes (HCM-EM) provides a route to model cellular metabolism.
-However, this framework is limited to small intracellular networks due to the computational burden of calculating elementary modes.
+Hybrid cybernetic models in combination with elementary modes (HCM-EM) is a tool to model cellular metabolism.
+However, HCM-EM is limited to reduced metabolic networks because of the computational burden of calculating elementary modes.
 In this study, we developed the hybrid cybernetic modeling with flux balance analysis or HCM-FBA technique which uses flux balance solutions instead of elementary modes
 to dynamically model metabolism.
-We show HCM-FBA has comparable performance to HCM-EM for a hypothetical proof of concept metabolic network and for a reduced anaerobic \textit{E. coli} network.
-HCM-FBA was applied to a larger metabolic network of aerobic \textit{E. coli} metabolism which was infeasible for HCM-EM (29 FBA modes versus more than 153,000 elementary modes).
-Global sensitivity analysis further reduced the number of FBA modes required to describe the data from 29 to 5, while maintaining model fit.
-Thus, HCM-FBA is an alternative to HCM-EM for large networks where the generation of elementary modes is infeasible.
+We show HCM-FBA has comparable performance to HCM-EM for a proof of concept metabolic network and for a reduced anaerobic \textit{E. coli} network.
+Next, HCM-FBA was applied to a larger metabolic network of aerobic \textit{E. coli} metabolism which was infeasible for HCM-EM (29 FBA modes versus more than 153,000 elementary modes).
+Global sensitivity analysis further reduced the number of FBA modes required to describe the aerobic \textit{E. coli} data, while maintaining model fit.
+Thus, HCM-FBA is a promising alternative to HCM-EM for large networks where the generation of elementary modes is infeasible.
 
@@
 \begin{IEEEkeywords}
-Dynamic metabolic models, flux balance analysis, cybernetic models
+Metabolic models, flux balance analysis, cybernetic models
 \end{IEEEkeywords}
@@
 model intracellular metabolism using the biochemical stoichiometry and other constraints such as thermodynamical feasibility under pseudo-steady state conditions.
-FBA has predicted the global performance of genome scale networks \cite{Covert:2004aa}.
+FBA has been used to efficiently estimate the performance of metabolic networks of arbitrary complexity, including genome scale networks, using linear programming  \cite{Covert:2004aa}.
 On the other hand, EMs (or EPs) catalog all possible metabolic behaviors such that any flux distribution predicted by FBA is a convex combination
 of the EMs (or EPs) \cite{Wiback:2003aa}.
-However, the calculation of elementary modes is computationally expensive and currently infeasible for genome scale networks \cite{2004_lee_varner_ko_ieee}.
+However, the calculation of EMs (or EPs) is computationally expensive and currently infeasible for genome scale networks \cite{2004_lee_varner_ko_ieee}.
 
@@
 Hybrid cybernetic models (HCM) have addressed earlier shortcomings of the approach by integrating cybernetic optimality concepts with EMs.
-HCMs dynamically choose combinations of modes to achieve hypothetical physiological objectives (Fig. \ref{fig:model-fitting}A).
-HCMs generate intracellular flux distributions consistent with other approaches such as metabolic flux analysis (MFA), and also describe extracellular measurements \cite{2008_kim_varner_ramkrishna_BiotechProg}.
-However, HCMs are restricted to networks which can be decomposed into EMs.
+HCMs dynamically choose combinations of biochemical modes (each catalyzed by a pseudo enzyme whose expression is controlled by an optimal decision) to achieve a physiological objective (Fig.
+\ref{fig:model-fitting}A).
+HCMs generate intracellular flux distributions consistent with other approaches such as metabolic flux analysis (MFA), and also describe dynamic extracellular measurements \cite{2008_kim_varner_ramkrishna_BiotechProg}.
+However, HCMs are restricted to networks which can be decomposed into EMs (or EPs).
 
@@
 HCM-FBA is a modification of the hybrid cybernetic approach of Ramkrishna and coworkers \cite{2008_kim_varner_ramkrishna_BiotechProg} which uses FBA solutions
-(instead of EMs) in conjunction with control variables to dynamically simulate metabolism.
-We compared the performance of HCM-FBA to HCM-EM for a proof of concept metabolic network along with two \emph{E. coli} networks.
-HCM-FBA performed comparably to HCM-EM for the hypothetical network and a reduced anaerobic \textit{E. coli} network, despite having fewer parameters in each case.
+(instead of EMs) in conjunction with cybernetic control variables to dynamically simulate metabolism.
+We compared the performance of HCM-FBA to HCM-EM for a prototypical metabolic network along with two \emph{E. coli} networks.
+HCM-FBA performed comparably to HCM-EM for the prototypical network and a reduced anaerobic \textit{E. coli} network, despite having fewer parameters in each case.
 Next, HCM-FBA was applied to an aerobic \textit{E. coli} metabolic network that was infeasible for HCM-EM.
@@
 The oval denotes the cell boundary, $q_{j}$ is the $j$th flux across the boundary, and $v_{k}$ denotes the $k$th intracellular flux.
-C: Simulation of extracellular metabolite trajectories using HCM-FBA (solid line) versus HCM-EM (points) for the proof of concept metabolic network.
+C: Simulation of extracellular metabolite trajectories using HCM-FBA (solid line) versus HCM-EM (points) for the prototypical network.
 }\label{fig:model-fitting}
@@
 \section{Results}
-HCM-FBA was equivalent to HCM-EM for a proof of concept metabolic network (Fig. \ref{fig:model-fitting}).
+HCM-FBA was equivalent to HCM-EM for a prototypical metabolic network (Fig. \ref{fig:model-fitting}).
 The proof of concept network, consisting of 6 metabolites and 7 reactions (Fig. \ref{fig:model-fitting}B), generated 3 FBA modes and 6 EMs.
@@
 
+%HCM-FBA is a modification of the hybrid cybernetic approach of Ramkrishna and coworkers \cite{2008_kim_varner_ramkrishna_BiotechProg}.
+
 \section{Discussion}
 In this study, we developed HCM-FBA, an effective modeling technique to simulate metabolic dynamics.
-HCM-FBA is a modification of the hybrid cybernetic approach of Ramkrishna and coworkers \cite{2008_kim_varner_ramkrishna_BiotechProg}.
-HCM-FBA uses flux balance analysis solutions (instead of elementary modes) in conjunction with control variables to dynamically simulate metabolism.
-We studied the performance of HCM-FBA on a proof of concept metabolic network, along with two \emph{E. coli} networks.
-First, we showed that the performance of HCM-FBA and HCM-EM were comparable for a proof of concept metabolic network and a small model of anaerobic \textit{E. coli} metabolism.
+HCM-FBA uses flux balance analysis solutions (instead of elementary modes) in conjunction with cybernetic control variables to dynamically simulate metabolism.
+We studied the performance of HCM-FBA on a prototypical metabolic network, along with two \emph{E. coli} networks.
+First, we showed that the performance of HCM-FBA and HCM-EM were comparable for the prototypical network and a small model of anaerobic \textit{E. coli} metabolism.
 For the anaerobic case, both approaches described the experimental data.
@@
 Next, HCM-FBA was applied to an aerobic \textit{E. coli} metabolic network that was not feasible for HCM-EM.
-Elementary mode decomposition of the core aerobic network generated over 150,000 elementary modes.
+Elementary mode decomposition of the aerobic network generated over 153,000 elementary modes.
 Conversely, the HCM-FBA approach described cellmass growth and the shift from glucose to acetate consumption with only 29 FBA modes.
@@
 
-We have shown that HCM-FBA is an alternative approach to model large metabolic networks where elementary modes calculations are infeasible.
-However, there are additional studies that should be performed to further test the HCM-FBA approach.
+HCM-FBA is a promising approach to model large metabolic networks where elementary modes calculations are infeasible.
+However, there are additional studies that should be performed.
 First, the intracellular flux distribution predicted by HCM-FBA should be compared to HCM-EM and to flux measurements calculated using MFA or FBA in combination with carbon labeling.
-HCM-EM predicted intracellular fluxes that were similar to MFA results \cite{2008_kim_varner_ramkrishna_BiotechProg}.
-Next, the performance of HCM-FBA to lumped hybrid cybernetic models (L-HCM) should be compared.
+HCM-EM predicted intracellular fluxes that were similar to MFA results \cite{2008_kim_varner_ramkrishna_BiotechProg}, however the fluxes predicted by HCM-FBA have not yet been validated.
+Next, the performance of HCM-FBA should be compared to lumped hybrid cybernetic models (L-HCM).
 L-HCMs, which combine elementary modes into mode families based upon metabolic function \cite{2010_song,Song:2011aa},
@@
 However, while L-HCM reduces the dimension of possible alternative modes that must be considered, it still requires the calculation of an initial set of modes.
-For metabolic networks of even moderate size, EM decomposition may not be possible.
-On the other hand, the generation of flux balance solutions (convex combinations of the elementary modes) is trivial, even for genome scale metabolic networks.
+For metabolic networks of even moderate size, EM (or EP) decomposition may not be possible.
+On the other hand, the generation of flux balance solutions (convex combinations of the elementary modes or extreme pathways) is trivial, even for genome scale metabolic networks.
 Thus, HCM-FBA opens up the possibility for dynamic genome scale models of bacterial and perhaps even of mammalian metabolism.
@@
 Thus, extracellular variables are dynamic while intracellular metabolites are at a pseudo steady state.
-The abundance of extracellular species $i$ ($x_{i}$), the pseudo enzyme $e_{l}$ (catalyzes flux through mode l), and cellmass are governed by:
+The abundance of extracellular species $i$ ($x_{i}$), the pseudo enzyme $e_{l}$ (catalyzes flux through mode $l$), and cellmass are governed by:
 \begin{eqnarray}\nonumber
@@
 Flux through each mode was catalyzed by a pseudo enzyme $e_{l}$, synthesized at the regulated specific rate $r_{E,l}\left(\mathbf{k},\mathbf{x}\right)$, and constitutively at the rate $\alpha_{l}$.
-The term $r_{E,l}$ denotes the specific enzyme synthesis rate for enzyme $l$ and $u_{l}$ denotes the variable controlling the synthesis of enzyme $l$.
+The term $u_{l}$ denotes the cybernetic variable controlling the synthesis of enzyme $l$.
 The term $\beta_{l}$ denotes the rate constant governing non-specific enzyme degradation, and $r_{G}$ denotes the specific growth rate through all modes.
 The specific uptake/secretion rates and the specific rate of enzyme synthesis were modeled using saturation kinetics.
-Lastly, cellmass was produced at the specific growth rate:
+The specific growth rate was given by:
 \begin{equation}
@@
 \end{align*}
-where $z_{sl}$ denotes the uptake flux of substrate $s$ from $\mathbf{x}$ through mode $l$.
-All model equations were implemented in Julia (v.0.4.2) \cite{Julia} and solved using SUNDIALS \cite{Sundials}. 
-The model code for each case study is available for download at http://www.varnerlab.org.
+where $z_{sl}$ denotes the uptake flux of substrate $s$ through mode $l$.
+The model equations were implemented in Julia (v.0.4.2) \cite{Julia} and solved using SUNDIALS \cite{Sundials}.
+The model code for each case study is available at http://www.varnerlab.org.
 
@@
 	 \; \; \mathbf{S}\mathbf{w}=\mathbf{0} \\
-\alpha_i \leq w_i \leq \beta_i  \qquad
+\alpha_i \leq w_i \leq \beta_i  \qquad i=1,2,\hdots,\mathcal{R}
  \end{multlined}
@@
 and $\alpha_i$ and $\beta_i$ denote the lower and upper bounds on flux $w_{i}$, respectively.
-The flux balance analysis problem was solved using the GNU Linear Programming Kit, GLPK version 4.52 \cite{GLPK}.
+The flux balance analysis problem was solved using the GNU Linear Programming Kit (v4.52) \cite{GLPK}.
 For each FBA mode, the objective $w_{obj}$ was to maximize either the specific growth rate or the specific rate of byproduct formation.
-Multiple FBA modes were calculated for each objective flux by allowing the oxygen and nitrate uptake rates to vary.
+Multiple FBA modes were calculated for each objective by allowing the oxygen and nitrate uptake rates to vary.
 For aerobic metabolism, the specific oxygen and nitrate uptake rates were constrained to allow a maximum flux of 10 mM/gDW$\cdot$hr and 0.05 mM/gDW$\cdot$hr, respectively.
@@
 value for species $j$ at time $\tau$, and $\omega_{j}\left(\tau\right)$ denotes the experimental measurement variance for species $j$ at time $\tau$.
-The outer summation is with respect to time, while the inner summation is with respect to state. The model residual was minimized using simulated annealing.
+The outer summation is with respect to time, while the inner summation is with respect to state.
+The model residual was minimized using simulated annealing implemented in the Julia programming language.
 

### 2016-03-29 Paper_v1.tex
Final Updates

@@
 The performance of HCM-FBA was equivalent to HCM-EM for anaerobic \textit{E. coli} metabolism (Fig. \ref{fig:ecoli}A).
-We constructed an anaerobic \textit{E. coli} network, consisting of 12 reactions and 19 metabolites, which generated 7 FBA modes and 9 EMs.
+We constructed an anaerobic \textit{E. coli} network \cite{2008_kim_varner_ramkrishna_BiotechProg}, consisting of 12 reactions and 19 metabolites, which generated 7 FBA modes and 9 EMs.
 HCM-EM reproduced cellmass, glucose, and byproduct trajectories using the kinetic parameters reported by Kim \textit{et al.} \cite{2008_kim_varner_ramkrishna_BiotechProg} (Fig.
@@
 For the anaerobic case, both approaches described the experimental data.
-However HCM-FBA (which was within 5\% of HCM-EM and slightly better than HCM-EM for lactate secretion) had fewer modes and parameters.
+However, HCM-FBA (which was within 5\% of HCM-EM and slightly better than HCM-EM for lactate secretion) had fewer modes and parameters.
 Next, HCM-FBA was applied to an aerobic \textit{E. coli} metabolic network that was not feasible for HCM-EM.
@@
 First, the intracellular flux distribution predicted by HCM-FBA should be compared to HCM-EM and to flux measurements calculated using MFA or FBA in combination with carbon labeling.
-HCM-EM predicted intracellular fluxes that were similar to MFA results \cite{2008_kim_varner_ramkrishna_BiotechProg}, however the fluxes predicted by HCM-FBA have not yet been validated.
+HCM-EM predicted intracellular fluxes that were similar to MFA results \cite{2008_kim_varner_ramkrishna_BiotechProg}; however, the fluxes predicted by HCM-FBA have not yet been validated.
 Next, the performance of HCM-FBA should be compared to lumped hybrid cybernetic models (L-HCM).
@@
 The specific growth rate was given by:
-\begin{equation}
+\begin{equation}\nonumber
 	r_{G}  = \sum_{l = 1}^{\mathcal{L}}z_{\mu l}q_{l}\left(\mathbf{e},\mathbf{k},\mathbf{x}\right)
@@
 The control variables $u_{l}$ and $v_{l}$, which control the synthesis and activity of each enzyme respectively, were given by:
+
 \begin{align*}
@@
 \end{align*}
+
 where $z_{sl}$ denotes the uptake flux of substrate $s$ through mode $l$.
@@
 The FBA problem was formulated as:
-\begin{equation}
+\begin{equation}\nonumber
  \begin{multlined}
@@
 
-\begin{equation}\label{eqn:objective-function}
+\begin{equation}\label{eqn:objective-function}\nonumber
 	\min_{\mathbf{k}} \sum_{\tau=1}^{\mathcal{T}}\sum_{j=1}^{\mathcal{S}}\left(\frac{\hat{x}_{j}\left(\tau\right) - x_{j}\left(\tau,\mathbf{k}\right)}{\omega_{j}\left(\tau\right)}\right)^{2}

### 2016-07-28 Paper_v1.tex
Add files via upload

@@
 Mathematical models of biochemical networks are useful tools to understand and ultimately predict how cells utilize nutrients to produce valuable products.
-Hybrid cybernetic models in combination with elementary modes (HCM-EM) is a tool to model cellular metabolism.
-However, HCM-EM is limited to reduced metabolic networks because of the computational burden of calculating elementary modes.
+Hybrid cybernetic models in combination with elementary modes (HCM) is a tool to model cellular metabolism.
+However, HCM is limited to reduced metabolic networks because of the computational burden of calculating elementary modes.
 In this study, we developed the hybrid cybernetic modeling with flux balance analysis or HCM-FBA technique which uses flux balance solutions instead of elementary modes
 to dynamically model metabolism.
-We show HCM-FBA has comparable performance to HCM-EM for a proof of concept metabolic network and for a reduced anaerobic \textit{E. coli} network.
-Next, HCM-FBA was applied to a larger metabolic network of aerobic \textit{E. coli} metabolism which was infeasible for HCM-EM (29 FBA modes versus more than 153,000 elementary modes).
+We show HCM-FBA has comparable performance to HCM for a proof of concept metabolic network and for a reduced anaerobic \textit{E. coli} network.
+Next, HCM-FBA was applied to a larger metabolic network of aerobic \textit{E. coli} metabolism which was infeasible for HCM (29 FBA modes versus more than 153,000 elementary modes).
 Global sensitivity analysis further reduced the number of FBA modes required to describe the aerobic \textit{E. coli} data, while maintaining model fit.
-Thus, HCM-FBA is a promising alternative to HCM-EM for large networks where the generation of elementary modes is infeasible.
+Thus, HCM-FBA is a promising alternative to HCM for large networks where the generation of elementary modes is infeasible.
 
@@
 \ref{fig:model-fitting}A).
-HCMs generate intracellular flux distributions consistent with other approaches such as metabolic flux analysis (MFA), and also describe dynamic extracellular measurements \cite{2008_kim_varner_ramkrishna_BiotechProg}.
+HCMs generate intracellular flux distributions consistent with other approaches such as metabolic flux analysis (MFA), and also describe dynamic extracellular measurements superior to dynamic FBA (DFBA) \cite{2008_kim_varner_ramkrishna_BiotechProg}.
 However, HCMs are restricted to networks which can be decomposed into EMs (or EPs).
@@
 (instead of EMs) in conjunction with cybernetic control variables to dynamically simulate metabolism.
-We compared the performance of HCM-FBA to HCM-EM for a prototypical metabolic network along with two \emph{E. coli} networks.
-HCM-FBA performed comparably to HCM-EM for the prototypical network and a reduced anaerobic \textit{E. coli} network, despite having fewer parameters in each case.
-Next, HCM-FBA was applied to an aerobic \textit{E. coli} metabolic network that was infeasible for HCM-EM.
+Since HCM showed superior performance to DFBA we compared the performance of HCM-FBA to HCM for a prototypical metabolic network to show comparable performance along with two \emph{E.
+coli} networks for real-world applicatons.
+HCM-FBA performed comparably to HCM for the prototypical network and a reduced anaerobic \textit{E. coli} network, despite having fewer parameters in each case.
+Next, HCM-FBA was applied to an aerobic \textit{E. coli} metabolic network that was infeasible for HCM.
 HCM-FBA described cellmass growth and the shift from glucose to acetate consumption with only a few modes.
@@
 Thus, HCM-FBA is a promising approach for the development of reduced order dynamic metabolic models
-and a viable alternative to HCM-EM, especially for large networks where the generation of EMs is infeasible.
+and a viable alternative to HCM, especially for large networks where the generation of EMs is infeasible.
 
-%HCM-FBA also captured the dynamics of a larger anaerobic \textit{E. coli} network, that was infeasible for HCM-EM.
-%Taken together, HCM-FBA has comparable model performace to HCM-EM and may be potentially applied to genome-scale networks which was a shortcoming of HCM-EM.
-%HCM-EM and HCM-FBA were then applied to a large metabolic network.
-%The number of elementary modes made it impractical to model with HCM-EM, but HCM-FBA was computationally feasible and performed well with respect to experimental observations.
+%HCM-FBA also captured the dynamics of a larger anaerobic \textit{E. coli} network, that was infeasible for HCM.
+%Taken together, HCM-FBA has comparable model performace to HCM and may be potentially applied to genome-scale networks which was a shortcoming of HCM.
+%HCM and HCM-FBA were then applied to a large metabolic network.
+%The number of elementary modes made it impractical to model with HCM, but HCM-FBA was computationally feasible and performed well with respect to experimental observations.
 
 \begin{figure}[!t]\centering
-\includegraphics[width=0.45\textwidth]{./figs/Fig-1-Method-General.pdf}
+\includegraphics[width=0.43\textwidth]{./figs/Fig-1-Method-General.pdf}
 \caption{HCM proof of concept metabolic study.
-A: HCMs distribute uptake and secretion fluxes amongst different pathways. For HCM-EM, these pathways are elementary modes; for HCM-FBA these are flux balance analysis solutions.
-HCM-EM combines all possible modes within a network; whereas HCM-FBA combines only steady-state paths estimated by flux balance analysis.
+A: HCMs distribute uptake and secretion fluxes amongst different pathways. For HCM, these pathways are elementary modes; for HCM-FBA these are flux balance analysis solutions.
+HCM combines all possible modes within a network; whereas HCM-FBA combines only steady-state paths estimated by flux balance analysis.
 B: Prototypical network with six metabolites and seven reactions.
@@
 The oval denotes the cell boundary, $q_{j}$ is the $j$th flux across the boundary, and $v_{k}$ denotes the $k$th intracellular flux.
-C: Simulation of extracellular metabolite trajectories using HCM-FBA (solid line) versus HCM-EM (points) for the prototypical network.
+C: Simulation of extracellular metabolite trajectories using HCM-FBA (solid line) versus HCM (points) for the prototypical network.
 }\label{fig:model-fitting}
@@
 \section{Results}
-HCM-FBA was equivalent to HCM-EM for a prototypical metabolic network (Fig. \ref{fig:model-fitting}).
+HCM-FBA was equivalent to HCM for a prototypical metabolic network (Fig. \ref{fig:model-fitting}).
 The proof of concept network, consisting of 6 metabolites and 7 reactions (Fig. \ref{fig:model-fitting}B), generated 3 FBA modes and 6 EMs.
@@
 The best fit HCM-FBA model replicated the synthetic data (Fig. \ref {fig:model-fitting}C).
-The HCM-EM and HCM-FBA kinetic parameters were not quantitatively identical, but had similar orders of magnitude;
+The HCM and HCM-FBA kinetic parameters were not quantitatively identical, but had similar orders of magnitude;
 the FBA approach had 3 fewer modes, thus identical parameter values were not expected.
-Taken together, the HCM-FBA approach replicated synthetic data generated by HCM-EM, despite having 3 fewer modes.
+Taken together, the HCM-FBA approach replicated synthetic data generated by HCM, despite having 3 fewer modes.
 Next, we tested the ability of HCM-FBA to replicate experimental data.
 
-%In the case, we demonstrated that HCM-FBA has comparable model performance to a published HCM-EM model \cite{2008_kim_varner_ramkrishna_BiotechProg}.
+%In the case, we demonstrated that HCM-FBA has comparable model performance to a published HCM model \cite{2008_kim_varner_ramkrishna_BiotechProg}.
 %and elementary modes reported
 %The kinetic parameters of HCM-FBA were varied to fit the same experimental observations.
-%HCM-FBA has 17 kinetic parameters while HCM-EM has 21 kinetic parameters and both models resulted in similar fits to the experimental data .
-%HCM-FBA had a better fit on lactate than HCM-EM by 35.6\% on the
-%However, HCM-EM had a lower residual standard error on the overall performance by 4.5\%.
+%HCM-FBA has 17 kinetic parameters while HCM has 21 kinetic parameters and both models resulted in similar fits to the experimental data .
+%HCM-FBA had a better fit on lactate than HCM by 35.6\% on the
+%However, HCM had a lower residual standard error on the overall performance by 4.5\%.
 %However, ZZZ has shown that the EM algorithm has exponential runtime complexity [REFHERE].
-%Thus, HCM-EM may not be possible for large networks, but HCM-FBA will not suffer from this shortcoming.
+%Thus, HCM may not be possible for large networks, but HCM-FBA will not suffer from this shortcoming.
 %In the case, we demonstrated that HCM-FBA is applicable for larger metabolic networks where elementary mode calculations increase exponentially.
 
-The performance of HCM-FBA was equivalent to HCM-EM for anaerobic \textit{E. coli} metabolism (Fig. \ref{fig:ecoli}A).
+The performance of HCM-FBA was equivalent to HCM for anaerobic \textit{E. coli} metabolism (Fig. \ref{fig:ecoli}A).
 We constructed an anaerobic \textit{E. coli} network \cite{2008_kim_varner_ramkrishna_BiotechProg}, consisting of 12 reactions and 19 metabolites, which generated 7 FBA modes and 9 EMs.
-HCM-EM reproduced cellmass, glucose, and byproduct trajectories using the kinetic parameters reported by Kim \textit{et al.} \cite{2008_kim_varner_ramkrishna_BiotechProg} (Fig.
+HCM reproduced cellmass, glucose, and byproduct trajectories using the kinetic parameters reported by Kim \textit{et al.} \cite{2008_kim_varner_ramkrishna_BiotechProg} (Fig.
 \ref{fig:ecoli}A, points versus dashed).
 HCM-FBA model parameters were estimated in this study from the Kim \textit{et al.} data set using simulated annealing.
-Overall, HCM-FBA performed within 5\% of HCM-EM (on a residual standard error basis) for the anaerobic \textit{E. coli} data (Fig. \ref {fig:ecoli}A, solid),
+Overall, HCM-FBA performed within 5\% of HCM (on a residual standard error basis) for the anaerobic \textit{E. coli} data (Fig. \ref {fig:ecoli}A, solid),
 despite having 2 fewer modes and 4 fewer parameters (17 versus 21 parameters).
-Thus, while both HCM-EM and HCM-FBA described the experimental data, HCM-FBA did so with fewer modes and parameters.
+Thus, while both HCM and HCM-FBA described the experimental data, HCM-FBA did so with fewer modes and parameters.
 
 \begin{figure}[!t]\centering
-\includegraphics[width=0.45\textwidth]{./figs/Fig-2-Ecoli-SimulationResults.pdf}
-\caption{HCM-FBA versus HCM-EM performance for small and large metabolic networks.
-A: Batch anaerobic \textit{E. coli} fermentation data versus HCM-FBA (solid) and HCM-EM (dashed).
+\includegraphics[width=0.43\textwidth]{./figs/Fig-2-Ecoli-SimulationResults.pdf}
+\caption{HCM-FBA versus HCM performance for small and large metabolic networks.
+A: Batch anaerobic \textit{E. coli} fermentation data versus HCM-FBA (solid) and HCM (dashed).
 The experimental data was reproduced from Kim \textit{et al.} \cite{2008_kim_varner_ramkrishna_BiotechProg}. Error bars represent the 90\% confidence interval.
@@
 
-HCM-FBA captured the shift from glucose to acetate consumption for a model of aerobic \textit{E. coli} metabolism that was infeasible for HCM-EM (Fig. \ref{fig:ecoli}B).
+HCM-FBA captured the shift from glucose to acetate consumption for a model of aerobic \textit{E. coli} metabolism that was infeasible for HCM (Fig. \ref{fig:ecoli}B).
 An \emph{E. coli} metabolic network (60 metabolites and 105 reactions) was constructed from literature \cite{2007_schuetz_etal_MolSysBio,2006_Palsson_model}.
-Elementary mode decomposition of this network (and thus HCM-EM) was not feasible; 153,000 elementary modes were generated before the calculation became infeasible.
+Elementary mode decomposition of this network (and thus HCM) was not feasible; 153,000 elementary modes were generated before the calculation became infeasible.
 Conversely, flux balance analysis generated only 29 modes for the same network.
@@
 HCM-FBA captured glucose consumption, cellmass formation, and the switch to acetate consumption following glucose exhaustion.
-HCM-FBA described the dynamics of a network that was infeasible for HCM-EM, thereby demonstrating the power of the approach for large networks.
+HCM-FBA described the dynamics of a network that was infeasible for HCM, thereby demonstrating the power of the approach for large networks.
 Next, we demonstrated a systematic strategy to identify the critical subset of FBA modes required for model performance.
@@
 removal of the most significant of these modes (encoding aerobic growth on glucose) destroyed model performance (Fig. \ref{fig:ecoli}B, dotted).
-Conversely, removing the remaining 24 modes had a negligible effect upon model performance (Fig. \ref{fig:ecoli}B, dashed).
+Conversely, removing the remaining 24 modes simultaneously had a negligible effect upon model performance (Fig. \ref{fig:ecoli}B, dashed).
 The sensitivity analysis identified the minimal model structure required to explain the experimental data.
@@
 \begin{figure*}[!t]\centering
-\includegraphics[width=0.80\textwidth]{./figs/Fig-3-Sensitivity-Results.pdf}
+\includegraphics[width=0.68\textwidth]{./figs/Fig-3-Sensitivity-Results.pdf}
 \caption{Global sensitivity analysis of the aerobic \textit{E. coli} model.
@@
 We studied the performance of HCM-FBA on a prototypical metabolic network, along with two \emph{E. coli} networks.
-First, we showed that the performance of HCM-FBA and HCM-EM were comparable for the prototypical network and a small model of anaerobic \textit{E. coli} metabolism.
+First, we showed that the performance of HCM-FBA and HCM were comparable for the prototypical network and a small model of anaerobic \textit{E. coli} metabolism.
 For the anaerobic case, both approaches described the experimental data.
-However, HCM-FBA (which was within 5\% of HCM-EM and slightly better than HCM-EM for lactate secretion) had fewer modes and parameters.
-Next, HCM-FBA was applied to an aerobic \textit{E. coli} metabolic network that was not feasible for HCM-EM.
+However, HCM-FBA (which was within 5\% of HCM and slightly better than HCM for lactate secretion) had fewer modes and parameters.
+Next, HCM-FBA was applied to an aerobic \textit{E. coli} metabolic network that was not feasible for HCM.
 Elementary mode decomposition of the aerobic network generated over 153,000 elementary modes.
@@
 Removal of these modes crippled the model, but removal of the remaining 24 modes had a negligible impact.
-Thus, HCM-FBA is an alternative approach to HCM-EM, especially for large networks where the generation of elementary modes is infeasible.
+These insiginificant modes are associated with maintenance states, thus they would not impact the model's predictive capability for a growing culture.
+A mode consuming a substrate is determined to be active which can give insight to the operating state of the cell.
+HCM-FBA is an alternative approach to HCM, especially for large networks where the generation of elementary modes is infeasible.
+Elementary modes show the complexity of a cell, displaying the many routes it can take but mathematically FBA has an objective superiority for large networks.
 
-HCM-FBA is a promising approach to model large metabolic networks where elementary modes calculations are infeasible.
+HCM-FBA is a promising approach to model large metabolic networks where elementary modes calculations are infeasible and kinetic models of such systems have uncertainty in parameter estimations.
 However, there are additional studies that should be performed.
-First, the intracellular flux distribution predicted by HCM-FBA should be compared to HCM-EM and to flux measurements calculated using MFA or FBA in combination with carbon labeling.
-HCM-EM predicted intracellular fluxes that were similar to MFA results \cite{2008_kim_varner_ramkrishna_BiotechProg}; however, the fluxes predicted by HCM-FBA have not yet been validated.
+First, the intracellular flux distribution predicted by HCM-FBA should be compared to HCM and to flux measurements calculated using MFA or FBA in combination with carbon labeling.
+HCM predicted intracellular fluxes that were similar to MFA results \cite{2008_kim_varner_ramkrishna_BiotechProg}; however, the fluxes predicted by HCM-FBA have not yet been validated.
 Next, the performance of HCM-FBA should be compared to lumped hybrid cybernetic models (L-HCM).
@@
 %The HCM framework has been shown to predict dynamic external fluxes of cell growth, substrate utilization and byproduct formation.
-%HCM-EM has good model performance for reduced networks, but is often impractical for larger networks.
+%HCM has good model performance for reduced networks, but is often impractical for larger networks.
 %However, Song and Ramkrishna address this shortcoming by lumping .
@@
 %, this has not been shown for HCM-FBA.
-%Just as HCM-EM has been applied to \textit{S. cerevisiae}, we can apply the HCM-FBA framework to eukaryotes as long as FBA accounts for compartmentalization.
+%Just as HCM has been applied to \textit{S. cerevisiae}, we can apply the HCM-FBA framework to eukaryotes as long as FBA accounts for compartmentalization.
 %FBA is frequently used to study genome-scale networks\cite{2010_orth_NatBiotech} and can be used to generate modes for the HCM framework.
@@
 %The internal fluxes still follow the pseudo steady state approximation applied in FBA, but HCM uses a combination of pathways.
-%HCM-EM has been shown to have comparable internal flux estimates to MFA results\cite{2008_kim_varner_ramkrishna_BiotechProg}, this has not been shown for HCM-FBA.
+%HCM has been shown to have comparable internal flux estimates to MFA results\cite{2008_kim_varner_ramkrishna_BiotechProg}, this has not been shown for HCM-FBA.
 %The HCM framework has been shown to predict dynamic external fluxes using both elementary and FBA modes.
@@
 %HCM overcomes this by incorparting the regulatory processes of substrate utilization based on its cybernetic arguments.
-%HCM-EM has good model performance for reduced networks, but is impractical for larger networks.
-%HCM-EM has been applied to a network with 67 reactions, requiring the elementary modes to be lumped into groups by complex weighting schemes and relying on experimental data\cite{2010_song_ramkrishna}.
-%A shortcoming of the HCM-EM framework is that EM calculations are computationally expensive and the number of modes exponentially increases for larger networks \cite{2004_lee_varner_ko_ieee}.
+%HCM has good model performance for reduced networks, but is impractical for larger networks.
+%HCM has been applied to a network with 67 reactions, requiring the elementary modes to be lumped into groups by complex weighting schemes and relying on experimental data\cite{2010_song_ramkrishna}.
+%A shortcoming of the HCM framework is that EM calculations are computationally expensive and the number of modes exponentially increases for larger networks \cite{2004_lee_varner_ko_ieee}.
 %In contrast, FBA does not have the computational burden associated with calculating elementary modes.
@@
 \section{Materials and Methods}
-The HCM-FBA approach is a modification of HCM-EM, where elementary modes are replaced with flux balance analysis solutions.
+The HCM-FBA approach is a modification of HCM, where elementary modes are replaced with flux balance analysis solutions.
 Thus, extracellular variables are dynamic while intracellular metabolites are at a pseudo steady state.
@@
 The term $q_{l}\left(\mathbf{e},\mathbf{k},\mathbf{x}\right)$ denotes the specific uptake/secretion rate for mode $l$
-where $\mathbf{e}$ denotes the pseudo enzyme vector, $\mathbf{k}$ denotes the unknown kinetic parameter vector, and $\mathbf{x}$ denotes the extracellular species vector;
+where $\mathbf{e}$ denotes the pseudo enzyme vector, $\mathbf{k}$ denotes the unknown kinetic parameter vector, $\mathbf{x}$ denotes the extracellular species vector, and $c$ denotes the cell mass;
 $q_{l}\left(\mathbf{e},\mathbf{k},\mathbf{x}\right)$ is the product of a kinetic term ($\bar{q}_{l}$) and a control variable governing enzyme activity.

### 2016-08-09 Paper_v2.tex
Final version

--- 2016-07-28 Paper_v1.tex
@@
 \author{\IEEEauthorblockN{Michael Vilkhovoy\IEEEauthorrefmark{1},
-Mason Minot\IEEEauthorrefmark{2},
+Mason Minot\IEEEauthorrefmark{1},
 Jeffrey D. Varner\IEEEauthorrefmark{1}}\\
-\IEEEauthorblockA{\IEEEauthorrefmark{1}School of Chemical Engineering, Purdue University, West Lafayette, IN 47907 USA}
-\IEEEauthorblockA{\IEEEauthorrefmark{2}School of Chemical and Biomolecular Engineering, Cornell University, Ithaca, NY 14850 USA}
-\thanks{Corresponding author: J. Varner (email:jdvarner@purdue.edu).}}
+\IEEEauthorblockA{\IEEEauthorrefmark{1}School of Chemical and Biomolecular Engineering, Cornell University, Ithaca, NY 14850 USA}
+\thanks{Corresponding author: J. Varner (email:jdv27@cornell.edu).}}
 
@@
 (instead of EMs) in conjunction with cybernetic control variables to dynamically simulate metabolism.
-Since HCM showed superior performance to DFBA we compared the performance of HCM-FBA to HCM for a prototypical metabolic network to show comparable performance along with two \emph{E.
-coli} networks for real-world applicatons.
+Since HCM showed superior performance to DFBA, we compared the performance of HCM-FBA to HCM for a prototypical metabolic network, along with two real-world \emph{E. coli} applications.
 HCM-FBA performed comparably to HCM for the prototypical network and a reduced anaerobic \textit{E. coli} network, despite having fewer parameters in each case.
@@
 Thus, HCM-FBA is a promising approach for the development of reduced order dynamic metabolic models
-and a viable alternative to HCM, especially for large networks where the generation of EMs is infeasible.
+and a viable alternative to HCM or DFBA, especially for large networks where the generation of EMs is infeasible.
 
@@
 the FBA approach had 3 fewer modes, thus identical parameter values were not expected.
-Taken together, the HCM-FBA approach replicated synthetic data generated by HCM, despite having 3 fewer modes.
-Next, we tested the ability of HCM-FBA to replicate experimental data.
+The HCM-FBA approach replicated synthetic data generated by HCM, despite having 3 fewer modes.
+Thus, we expect HCM-FBA will perform similarly to HCM, despite having fewer parameters.
+Next, we tested the ability of HCM-FBA to replicate real-world experimental data.
 
@@
 In this study, we developed HCM-FBA, an effective modeling technique to simulate metabolic dynamics.
-HCM-FBA uses flux balance analysis solutions (instead of elementary modes) in conjunction with cybernetic control variables to dynamically simulate metabolism.
+HCM-FBA uses flux balance analysis solutions in conjunction with cybernetic control variables to dynamically simulate metabolism.
 We studied the performance of HCM-FBA on a prototypical metabolic network, along with two \emph{E. coli} networks.
@@
 Removal of these modes crippled the model, but removal of the remaining 24 modes had a negligible impact.
-These insiginificant modes are associated with maintenance states, thus they would not impact the model's predictive capability for a growing culture.
-A mode consuming a substrate is determined to be active which can give insight to the operating state of the cell.
+These insignificant modes were associated with maintenance, thus they would likely not impact model predictions for a growing culture.
 HCM-FBA is an alternative approach to HCM, especially for large networks where the generation of elementary modes is infeasible.
@@
 
-HCM-FBA is a promising approach to model large metabolic networks where elementary modes calculations are infeasible and kinetic models of such systems have uncertainty in parameter estimations.
+% Mike: I don't understand this sentence?
+%A mode consuming a substrate was determined to be active which can give insight to the operating state of the cell.
+
+HCM-FBA is a promising approach to model large metabolic networks where elementary modes calculations are infeasible,
+and where kinetic models of such systems have intractable identification problems.
 However, there are additional studies that should be performed.
-First, the intracellular flux distribution predicted by HCM-FBA should be compared to HCM and to flux measurements calculated using MFA or FBA in combination with carbon labeling.
+First, the intracellular flux distribution predicted by HCM-FBA should be compared to HCM and to flux measurements calculated using MFA or FBA/DFBA in combination with carbon labeling.
 HCM predicted intracellular fluxes that were similar to MFA results \cite{2008_kim_varner_ramkrishna_BiotechProg}; however, the fluxes predicted by HCM-FBA have not yet been validated.
@@
 The control variables $u_{l}$ and $v_{l}$, which control the synthesis and activity of each enzyme respectively, were given by:
-
 \begin{align*}
@@
 \end{align*}
-
 where $z_{sl}$ denotes the uptake flux of substrate $s$ through mode $l$.
 The model equations were implemented in Julia (v.0.4.2) \cite{Julia} and solved using SUNDIALS \cite{Sundials}.
-The model code for each case study is available at http://www.varnerlab.org.
+The model code is available at http://www.varnerlab.org under a MIT license.
 
